### 2011-02-21 paper.tex
Added my name and affiliation

%% BioMed_Central_Tex_Template_v1.05
%%                                      %
%  bmc_article.tex            ver: 1.05 %
%                                       %

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%         <27 January 2006>           %%
%%                                     %%
%%                                     %%
%% Uses:                               %%
%% cite.sty, url.sty, bmc_article.cls  %%
%% ifthen.sty. multicol.sty		       %%
%%									   %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%	
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.pdf and the instructions for    %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%   
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %% 
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %% 
%%  submitted article.                         %% 
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{A sample article title}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}%
      }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
}%

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the Instructions for       %%
%% authors on http://www.biomedcentral.com  %%
%% and change the section headings          %%
%% accordingly.                             %%   
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} Text for this section of the abstract.

        \paragraph*{Results:} Text for this section of the abstract \ldots

        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
\end{abstract}

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the instructions for       %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and change the section headings          %%
%% accordingly.                             %% 
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
 Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{Results sub-heading}
    \subsubsection*{This is a sub-sub-heading}
      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
      pb at end of lines ensures correct paragraph spacing.\pb
	  Text for this sub-sub-section \ldots
    \subsubsection*{Another sub-sub-sub-heading}
      Text for this sub-sub-section \ldots

  \subsection*{Another results sub-heading}
    Text for this sub-section \ldots

  \subsection*{Yet another results sub-heading}
    Text for this sub-section.  More results \ldots

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%
\section*{Methods}
  \subsection*{Methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Another methods sub-heading for this section}
    Text for this sub-section \ldots

  \subsection*{Yet another sub-heading for this section}
    Text for this sub-section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%              
%%  Bmc_article.bst  will be used to                       %%
%%  create a .BBL file for submission, which includes      %%
%%  XML structured for BMC.                                %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %% 
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %% 
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\ifthenelse{\boolean{publ}}{\end{multicols}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{bmcformat}
\end{document}

### 2011-03-01 paper.tex
Added a draft title
Added name to paper
Some initial text

@@
 
-\title{A sample article title}
+\title{The Blue Obelisk five years on}
 
@@
       \and 
+         Egon L Willighagen$^3$%
+         \email{Egon L Willighagen - egon.willighagen@ki.se}
+     \and 
+         Rajarshi Guha$^4$%
+         \email{Rajarshi Guha - guhar@mail.nih.gov}
+     \and 
+         Noel M O'Boyle$^5$%
+         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+      \and
          Jane E Doe\correspondingauthor$^2$%
          \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
-      \and 
-         Egon L Willighagen$^3$%
-         \email{Egon L Willighagen - egon.willighagen@ki.se}%
       }
@@
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
+    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland
 }%
@@
 \section*{Background}
- Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
+The Blue Obelisk group was established in 200X at the 2XXth National Meeting of the American Chemistry Society as a response to the lack of open data, open standards and open source (ODOSOS) in chemistry.
+While other scientific disciplines such as physics, biology and astronomy (to name a few) were embracing new ways of doing science and reaping the benefits of community efforts, there was little if any innovation in the field of chemistry and scientific progress was actively hampered by the lack of access to data and tools.
+
+The goals of the Blue Obelisk group were described in Guha et al.\cite{guha2006}. but are briefly summarised here...
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Main section %%
+%%
+\section*{Open Source}
+  \subsection*{Progress}
+Progress to date...
+  \subsection*{Remaining challenges}
+What's left to do?
 
@@
 
-%%%%%%%%%%%%%%%%%%
-\section*{Methods}
-  \subsection*{Methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Another methods sub-heading for this section}
-    Text for this sub-section \ldots
-
-  \subsection*{Yet another sub-heading for this section}
-    Text for this sub-section \ldots
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Authors contributions}
-    Text for this section \ldots
+   Charles Darwin did all the work. The others stole the glory.
 
@@
   \ifthenelse{\boolean{publ}}{\small}{}
-  Text for this section \ldots
+  Thanks to everyone.
 
@@
  \bibliographystyle{bmc_article}  % Style BST file
-  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )
+  \bibliography{paper} }     % Bibliography file (usually '*.bib' )
 

### 2011-03-02 paper.tex
Initial start on software table

@@
 \section*{Tables}
-  \subsection*{Table 1 - Sample table title}
-    Here is an example of a \emph{small} table in \LaTeX\ using  
-    \verb|\tabular{...}|. This is where the description of the table 
-    should go. \par \mbox{}
+  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
+    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
+    \par \mbox{}
     \par
@@
       \begin{tabular}{|c|c|c|}
-        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        A1 & B2  & C3 \\ \hline
+        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
+        \hline Name & Website & Description? or Lead Developer? \\ \hline
+        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
+        Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
+        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
         A2 & ... & .. \\ \hline

### 2011-03-02 paper.tex
Bit more on software table

@@
   \subsection*{Progress}
+
 Progress to date...
@@
         Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
+        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
+        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
         Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
-        A2 & ... & .. \\ \hline
+        RDKit & http://rdkit.org & Greg Landrum \\ \hline
+        \multicolumn{3}{|c|}{Interconversion} \\ \hline
+        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
         A3 & ..  & .  \\ \hline

### 2011-03-02 paper.tex
Something about applications building on toolkits

@@
 
+...something about toolkits...
+
+Although feature-rich and robust cheminformatics toolkits are useful in and of themselves, they can also be seen as providing a base layer on which tools and applications can be built.
+Now that these toolkits have reached an acceptable level of stability and functionality, an increasing amount of software is being developed using these toolkits.
+As an example of the interaction between the commercial and open source worlds, Silicos (a Belgian company that provides services in the area of cheminformatics XXXX rephrase), has released several command line applications based on Open Babel; Pharao for pharmacophore generation (Sept 2010?), Sieve for filtering by molecular property, .....
+
+It is worth noting that Pharao is the first Open Source pharmacophore program (\emph{Is this a reasonable statement?) (although pharmacophore-related functionality has been available in the CDK and RDKit for some time).
+
+The availability of feature-rich and robust cheminformatics toolkits has started to bear fruit in the development of higher-level tools and applications that are built using these toolkits.
+
 Progress to date...

### 2011-03-02 paper.tex
More on OSOS

@@
 
-It is worth noting that Pharao is the first Open Source pharmacophore program (\emph{Is this a reasonable statement?) (although pharmacophore-related functionality has been available in the CDK and RDKit for some time).
+It is worth noting that Pharao is the first Open Source pharmacophore program (\emph{Is this a reasonable statement?}) (although pharmacophore-related functionality has been available in the CDK and RDKit for some time).
 
-The availability of feature-rich and robust cheminformatics toolkits has started to bear fruit in the development of higher-level tools and applications that are built using these toolkits.
+  \subsection*{Remaining challenges}
 
-Progress to date...
+Accuracy.
+Very often software work at the 95\% to 98\% level.
+The variety of chemical structures is such that with a large enough dataset, 'unusual' structures are always found which may be mishandled by software.
+Given that much of the development of open source software is unfunded, and relies heavily on developer motivation, it is understandable that working on the final N\% of problem structures (which may require substantial work) is not the most exciting of tasks.
+Putting a bounty system in place may be useful for these situations where a researcher needs a problem to be fixed because it affects a particular dataset of interest.
+
+Performance. There is a famous quote by Knuth that "Premature optimization is root of all evil (or something)". Given that compute time is cheap, ....
+
+Databases?
+
+\section*{Open Standards}
+  \subsection*{Progress}
+
+OpenSMILES.
+
+InChI
+
+CML?
+
   \subsection*{Remaining challenges}
-What's left to do?
+
+OpenSMARTS.
+
+Structure Normalisation
+
+\section*{Open Data}
+  \subsection*{Progress}
+  \subsection*{Remaining challenges}
 

### 2011-03-04 paper.tex
Test commit
Added test on struct stds

@@
 \section*{Background}
-The Blue Obelisk group was established in 200X at the 2XXth National Meeting of the American Chemistry Society as a response to the lack of open data, open standards and open source (ODOSOS) in chemistry.
-While other scientific disciplines such as physics, biology and astronomy (to name a few) were embracing new ways of doing science and reaping the benefits of community efforts, there was little if any innovation in the field of chemistry and scientific progress was actively hampered by the lack of access to data and tools.
+The Blue Obelisk group was established in 2005 at the
+229\textsuperscript{th} National Meeting of the American Chemistry
+Society as a response to the lack of open data, open standards and
+open source (ODOSOS) in chemistry. While other scientific disciplines
+such as physics, biology and astronomy (to name a few) were embracing
+new ways of doing science and reaping the benefits of community
+efforts, there was little if any innovation in the field of chemistry
+and scientific progress was actively hampered by the lack of access to
+data and tools.
 
@@
 
+Pistoia Alliance (and others?) should be mentioned at this point
+
   \subsection*{Remaining challenges}
@@
 
-Structure Normalisation
+A core requirement for chemical structure databases and chemical
+registration systems in general is the notion of structure
+standardization.  That is,  for a given input structure, multiple
+representations should be converted to one canonical form. 
+Structure canonicalization routines partially address this aspect,
+converting multiple alternative topologies to a single canonical
+form. However, the problem of standardization is broader than just
+topological canonicalization. Features that must be considered include
+\begin{itemize}
+\item topological canonicalization
+\item handling of charges
+\item tautomer enumeration and canonicalization
+\item normalization of functional groups
+\end{itemize}
+Currently, most of the individual components of a ``standardization
+pipeline'' can be implemented using BO tools. The larger problem is
+that there is no agreed upon list of steps for a standardization
+process. While some specifications have been published (e.g., Pubchem)
+and some standardization services and tools are available (Pubchem
+provides an online service to standaridize molecules and the NCGC
+provides a stand alone tool) each group has their own set of rules. A
+common reference specification for standardization would be of immense
+value in interoperability between structure repositories as well as
+between toolkits (though the latter is still confounded by differences
+in lower level cheminformatic features such as aromaticity models).
 

### 2011-03-04 paper.tex
First para on OpenSMILES

@@
 
-OpenSMILES.
+    \subsubsection*{OpenSMILES}
+
+One of the most widely used ways to store chemical structures is the SMILES format (or SMILES string).
+This is a linear notation depicted by Daylight Information Systems that describes the connection table of a molecule and may optionally encode chirality.
+Its popularity stems from the fact that it is a compact representation of the chemical structure that is human readable and writable, and is convenient to manipulate (e.g.
+to include in spreadsheets, or copy from a Wikipedia article).
 

### 2011-03-08 paper.tex
Added some text on collaboration between projects in Open Source

@@
 
+  \subsubsection*{Collaboration and interoperability}
+
+One of the effects of the Blue Obelisk has been to bring developers together from different Open Source chemistry projects so that they look for opportunities to collaborate rather than compete, and to leverage work done by other projects to avoid duplication of effort.
+
+As an example of this, when in March 2008 the Jmol development team were looking to add support for energy minimisation, rather than implement a forcefield from scratch they ported the UFF forcefield implementation from Open Babel to Jmol.
+This code has allowed Jmol to support 2D to 3D conversion of structures (through energy minimisation).
+
+In the area of cheminformatics tookits, two of the existing toolkits Open Babel and RDKit are planning to work together on a common underlying framework called MolCore.
+This project is still in the planning stage, but if it is a success it will mean that the the two libraries will be interoperable (while retaining their existing focus) but also that the cost of maintaining the code will be shared among more developers, freeing time for the development of new features.
+
+One of the goals of the Blue Obelisk is to promote interoperability in chemical informatics.
+When barriers exist to moving chemical data between different software, the community becomes fragmented and there is the danger of vendor lock-in (where users are constrained to using a particular software, a situation which puts them at a disadvantage).
+
+...cinfony...
+
   \subsection*{Remaining challenges}

### 2011-03-08 paper.tex
Added some text on toolkit progress

@@
 
-...something about toolkits...
+Open Source toolkits for cheminformatics have now existed for nearly
+ten years. During this period, some toolkits were developed form
+scratch, whereas others were made Open Source by releasing in-house
+codebases under liberal licenses. The primary toolkits currently in
+active development are the Chemistry Development Kit, OpenBabel and
+RDKit. These projects support a variety of languages natively (Java,
+C++, Python), but by virtue of language wrappers, they can easily be
+used from languages that they were not originally developed in. For
+example, the CDK can be accessed from Python and OpenBabel can be
+accesed from Java via SWIG wrappers. One important aspect of the
+toolkits noted here is that they originated for the support of desktop
+and server-side applications.
 
-Although feature-rich and robust cheminformatics toolkits are useful in and of themselves, they can also be seen as providing a base layer on which tools and applications can be built.
-Now that these toolkits have reached an acceptable level of stability and functionality, an increasing amount of software is being developed using these toolkits.
-As an example of the interaction between the commercial and open source worlds, Silicos (a Belgian company that provides services in the area of cheminformatics XXXX rephrase), has released several command line applications based on Open Babel; Pharao for pharmacophore generation (Sept 2010?), Sieve for filtering by molecular property, .....
+More recently, a number of workers have developed light-weight
+toolkits that are targeted towards web applications. One example of
+this form is the Chemdoodle library from XXX. The focus of these
+toolkits is to support client-side cheminformatics functionality,
+allowing for rich, chemically-aware browser-based applications.
 
-It is worth noting that Pharao is the first Open Source pharmacophore program (\emph{Is this a reasonable statement?}) (although pharmacophore-related functionality has been available in the CDK and RDKit for some time).
+In terms of functionality, most Open Source libraries support commonly
+ued cheminformatics features. At the same time, individual toolkits
+have their own strenghts. For example, the CDK has an extensive set of
+molecular descriptors useful for QSAR modeling applictions, but lacks
+support for 3D structure generation. In contrast OpenBabel, supports
+3D structure generation and optimization using a variety of
+forcefields, but lacks a built in set of molecular descriptors. RDKit
+on the other hand supports both features and in general is the most
+comprehensive Open Source toolkit in terms of functionality. Table XXX
+summarizes the features of these three toolkits.
+
+Although feature-rich and robust cheminformatics toolkits are useful
+in and of themselves, they can also be seen as providing a base layer
+on which tools and applications can be built. Now that these toolkits
+have reached an acceptable level of stability and functionality, an
+increasing amount of software is being developed using these
+toolkits. A number of Open Source applications including Bioclipse,
+XXX have been developed.
+
+More interestingly, commercial groups have released Open Source tools
+based on these toolkits. An recent example of such an interaction 
+is Silicos (a Belgian company that provides services
+in the area of cheminformatics XXXX rephrase), has released several
+command line applications based on Open Babel. For example, the Pharao
+tool released by them is a comprehensive solution for pharmacophore
+searching that provides extensive support for a variety of
+pharmacophore searches. Other tools released by them include Seive for
+fitering by molecular property and XXX. Another example of Open Source
+tools originating from commercial groups is the ChemCraft tool from
+Molecular Networks GmBH, that does XXX.
 

### 2011-03-09 paper.tex
Companies and open source

@@
 
+\subsubsection*{Second-generation tools}
+
 Although feature-rich and robust cheminformatics toolkits are useful
 in and of themselves, they can also be seen as providing a base layer
-on which tools and applications can be built. Now that these toolkits
-have reached an acceptable level of stability and functionality, an
-increasing amount of software is being developed using these
-toolkits. A number of Open Source applications including Bioclipse,
-XXX have been developed.
+on which additional tools and applications can be built.
+This is one of the reasons that cheminformatics toolkits are so important to the open source `ecosystem'; their availability lowers the barrier for the development of a `second generation' of chemistry software that no longer needs to concern itself with the low-level details of manipulating chemical structures, and can focus on providing additional functionality and ease-of-use.
 
-More interestingly, commercial groups have released Open Source tools
-based on these toolkits. An recent example of such an interaction 
-is Silicos (a Belgian company that provides services
-in the area of cheminformatics XXXX rephrase), has released several
-command line applications based on Open Babel. For example, the Pharao
+Bioclipse (v1.0 released in Aug 2006 - ????within the scope of the paper???) and Open Babel (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively.
+Bioclipse is an award-winning `molecular workbench' for life sciences that ..., while Avogadro is a 3D molecular editor and viewer aimed at preparing and analysis computational chemistry calculations.
+An interesting aspect of both of these projects is that they share some developers with the underlying toolkit and this has driven the development of new features in the CDK and Open Babel (???
+is this true for the CDK too?
+???).
+
+Some other recent projects that build on Blue Obelisk software include AMBIT (a GUI that facilitates registration of chemicals for the REACH EU directive on toxicity, based on the CDK), etc.
+
+\subsubsection*{The business end}
+
+Developing a business model based on open source software generally relies on providing support, customisation and services around the software.
+
+(Insert about eMolecules and Open Babel SMILES code...)
+
+A recent example of such an interaction 
+is Silicos, a Belgian company that provides services
+in the area of cheminformatics, which has released several
+command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
 tool released by them is a comprehensive solution for pharmacophore
@@
 pharmacophore searches. Other tools released by them include Seive for
-fitering by molecular property and XXX. Another example of Open Source
+fitering by molecular property and XXX.
+
+Another example of Open Source
 tools originating from commercial groups is the ChemCraft tool from
 Molecular Networks GmBH, that does XXX.
+
+Rather than releasing new tools, an alternative approach to provide an interface to existing tools.
+hBar Solutions has developed an online portal, hBar Lab, for managing and performing computational chemistry calculations in the cloud.
+To do so it leverages two Blue Obelisk projects, Jmol and Open Babel, as well as the open source quantum mechanics packages MPQC.
 

### 2011-03-10 paper.tex
More on business

@@
 
-Developing a business model based on open source software generally relies on providing support, customisation and services around the software.
+Although the software developed by the Blue Obelisk is Open Source on principle, this does not rule out the establishment of a business based on open source chemistry software.
+Rather than supporting itself through sales, such a business generally relies on providing support, customisation and services around the software.
+The interaction between an open source project and a company is important, and works best where there is some form of ``giving back" to the project (such as developer time, code contributions, or finanical assistance with server costs).
 
-(Insert about eMolecules and Open Babel SMILES code...)
-
-A recent example of such an interaction 
-is Silicos, a Belgian company that provides services
-in the area of cheminformatics, which has released several
-command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
+An example of such an interaction was the donation by eMolecules, Inc.
+to Open Babel of code for the canonicalisation of molecules and fragments (Nov 2006).
+eMolecules is an online vendor of chemicals that uses Open Babel under-the-hood to manage its compound collections.
+Another more recent example occurred in July 2010 when Silicos, a Belgian company that provides services
+in the area of cheminformatics, released several command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
 tool released by them is a comprehensive solution for pharmacophore

### 2011-03-10 paper.tex
InChI

@@
 
+The IUPAC InChI identifier is a non-proprietary and unique identifier for chemical substances designed to enable linking of diverse data compilations.
+Although its development predates the Blue Obelisk, software such as Open Babel has included InChI support since 2005.
+Since the official InChI implementation is in C, it is difficult to access from the other widely used language for cheminformatics toolkits, Java.
+The Blue Obelisk project JNI-InChI has been set up to solve this problem by using the Java Native Interface to link the InChI binary to Java.
+In this way, it promotes the wider adoption of this standard identifier by the chemistry community.
+
     \subsubsection*{OpenSMILES}

### 2011-03-11 paper.tex
Conclusions

@@
 
-InChI
-
 CML?
@@
   \subsection*{Remaining challenges}
-
-OpenSMARTS.
 
@@
 
+We have already discussed the development of an Open SMILES standard.
+While much progress has been made towards a complete specification, more remains to be done before this can be considered finished.
+After that point, the next logical step would be to start work on a standard for the SMARTS language, the extension to SMILES that specifies patterns that match chemical substructures.
+
 \section*{Open Data}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Results and Discussion %%
-%%
-\section*{Results and Discussion}
-  \subsection*{Results sub-heading}
-    \subsubsection*{This is a sub-sub-heading}
-      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
-      pb at end of lines ensures correct paragraph spacing.\pb
-	  Text for this sub-sub-section \ldots
-    \subsubsection*{Another sub-sub-sub-heading}
-      Text for this sub-sub-section \ldots
-
-  \subsection*{Another results sub-heading}
-    Text for this sub-section \ldots
-
-  \subsection*{Yet another results sub-heading}
-    Text for this sub-section.  More results \ldots
-
 %%%%%%%%%%%%%%%%%%%%%%
 \section*{Conclusions}
-  Text for this section \ldots
+
+We have shown that the Blue Obelisk has been very successful
+in bringing together researchers and developers with common interests
+in ODOSOS, leading to development of many useful resources freely
+available to the chemistry community. But how best to engage with the
+wider chemistry community outside of the Blue Obelisk remains an open
+question. If the Blue Obelisk is truly to make an impact,
+then an attempt must be made to reach beyond the subscribers to the
+BO mailing list and blogs of members.
+
+One recent development
+that attempts to address this issue was the establishment in Apr 2010 of a
+question-and-answer
+website related to Blue Obelisk projects and themes at
+http://blueobelisk.shapado.com. This is a website in the
+style of Stack Overflow that encourages high quality answers (and
+questions) through the use of a voting system. In the year since it
+was established, over
+200 users have registered, many of whom had no previous involvement
+with the Blue Obelisk.
+
+We hope to see this involvement between the Blue Obelisk and the wider
+community grow in future. To this end, we encourage the reader to
+visit http://blueobelisk.org, send a message to our mailing list,
+investigate related projects or read our blogs.
 

### 2011-03-14 paper.tex
Info on ChemiSQL

@@
 One of the effects of the Blue Obelisk has been to bring developers together from different Open Source chemistry projects so that they look for opportunities to collaborate rather than compete, and to leverage work done by other projects to avoid duplication of effort.
-
 As an example of this, when in March 2008 the Jmol development team were looking to add support for energy minimisation, rather than implement a forcefield from scratch they ported the UFF forcefield implementation from Open Babel to Jmol.
 This code has allowed Jmol to support 2D to 3D conversion of structures (through energy minimisation).
+
+Another collaborative initiative between Blue Obelisk projects was the establishment in May 2008 of
+the ChemiSQL project. This brought together the developers of several
+open source chemistry database cartridges (PgChem, MyChem, OrChem and
+more recently Bingo) with a view to making their database APIs more
+similar and collaborating on benchmark datasets for assessing
+performance. For two of these projects, PgChem and MyChem, which are both based on
+Open Babel, there is the additional possibility of working together on a shared
+codebase.
 

### 2011-03-15 paper.tex
Cinfony

@@
 
-One of the goals of the Blue Obelisk is to promote interoperability in chemical informatics.
-When barriers exist to moving chemical data between different software, the community becomes fragmented and there is the danger of vendor lock-in (where users are constrained to using a particular software, a situation which puts them at a disadvantage).
-
-...cinfony...
+One of the goals of the Blue Obelisk is to promote interoperability in chemical
+informatics. When barriers exist to moving chemical data between
+different software, the community becomes fragmented and there is
+the danger of vendor lock-in (where users are constrained to using
+a particular software, a situation which puts them at a
+disadvantage). This applies as much to Open Source software as to
+proprietary software. Cinfony is a project (first release May 2008)
+whose goal is to tackle this problem in the area of cheminformatics toolkits.
+It is a Python library that enables Open Babel, the CDK, and RDKit to
+be used using the same API; this makes it easy, for example, to read a
+molecule using Open Babel, calculate descriptors using the CDK and
+create a depiction using RDKit.
 

### 2011-03-15 paper.tex
More on intro, plus suggestion on other areas of activity

@@
 
-The goals of the Blue Obelisk group were described in Guha et al.\cite{guha2006}. but are briefly summarised here...
+The formation of the Blue Obelisk group is somewhat unusual in that it
+is not a funded network, nor does it follow the industry consortium
+model. Rather it is a grassroots organisation, catalysed by an initial
+core of interested scientists, but with membership open to all who
+share one or more of the goals of the group:
+\begin{itemize}
+\item Open Source. One can use other people's code without further permission, including changing it for one's own use and distributing it again.
+\item Open Standards.
+One can find visible community mechanisms for protocols and communicating information.
+The mechanisms for creating and maintaining these standards cover a wide spectrum of human organizations, including various degrees of consent.
+\item Open Data. One can obtain all data in the public domain when wanted and reuse it for whatever purpose.
+\end{itemize}
+
+The Blue Obelisk was first described in Guha et al.\cite{guha2006}. This
+contribution looks back on the past 5 years and surveys progress and
+remaining challenges in the areas of Open Source, Open Data and Open
+Standards in chemistry.
 
@@
 toolkits that are targeted towards web applications. One example of
-this form is the Chemdoodle library from XXX. The focus of these
+this form is the ChemDoodle library from iChemLabs. The focus of these
 toolkits is to support client-side cheminformatics functionality,
@@
 
+\section*{Other areas of activity}
+Web services? (NIH, CDK and Chemspider)
+
+Books (Angel, Egon, OB)
+
 %%%%%%%%%%%%%%%%%%%%%%

### 2011-04-13 paper.tex
Change title to include a description of the Blue Obelisk. I think this is better if people don't know what the BO is in the first place.
Add Figure 1 - The Blue Obelisk logo
Finish off OpenSMILES

@@
 
-\title{The Blue Obelisk five years on}
+\title{Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on}
 
@@
 
-One of the most widely used ways to store chemical structures is the SMILES format (or SMILES string).
-This is a linear notation depicted by Daylight Information Systems that describes the connection table of a molecule and may optionally encode chirality.
-Its popularity stems from the fact that it is a compact representation of the chemical structure that is human readable and writable, and is convenient to manipulate (e.g.
-to include in spreadsheets, or copy from a Wikipedia article).
+One of the most widely used ways to store chemical structures is the
+SMILES format (or SMILES string). This is a linear notation depicted
+by Daylight Information Systems that describes the connection table
+of a molecule and may optionally encode chirality. Its popularity
+stems from the fact that it is a compact representation of the
+chemical structure that is human readable and writable, and is
+convenient to manipulate (e.g. to include in spreadsheets, or copy
+from a Wikipedia article).
+
+Despite its widespread use, a formal
+definition of the language did not exist beyond Daylight's SMILES
+Theory Manual and tutorials. This caused some confusion in the
+implementation and interpretation of corner cases, for example the
+handling of cis/trans bond symbols at ring closures. In 2007, Craig
+James (eMolecules) initiated work on the OpenSMILES specification, a
+complete specification of the SMILES language as an Open Standard
+developed through a community process. The specification is largely
+complete and contains guidelines on reading SMILES, a formal
+grammar, recommendations on standard forms when writing SMILES, as
+well as proposed extensions.
 
@@
 \section*{Figures}
-  \subsection*{Figure 1 - Sample figure title}
-      A short description of the figure content
-      should go here.
+  \subsection*{Figure 1 - Blue Obelisk logo}
 

### 2011-04-14 paper.tex
Panton Principles

@@
   \subsection*{Progress}
+
+A considerable stumbling block in advocating the release of scientific
+data as Open Data has been how exactly to define Open. A major step
+forward was the launch in 2010 of the Panton Principles for Open Data
+in Science. This formalises the idea that Open Data maximises the
+possibility of reuse and repurposing, the fundamental basis
+of how science works. These principles recommend that published data
+be licensed explicitly, and preferably under CC0 (Creative Commons "No
+Rights Reserved"). This license allows others to use the data for any
+purpose whatsoever without any barriers.
+
   \subsection*{Remaining challenges}

### 2011-04-14 paper.tex
Adding references

@@
 
-The Blue Obelisk was first described in Guha et al.\cite{guha2006}. This
+The Blue Obelisk was first described in Guha et al.\cite{Guha2006} in
+2006. This
 contribution looks back on the past 5 years and surveys progress and
@@
 
-Bioclipse (v1.0 released in Aug 2006 - ????within the scope of the paper???) and Open Babel (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively.
+Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
+\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively.
 Bioclipse is an award-winning `molecular workbench' for life sciences that ..., while Avogadro is a 3D molecular editor and viewer aimed at preparing and analysis computational chemistry calculations.
@@
 Rather than supporting itself through sales, such a business generally relies on providing support, customisation and services around the software.
-The interaction between an open source project and a company is important, and works best where there is some form of ``giving back" to the project (such as developer time, code contributions, or finanical assistance with server costs).
+The interaction between an open source project and a company is important, and works best where there is some form of `giving back' to the project (such as developer time, code contributions, or finanical assistance with server costs).
 
@@
 disadvantage). This applies as much to Open Source software as to
-proprietary software. Cinfony is a project (first release May 2008)
-whose goal is to tackle this problem in the area of cheminformatics toolkits.
+proprietary software. Cinfony is a project (first release in May 2008)
+whose goal is to tackle this problem in the area of cheminformatics
+toolkits \cite{OBoyleCinfony2008}.
 It is a Python library that enables Open Babel, the CDK, and RDKit to
@@
 \end{itemize}
-Currently, most of the individual components of a ``standardization
-pipeline'' can be implemented using BO tools. The larger problem is
+Currently, most of the individual components of a `standardization
+pipeline' can be implemented using BO tools. The larger problem is
 that there is no agreed upon list of steps for a standardization
@@
 forward was the launch in 2010 of the Panton Principles for Open Data
-in Science. This formalises the idea that Open Data maximises the
+in Science \cite{WebPanton}. This formalises the idea that Open Data maximises the
 possibility of reuse and repurposing, the fundamental basis
 of how science works. These principles recommend that published data
-be licensed explicitly, and preferably under CC0 (Creative Commons "No
-Rights Reserved"). This license allows others to use the data for any
+be licensed explicitly, and preferably under CC0 (Creative Commons `No
+Rights Reserved') \cite{WebCC0}. This license allows others to use the data for any
 purpose whatsoever without any barriers.
@@
  \bibliographystyle{bmc_article}  % Style BST file
-  \bibliography{paper} }     % Bibliography file (usually '*.bib' )
+  \bibliography{websites,paper} }     % Bibliography file (usually '*.bib' )
 

### 2011-04-14 paper.tex
Fix spellings

@@
 codebases under liberal licenses. The primary toolkits currently in
-active development are the Chemistry Development Kit, OpenBabel and
+active development are the Chemistry Development Kit, Open Babel and
 RDKit. These projects support a variety of languages natively (Java,
@@
 used from languages that they were not originally developed in. For
-example, the CDK can be accessed from Python and OpenBabel can be
-accesed from Java via SWIG wrappers. One important aspect of the
+example, the CDK can be accessed from Python and Open Babel can be
+accessed from Java via SWIG wrappers. One important aspect of the
 toolkits noted here is that they originated for the support of desktop
@@
 In terms of functionality, most Open Source libraries support commonly
-ued cheminformatics features. At the same time, individual toolkits
-have their own strenghts. For example, the CDK has an extensive set of
-molecular descriptors useful for QSAR modeling applictions, but lacks
-support for 3D structure generation. In contrast OpenBabel, supports
+used cheminformatics features. At the same time, individual toolkits
+have their own strengths. For example, the CDK has an extensive set of
+molecular descriptors useful for QSAR modeling applications, but lacks
+support for 3D structure generation. In contrast, Open Babel supports
 3D structure generation and optimization using a variety of
@@
 searching that provides extensive support for a variety of
-pharmacophore searches. Other tools released by them include Seive for
-fitering by molecular property and XXX.
+pharmacophore searches. Other tools released by them include Sieve for
+filtering by molecular property and XXX.
 
@@
 and some standardization services and tools are available (Pubchem
-provides an online service to standaridize molecules and the NCGC
+provides an online service to standardise molecules and the NCGC
 provides a stand alone tool) each group has their own set of rules. A

### 2011-04-15 paper.tex
More references
Update section on cheminformatics toolkit. Add stubs for Indigo and RDKit. Remove comparison of toolkits as not relevant.

@@
 
+\subsection*{Cheminformatics toolkits}
+
 Open Source toolkits for cheminformatics have now existed for nearly
-ten years. During this period, some toolkits were developed form
-scratch, whereas others were made Open Source by releasing in-house
-codebases under liberal licenses. The primary toolkits currently in
-active development are the Chemistry Development Kit, Open Babel and
-RDKit. These projects support a variety of languages natively (Java,
-C++, Python), but by virtue of language wrappers, they can easily be
-used from languages that they were not originally developed in. For
-example, the CDK can be accessed from Python and Open Babel can be
-accessed from Java via SWIG wrappers. One important aspect of the
-toolkits noted here is that they originated for the support of desktop
-and server-side applications.
+ten years. During this period, some toolkits were developed from
+scratch in academia, whereas others were made Open Source by releasing in-house
+codebases under liberal licenses. When the Blue Obelisk was
+established five years ago, the primary toolkits under active development
+were the Chemistry Development Kit (CDK)
+\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpenBabel},
+and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
+continue to be actively developed. [Insert main focus of work on CDK
+in last 5 years] Since 2006, major new features of Open Babel include 3D structure
+generation and 2D structure-diagram generation, UFF and MMFF94
+forcefields, and significantly expanded support for computational
+chemistry calculations. In addition, a major focus of Open Babel development
+has been to fix problems in the areas of stereochemistry, kekulisation
+and canonicalisation.
 
-More recently, a number of workers have developed light-weight
+Two new Open Source cheminformatics toolkits have appeared since the
+original paper. In 2006 Rational Discovery, a cheminformatics service
+company (since closed down), released RDKit \cite{WebRDKit} under the
+BSD License. RDKit continues to be actively developed and includes
+code donated by Novartis. More recently, GGA Software Services (a
+contract programming company) released the Indigo toolkit
+\cite{WebIndigo} and associated software in 2009 under the GPL. [More
+text on Indigo here]
+
+One important aspect of the above
+toolkits is that they originated for the support of desktop
+and server-side applications. More recently, a number of workers have developed light-weight
 toolkits that are targeted towards web applications. One example of
@@
 allowing for rich, chemically-aware browser-based applications.
-
-In terms of functionality, most Open Source libraries support commonly
-used cheminformatics features. At the same time, individual toolkits
-have their own strengths. For example, the CDK has an extensive set of
-molecular descriptors useful for QSAR modeling applications, but lacks
-support for 3D structure generation. In contrast, Open Babel supports
-3D structure generation and optimization using a variety of
-forcefields, but lacks a built in set of molecular descriptors. RDKit
-on the other hand supports both features and in general is the most
-comprehensive Open Source toolkit in terms of functionality. Table XXX
-summarizes the features of these three toolkits.
 

### 2011-04-15 paper.tex
Minor fix to heading
Fix column width to 70. I have learned the magic of 'set textwidth=70' in vim.

@@
 \begin{itemize}
-\item Open Source. One can use other people's code without further permission, including changing it for one's own use and distributing it again.
-\item Open Standards.
-One can find visible community mechanisms for protocols and communicating information.
-The mechanisms for creating and maintaining these standards cover a wide spectrum of human organizations, including various degrees of consent.
-\item Open Data. One can obtain all data in the public domain when wanted and reuse it for whatever purpose.
+\item Open Source. One can use other people's code without further
+permission, including changing it for one's own use and distributing
+it again.
+\item Open Standards. One can find visible community mechanisms for
+protocols and communicating information. The mechanisms for creating
+and maintaining these standards cover a wide spectrum of human
+organizations, including various degrees of consent.
+\item Open Data. One can obtain all data in the public domain when
+wanted and reuse it for whatever purpose.
 \end{itemize}
@@
 
-\subsection*{Cheminformatics toolkits}
+\subsubsection*{Cheminformatics toolkits}
 
@@
 in and of themselves, they can also be seen as providing a base layer
-on which additional tools and applications can be built.
-This is one of the reasons that cheminformatics toolkits are so important to the open source `ecosystem'; their availability lowers the barrier for the development of a `second generation' of chemistry software that no longer needs to concern itself with the low-level details of manipulating chemical structures, and can focus on providing additional functionality and ease-of-use.
+on which additional tools and applications can be built. This is one
+of the reasons that cheminformatics toolkits are so important to the
+open source `ecosystem'; their availability lowers the barrier for the
+development of a `second generation' of chemistry software that no
+longer needs to concern itself with the low-level details of
+manipulating chemical structures, and can focus on providing
+additional functionality and ease-of-use.
 
 Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
-\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively.
-Bioclipse is an award-winning `molecular workbench' for life sciences that ..., while Avogadro is a 3D molecular editor and viewer aimed at preparing and analysis computational chemistry calculations.
-An interesting aspect of both of these projects is that they share some developers with the underlying toolkit and this has driven the development of new features in the CDK and Open Babel (???
-is this true for the CDK too?
-???).
+\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such
+software, based on the CDK and Open Babel, respectively. Bioclipse is
+an award-winning `molecular workbench' for life sciences that ...,
+while Avogadro is a 3D molecular editor and viewer aimed at preparing
+and analysing computational chemistry calculations. An interesting
+aspect of both of these projects is that they share some developers
+with the underlying toolkit and this has driven the development of new
+features in the CDK and Open Babel.
 
-Some other recent projects that build on Blue Obelisk software include AMBIT (a GUI that facilitates registration of chemicals for the REACH EU directive on toxicity, based on the CDK), etc.
+Some other recent projects that build on Blue Obelisk software include
+AMBIT (a GUI that facilitates registration of chemicals for the REACH
+EU directive on toxicity, based on the CDK), etc.
 
@@
 
-Although the software developed by the Blue Obelisk is Open Source on principle, this does not rule out the establishment of a business based on open source chemistry software.
-Rather than supporting itself through sales, such a business generally relies on providing support, customisation and services around the software.
-The interaction between an open source project and a company is important, and works best where there is some form of `giving back' to the project (such as developer time, code contributions, or finanical assistance with server costs).
+Although the software developed by the Blue Obelisk is Open Source on
+principle, this does not rule out the establishment of a business
+based on open source chemistry software. Rather than supporting itself
+through sales, such a business generally relies on providing support,
+customisation and services around the software. The interaction
+between an open source project and a company is important, and works
+best where there is some form of `giving back' to the project (such as
+developer time, code contributions, or financial assistance with server costs).
 
 An example of such an interaction was the donation by eMolecules, Inc.
-to Open Babel of code for the canonicalisation of molecules and fragments (Nov 2006).
-eMolecules is an online vendor of chemicals that uses Open Babel under-the-hood to manage its compound collections.
-Another more recent example occurred in July 2010 when Silicos, a Belgian company that provides services
-in the area of cheminformatics, released several command line applications based on Open Babel as well as donating code to the project. For example, the Pharao
+to Open Babel of code for the canonicalisation of molecules and
+fragments (Nov 2006). eMolecules is an online vendor of chemicals that
+uses Open Babel under-the-hood to manage its compound collections.
+Another more recent example occurred in July 2010 when Silicos, a
+Belgian company that provides services
+in the area of cheminformatics, released several command line
+applications based on Open Babel as well as donating code to the
+project. For example, the Pharao
 tool released by them is a comprehensive solution for pharmacophore
@@
 
-Rather than releasing new tools, an alternative approach to provide an interface to existing tools.
-hBar Solutions has developed an online portal, hBar Lab, for managing and performing computational chemistry calculations in the cloud.
-To do so it leverages two Blue Obelisk projects, Jmol and Open Babel, as well as the open source quantum mechanics packages MPQC.
+Rather than releasing new tools, an alternative approach to provide an
+interface to existing tools. hBar Solutions has developed an online
+portal, hBar Lab, for managing and performing computational chemistry
+calculations in the cloud. To do so it leverages two Blue Obelisk
+projects, Jmol and Open Babel, as well as the open source quantum
+mechanics packages MPQC.
 
@@
 
-One of the effects of the Blue Obelisk has been to bring developers together from different Open Source chemistry projects so that they look for opportunities to collaborate rather than compete, and to leverage work done by other projects to avoid duplication of effort.
-As an example of this, when in March 2008 the Jmol development team were looking to add support for energy minimisation, rather than implement a forcefield from scratch they ported the UFF forcefield implementation from Open Babel to Jmol.
-This code has allowed Jmol to support 2D to 3D conversion of structures (through energy minimisation).
+One of the effects of the Blue Obelisk has been to bring developers
+together from different Open Source chemistry projects so that they
+look for opportunities to collaborate rather than compete, and to
+leverage work done by other projects to avoid duplication of effort.
+As an example of this, when in March 2008 the Jmol development team
+were looking to add support for energy minimisation, rather than
+implement a forcefield from scratch they ported the UFF forcefield
+implementation from Open Babel to Jmol. This code has allowed Jmol to
+support 2D to 3D conversion of structures (through energy
+minimisation).
 
@@
 
-In the area of cheminformatics tookits, two of the existing toolkits Open Babel and RDKit are planning to work together on a common underlying framework called MolCore.
-This project is still in the planning stage, but if it is a success it will mean that the the two libraries will be interoperable (while retaining their existing focus) but also that the cost of maintaining the code will be shared among more developers, freeing time for the development of new features.
+In the area of cheminformatics tookits, two of the existing toolkits
+Open Babel and RDKit are planning to work together on a common
+underlying framework called MolCore. This project is still in the
+planning stage, but if it is a success it will mean that the the two
+libraries will be interoperable (while retaining their existing focus)
+but also that the cost of maintaining the code will be shared among
+more developers, freeing time for the development of new features.
 
@@
 
-Accuracy.
-Very often software work at the 95\% to 98\% level.
-The variety of chemical structures is such that with a large enough dataset, 'unusual' structures are always found which may be mishandled by software.
-Given that much of the development of open source software is unfunded, and relies heavily on developer motivation, it is understandable that working on the final N\% of problem structures (which may require substantial work) is not the most exciting of tasks.
-Putting a bounty system in place may be useful for these situations where a researcher needs a problem to be fixed because it affects a particular dataset of interest.
+Accuracy. Very often software work at the 95\% to 98\% level. The
+variety of chemical structures is such that with a large enough
+dataset, 'unusual' structures are always found which may be mishandled
+by software. Given that much of the development of open source
+software is unfunded, and relies heavily on developer motivation, it
+is understandable that working on the final N\% of problem structures
+(which may require substantial work) is not the most exciting of
+tasks. Putting a bounty system in place may be useful for these
+situations where a researcher needs a problem to be fixed because it
+affects a particular dataset of interest.
 
-Performance. There is a famous quote by Knuth that "Premature optimization is root of all evil (or something)". Given that compute time is cheap, ....
+Performance. There is a famous quote by Knuth that "Premature
+optimization is root of all evil (or something)". Given that compute
+time is cheap, ....
 
@@
 
-The IUPAC InChI identifier is a non-proprietary and unique identifier for chemical substances designed to enable linking of diverse data compilations.
-Although its development predates the Blue Obelisk, software such as Open Babel has included InChI support since 2005.
-Since the official InChI implementation is in C, it is difficult to access from the other widely used language for cheminformatics toolkits, Java.
-The Blue Obelisk project JNI-InChI has been set up to solve this problem by using the Java Native Interface to link the InChI binary to Java.
-In this way, it promotes the wider adoption of this standard identifier by the chemistry community.
+The IUPAC InChI identifier is a non-proprietary and unique identifier
+for chemical substances designed to enable linking of diverse data
+compilations. Although its development predates the Blue Obelisk,
+software such as Open Babel has included InChI support since 2005.
+Since the official InChI implementation is in C, it is difficult to
+access from the other widely used language for cheminformatics
+toolkits, Java. The Blue Obelisk project JNI-InChI has been set up to
+solve this problem by using the Java Native Interface to link the
+InChI binary to Java. In this way, it promotes the wider adoption of
+this standard identifier by the chemistry community.
 
@@
 We have already discussed the development of an Open SMILES standard.
-While much progress has been made towards a complete specification, more remains to be done before this can be considered finished.
-After that point, the next logical step would be to start work on a standard for the SMARTS language, the extension to SMILES that specifies patterns that match chemical substructures.
+While much progress has been made towards a complete specification,
+more remains to be done before this can be considered finished. After
+that point, the next logical step would be to start work on a standard
+for the SMARTS language, the extension to SMILES that specifies
+patterns that match chemical substructures.
 

### 2011-04-16 paper.tex
Added text to introduction. Added two references to paper.bib. Added paper.bib, which may have been a mistake.

@@
 	 \and 
+	     Christoph Steinbeck$^6$%
+	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+     \and 
          Noel M O'Boyle$^5$%
@@
     \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
-    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
+    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
+    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
 }%
@@
         % Do not use inserted blank lines (ie \\) until main body of text.
-        \paragraph*{Background:} Text for this section of the abstract.
+        \paragraph*{Background:} The Blue Obelisk movement was established in 2005 as a 
+		response to the lack of open data, open standards and
+		open source (ODOSOS) in chemistry. While other scientific disciplines
+		such as physics, biology and astronomy (to name a few) were embracing
+		new ways of doing science and reaping the benefits of community
+		efforts, there was little if any innovation in the field of chemistry
+		and scientific progress was actively hampered by the lack of access to
+		data and tools.
 
-        \paragraph*{Results:} Text for this section of the abstract \ldots
+        \paragraph*{Results:} This
+		contribution looks back on the past 5 years and surveys progress and
+		remaining challenges in the areas of Open Source, Open Data and Open
+		Standards in chemistry.
 
-        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
+        \paragraph*{Conclusions:} Here we show that the Blue Obelisk has been very successful
+		in bringing together researchers and developers with common interests
+		in ODOSOS, leading to development of many useful resources freely
+		available to the chemistry community. But how best to engage with the
+		wider chemistry community outside of the Blue Obelisk remains an open
+		question.
 \end{abstract}
@@
 \section*{Background}
-The Blue Obelisk group was established in 2005 at the
+The Blue Obelisk movement was established in 2005 at the
 229\textsuperscript{th} National Meeting of the American Chemistry
@@
 data and tools.
+
+Since 2005 it has become evident that a good amount of development in open 
+chemical information and information is driven by the demands of neighbouring
+scientific fields. In many areas in biology, for example, the importance of 
+small molecules and their interactions and reactions in biological systems
+has been realised. In fact, one of the first free and open databases and ontologies 
+of small molecules was created as a resource about chemical structure and nomenclature
+by biologists \cite{DeMatos:2009p3839}.
+Open workflow systems such as Taverna (XXX add citation) have opened the door for 
+an easy and seamless integration of FOS components into executable workflows 
+or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
+reported in 2010 \cite{Kuhn:2010p4001}.
 
@@
 and maintaining these standards cover a wide spectrum of human
-organizations, including various degrees of consent.
+organisations, including various degrees of consent.
 \item Open Data. One can obtain all data in the public domain when

### 2011-04-18 paper.tex
Added more text about CDK taverna. Added a section about databases. Added rows to BO table and added subheadings to table.

@@
 data and tools.
-
 Since 2005 it has become evident that a good amount of development in open 
@@
 by biologists \cite{DeMatos:2009p3839}.
-Open workflow systems such as Taverna (XXX add citation) have opened the door for 
-an easy and seamless integration of FOS components into executable workflows 
+Open workflow systems such as Taverna \cite{Hull:2006p60} have opened the door for 
+an easy and seamless integration of free and open software (FOS) components into executable workflows 
 or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
@@
 features in the CDK and Open Babel.
+
+CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
+Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharma- cophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
+CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
+Taverna version 2.x.
+The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.
 
@@
 
-Databases?
+\section*{Chemical Structure Registration Systems, Databases}
+Databases? (XXX CS: A couple of database-related paragraphs could go here XXX)
+
+XXX the paragraph below on OrChem needs some work
+Registration, indexing and searching of chemical structures in relational databases is one of the core areas of cheminformatics. 
+A number of structure registration systems have been published in the last five years, exploiting the fact that 
+free cheminformatics toolkits such as OpenBabel and the CDK were available. 
+OrChem, for example, is an extension for the Oracle 11G database that adds registration and indexing of chemical structures to support fast substructure and similarity searching.
+The cheminformatics functionality is provided by the Chemistry Development Kit.
+OrChem provides similarity searching with response times in the order of seconds for databases with millions of compounds, depending on a given similarity cut-off.
+For substructure searching, it can make use of multiple processor cores on today's powerful database servers to provide fast response times in equally large data sets.
+OrChem is free software and can be redistributed and/or modified under the terms of the GNU Lesser General Public License as published by the Free Software Foundation.
+All software is available via http://orchem.sourceforge.net.
 
@@
         \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
-        Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
+        Chemistry Development Kit (CDK) \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
         Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
@@
         RDKit & http://rdkit.org & Greg Landrum \\ \hline
+        \multicolumn{3}{|c|}{Integration} \\ \hline
+        CDK-Taverna \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
+        A3 & ..  & .  \\ \hline
         \multicolumn{3}{|c|}{Interconversion} \\ \hline
@@
         A3 & ..  & .  \\ \hline
+        \multicolumn{3}{|c|}{Structure Databases} \\ \hline
+        OrChem \cite{RijnbeekS10} & http://orchem.sourceforge.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
+        MyChem  & http://XXX.sourceforge.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
+        PGChem  & http://XXX.sourceforge.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
+        Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
+
       \end{tabular}

### 2011-04-18 paper.tex
Added reference to CDK News.
Edited pm286 as author

@@
      \and 
+         Peter Murray-Rust$^7$%
+         \email{Peter Murray-Rust - pm286@cam.ac.uk}
+      \and
          Jane E Doe\correspondingauthor$^2$%
@@
     \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
+    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK
 }%
@@
 
-The Blue Obelisk was first described in Guha et al.\cite{Guha2006} in
+The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
+later as a formal paper by Guha et al.\cite{Guha2006} in
 2006. This

### 2011-04-18 paper.tex
Added some brief information about Indigo, KNIME, and CurlySMILES

@@
          \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
+      \and
+         Dmitry Pavlov$^8$%
+         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
       }
@@
     \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK
+    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia
 }%
@@
 by biologists \cite{DeMatos:2009p3839}.
-Open workflow systems such as Taverna \cite{Hull:2006p60} have opened the door for 
-an easy and seamless integration of free and open software (FOS) components into executable workflows 
+Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
+have opened the door for 
+an easy and seamless integration of FOS components into executable workflows 
 or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
 reported in 2010 \cite{Kuhn:2010p4001}.
+
+KNIME currently has a built-in collection of CDK-based and
+OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
+chemistry workflows. Another collection of chemistry nodes---based on
+the RDKit and its Java wrapper---appeared in 2010 and is available to
+KNIME users from the KNIME's official ``Community Updates'' site.
+Indigo nodes for KNIME are in preparation and will appear nearby the
+RDKit nodes by May 2011.
 
@@
 BSD License. RDKit continues to be actively developed and includes
-code donated by Novartis. More recently, GGA Software Services (a
-contract programming company) released the Indigo toolkit
-\cite{WebIndigo} and associated software in 2009 under the GPL. [More
-text on Indigo here]
+code donated by Novartis.
+
+More recently, GGA Software Services (a contract programming company)
+released the Indigo toolkit \cite{WebIndigo} and associated software
+in 2009 under the GPL. Indigo is written from scratch in C++ and has
+high-level wrappers for using it in C, Java, Python, or .NET
+environment. Like other toolkits, Indigo provides support for
+tetrahedral ans cis-trans stereochemistry, 2D coordinate generation,
+exact/substructure/SMARTS matching, fingerprint generation (all these
+both for molecules and reactions), and canonical SMILES computation.
+It also provides some less common functionality, like matching
+tautomers ans resonance substructures, enumeration of subgraps,
+finding maximum common substructure of $N$ input structures, and
+enumerating reaction products. Indigo also has its rendering plugin,
+which is used in a lot of commercial and open-sourse projects as the
+main 2D chemical rendering engine.
 
@@
 
+Also, Indigo is going to provide its own plugin with an interface to
+the InChI implementation in May 2011. This will enable developers to
+calculate InChI and InChIKey-s within Python and .NET projects (not to
+mention Java), eliminating the trouble of compiling and using binary
+modules directly.
+
     \subsubsection*{OpenSMILES}
@@
 well as proposed extensions.
+
+Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
+SMILES notation, which allows to define crystals structures, polymers,
+electron delocalization charges, molecule interations, and many other
+features absent in the initial SMILES specification. There has been
+discussions about including parts of the CurlySMILES notation into
+OpenSMILES, especially polymers.
 

### 2011-04-18 paper.tex
Added a second example of between-project code migration
Pasted in Roberts contribution

@@
          \email{Peter Murray-Rust - pm286@cam.ac.uk}
+     \and
+         Robert Hanson$^8$%
+         \email{Robert Hanson - hansonr@stolaf.edu}
       \and
@@
 
+Some project have also been adopted by other fields, including biology
+and crystallography. A good example
+here is Jmol, which has, in the past five years, gone through two major revisions
+as part of an accelerated development focus. Starting with Jmol 10.2, released in
+April of 2006, Jmol 11.0 added over 100 new features and was released in March of
+2007. Three years later, Jmol 12.0 followed with over 700 new features.
+Based on the outstanding work of Michael (Miguel) Howard and others, this development
+path took Jmol from a "Rasmol/Chime replacement" to a fully fledged molecular
+visualization package, including full support for crystallography~\cite{Hanson2010},
+efficient delivery of surface data using fast solvent-excluded surface
+generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
+ display of molecular orbitals from standard basis set/coefficient data,
+the inclusion of dynamic minimization using the UFF force field, and
+a full implementation of Daylight SMILES and SMART, with extensions to
+conformational and biomolecular substructure searching in the form of Jmol
+BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
+Behind the scenes, Jmol has been essentially completely rewritten. A full
+mathematical expression evaluator was added, and the scripting language
+was folded into a versatile JavaScript-like command language, including a
+complete set of program flow commands, several new data types, and
+user-defined variables and functions. An exciting non-chemical application
+of Jmol has been its incorporation into the open source Sage
+Mathematics Project [http://www.sagemath.org/], where Jmol is being used
+to deliver three-dimensional depiction of mathematical functions.
+
 Two new Open Source cheminformatics toolkits have appeared since the
@@
 support 2D to 3D conversion of structures (through energy
-minimisation).
+minimisation). Similarly, efficient Jmol code for atom-atom rebonding
+has been ported to the CDK.
 

### 2011-04-19 paper.tex
fixed spelling errors

@@
 environment. Like other toolkits, Indigo provides support for
-tetrahedral ans cis-trans stereochemistry, 2D coordinate generation,
+tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
 exact/substructure/SMARTS matching, fingerprint generation (all these
@@
 enumerating reaction products. Indigo also has its rendering plugin,
-which is used in a lot of commercial and open-sourse projects as the
+which is used in a lot of commercial and open-source projects as the
 main 2D chemical rendering engine.
@@
 underlying framework called MolCore. This project is still in the
-planning stage, but if it is a success it will mean that the the two
+planning stage, but if it is a success it will mean that the two
 libraries will be interoperable (while retaining their existing focus)
@@
 We hope to see this involvement between the Blue Obelisk and the wider
-community grow in future. To this end, we encourage the reader to
+community grow in the future. To this end, we encourage the reader to
 visit http://blueobelisk.org, send a message to our mailing list,

### 2011-04-20 paper.tex
Added sections on Bioclipse and QSAR-ML with references.

@@
      \and 
+         Ola Spjuth$^9$%
+         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+     \and 
          Peter Murray-Rust$^7$%
@@
     \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
-    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
-    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK
-    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia
+    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
+    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
+    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
+    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden 
 }%
@@
 
-Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
+Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
 \cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such
 software, based on the CDK and Open Babel, respectively. Bioclipse is
-an award-winning `molecular workbench' for life sciences that ...,
-while Avogadro is a 3D molecular editor and viewer aimed at preparing
+an award-winning `molecular workbench' for life sciences with a dynamic plugin architectur, and which wraps cheminformatics functionality behind user-friendly interfaces and graphical editors.
+All functionality in Bioclipse is completely scriptable using an integrated high-level language called Bioclipse Scripting Language (BSL), which can be used for invoking functionality or to automate complex tasks~\cite{Bioclipse2}.
+
+Avogadro is a 3D molecular editor and viewer aimed at preparing
 and analysing computational chemistry calculations. An interesting
@@
 
-CML?
+\subsubsection*{CML}
+?
+
+\subsubsection*{QSAR-ML}
+The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses.
+QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}.
+A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation.
+Hence, a dataset described in QSAR-ML is completely reproducible.
+To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations.
+Descriptor implementations were initially available from CDK and JoeLIB, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
+
+\subsubsection*{Others...?}
 
@@
 \section*{Other areas of activity}
-Web services? (NIH, CDK and Chemspider)
+Web services? XMPP services? (NIH, CDK and Chemspider)
 

### 2011-04-20 paper.tex
Some spelling corrections.

@@
 software, based on the CDK and Open Babel, respectively. Bioclipse is
-an award-winning `molecular workbench' for life sciences with a dynamic plugin architectur, and which wraps cheminformatics functionality behind user-friendly interfaces and graphical editors.
-All functionality in Bioclipse is completely scriptable using an integrated high-level language called Bioclipse Scripting Language (BSL), which can be used for invoking functionality or to automate complex tasks~\cite{Bioclipse2}.
+an award-winning molecular workbench for life sciences with a dynamic plugin architecture, and which wraps cheminformatics functionality behind user-friendly interfaces and graphical editors.
+All functionality in Bioclipse is completely scriptable using an integrated high-level language called Bioclipse Scripting Language (BSL), which can be used for scripted research or to automate complex tasks~\cite{Bioclipse2}.
 

### 2011-04-20 paper.tex
added mentioning of Brunn as per Egon's suggestion
It goes into the Bioclipse paragraph which also got changed a little...

--- 2011-04-20 paper.tex
@@
          \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+     \and 
+         Jonathan Alvarsson$^9$%
+         \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
      \and 
@@
 Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
-\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such
-software, based on the CDK and Open Babel, respectively. Bioclipse is
-an award-winning `molecular workbench' for life sciences with a dynamic plugin architectur, and which wraps cheminformatics functionality behind user-friendly interfaces and graphical editors.
-All functionality in Bioclipse is completely scriptable using an integrated high-level language called Bioclipse Scripting Language (BSL), which can be used for invoking functionality or to automate complex tasks~\cite{Bioclipse2}.
+\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
+on the CDK and Open Babel, respectively. Bioclipse is an award-winning
+`molecular workbench' for life sciences with a dynamic plugin architecture, and
+which wraps cheminformatics functionality behind user-friendly interfaces and
+graphical editors. All functionality in Bioclipse is completely scriptable
+using an integrated high-level language called Bioclipse Scripting Language
+(BSL), which can be used for invoking functionality or to automate complex
+tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
+such as for example Brunn~\cite{Brunn}, which is a laboratory information
+system for microplate based high-throughput screening. In this manner open
+source projects empower other open source projects.
 
@@
 and analysing computational chemistry calculations. An interesting
-aspect of both of these projects is that they share some developers
-with the underlying toolkit and this has driven the development of new
+aspect of both Avogadro and Bioclipse is that they share some developers
+with the underlying toolkits and this has driven the development of new
 features in the CDK and Open Babel.

### 2011-04-20 paper.tex
minor proof reading fixes

@@
 Based on the outstanding work of Michael (Miguel) Howard and others, this development
-path took Jmol from a "Rasmol/Chime replacement" to a fully fledged molecular
+path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
 visualization package, including full support for crystallography~\cite{Hanson2010},
@@
 It also provides some less common functionality, like matching
-tautomers ans resonance substructures, enumeration of subgraps,
+tautomers and resonance substructures, enumeration of subgraps,
 finding maximum common substructure of $N$ input structures, and

### 2011-04-20 paper.tex

@@
 on the CDK and Open Babel, respectively. Bioclipse is an award-winning
-`molecular workbench' for life sciences with a dynamic plugin architecture, and
+molecular workbench for life sciences with a dynamic plugin architecture, and
 which wraps cheminformatics functionality behind user-friendly interfaces and
@@
 using an integrated high-level language called Bioclipse Scripting Language
-(BSL), which can be used for invoking functionality or to automate complex
+(BSL), which can be used for scripted research or to automate complex
 tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
-such as for example Brunn~\cite{Brunn}, which is a laboratory information
-system for microplate based high-throughput screening. In this manner open
-source projects empower other open source projects.
+such as Brunn~\cite{Brunn}, which is a laboratory information system for
+microplate based high-throughput screening. In this manner open source projects
+empower other open source projects.
 

### 2011-04-20 paper.tex
spelling

@@
 
-In the area of cheminformatics tookits, two of the existing toolkits
+In the area of cheminformatics toolkits, two of the existing toolkits
 Open Babel and RDKit are planning to work together on a common

### 2011-04-20 paper.tex
Added section on IUPAC nomenclature/OPSIN

@@
          \email{Peter Murray-Rust - pm286@cam.ac.uk}
+     \and 
+         Daniel M Lowe$^7$%
+         \email{Daniel Lowe - dl387@cam.ac.uk}
      \and
@@
 
+    \subsubsection*{IUPAC Nomenclature Conversion}
+The oldest form of communication used to describe chemicals, predating even the knowledge of the atomic structure of compounds, is the chemical name.
+Even now chemical names are abundant in the scientific literature and encode valuable structural information.
+The IUPAC have through successive books of recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993} aimed to codify and to an extent standardise naming practices.
+The latest provisional recommendations are freely available through the IUPAC's website~\cite{iupac_draft_2004}.
+
+The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC Nomenclature)~\cite{lowe_chemical_2011} aims to make this abundance of chemical names machine readable by translating them to SMILES, CML or InChI.
+The program is based around the use of a regular grammar to guide tokenisation and parsing of chemical names, followed by step-wise application of nomenclature rules.
+OPSIN is able to offer fast and precise conversions for the majority of names using IUPAC organic nomenclature.
+OPSIN is available as a web service, Java library and standalone application for maximum interoperability.
+
 \subsubsection*{QSAR-ML}
@@
         OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
+        OPSIN & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
         A3 & ..  & .  \\ \hline

### 2011-04-22 paper.tex
Improved section on web based tools and ChemDoodle Web Components. Added ChemDoodle Web Components to projects table. I don't know where to place references, but the urls are as follows:
ChemDoodle Web Components - http://web.chemdoodle.com
HTML5 - http://dev.w3.org/html5/spec/

@@
 
-One important aspect of the above
-toolkits is that they originated for the support of desktop
-and server-side applications. More recently, a number of workers have developed light-weight
-toolkits that are targeted towards web applications. One example of
-this form is the ChemDoodle library from iChemLabs. The focus of these
-toolkits is to support client-side cheminformatics functionality,
-allowing for rich, chemically-aware browser-based applications.
+While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented.
+As such, there is a growing need to present chemical and scientific information in a web-based medium.
+Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in Javascript.
+This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education.
+It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial.
+As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilize these new technologies to improve products and education.
+A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs.
+The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception.
+ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra.
+Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices.
+It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets.
+iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry.
+This goal helps to make sure that the cost of education decreases while using the web to further spread science.
 
@@
         RDKit & http://rdkit.org & Greg Landrum \\ \hline
+        ChemDoodle Web Components & http://web.chemdoodle.com & iChemLabs \\ \hline
         \multicolumn{3}{|c|}{Integration} \\ \hline

### 2011-04-24 paper.tex
Improved section on business. Other parties still need to finish their paragraphs. References:
iBabel - http://homepage.mac.com/swain/Sites/Macinchem/page65/ibabel3.html
ChemSpotlight - http://chemspotlight.openmolecules.net/
ChemSpider/ChemDoodle - http://www.chemspider.com/blog/ichemlabs-and-rsc-chemspider-announce-partnership.html

@@
 
-Although the software developed by the Blue Obelisk is Open Source on
-principle, this does not rule out the establishment of a business
-based on open source chemistry software. Rather than supporting itself
-through sales, such a business generally relies on providing support,
-customisation and services around the software. The interaction
-between an open source project and a company is important, and works
-best where there is some form of `giving back' to the project (such as
-developer time, code contributions, or financial assistance with server costs).
+Open Source provides a unique opportunity for commercial organizations to work with the community.
+As opposed to traditional business models which focus on maximizing revenue through monetization of all aspects of software, working with the community allows a business to maximize profits through greater community adoption and popularity.
+This is an ideal solution for some commercial organizations because only they can afford the significant investment in funding and time that is required to develop a successful open source project.
+Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry.
+A goal of the Blue Obelisk is to reduce these challenges and bring commercial organizations and the community together to benefit science.
+
+Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license.
+With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources.
+Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development.
+Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it.
+These benefits allow a company to create a successful product with significantly reduced cost and effort.
+
+Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded.
+In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass.
+This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities.
+On the other hand, commercial organizations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption.
+In return, that commercial organization gains recognition from that product and is the first choice for commercial support, consulting and custom development.
+This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.
+
+In the scientific industry, it is incredibly difficult to start successful open projects.
+This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment.
+Because of this, there is an increased barrier to starting open projects in the scientific industry.
+The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects.
+Commercial organizations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.
+
+For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing.
+The license is chosen to best ensure the success of the project.
+iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception.
+iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilize the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content.
+Other open source developers can benefit from this library freely and expect continued development, while commercial organizations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market.
+Of course, no one will need to waste time rewriting functionality that is already provided with quality.
+Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
 

### 2011-04-26 paper.tex
Added a screenshot for the BOx website

@@
 
+\subsection*{Getting the message out}
+
+One recent development
+that attempts to address this issue was the establishment in Apr 2010 of a
+question-and-answer (QA)
+website related to Blue Obelisk projects and themes at
+http://blueobelisk.shapado.com (see Figure~3). This is a website in the
+style of Stack Overflow that encourages high quality answers (and
+questions) through the use of a voting system. In the year since it
+was established, over
+200 users have registered, many of whom had no previous involvement
+with the Blue Obelisk, showing that the QA website is a nice complement
+to earlier existing channels of communication.
+
 Books (Angel, Egon, OB)
@@
 BO mailing list and blogs of members.
-
-One recent development
-that attempts to address this issue was the establishment in Apr 2010 of a
-question-and-answer
-website related to Blue Obelisk projects and themes at
-http://blueobelisk.shapado.com. This is a website in the
-style of Stack Overflow that encourages high quality answers (and
-questions) through the use of a voting system. In the year since it
-was established, over
-200 users have registered, many of whom had no previous involvement
-with the Blue Obelisk.
 
@@
       Figure legend text.
+
+  \subsection*{Figure 3 - Screenshot of the Blue Obelisk eXchange Question
+    and Answer website.}
 

### 2011-04-26 paper.tex
Added text on Open Data

@@
 be licensed explicitly, and preferably under CC0 (Creative Commons `No
-Rights Reserved') \cite{WebCC0}. This license allows others to use the data for any
-purpose whatsoever without any barriers.
+Rights Reserved', also known as CCZero) \cite{WebCC0}. This license allows others to use the 
+data for any purpose whatsoever without any barriers. Other licenses
+compatible with the Panton Principles include the
+Open Data Commons Public Domain Dedication and Licence (PDDL), the
+Open Data Commons Attribution License, and the
+Open Data Commons Open Database License (ODbL) (see
+http://www.opendefinition.org/licenses/\#Data).
+
+Despite this positive news, little chemical data has become
+available from the traditional chemical fields of organic,
+inorganic, solid state chemistry. Table~2 lists a few notable
+exceptions: ChemPedia (a now discontinued crowd-sourcing project),
+CrystalEye (http://wwmm.ch.cam.ac.uk/crystaleye/),
+and the Open Notebook Science Solubility
+data~\cite{ONS2010}. There is also data available using licenses
+not compatible with the Panton Principles, but where the user
+is allows to modify and redistribute the data. A new data
+set in this category is the data from the ChEMBL database, 
+which is available under the Creative Commons Share-Alike
+Attribution license~\cite{Overington2009}.
+
+Importantly, publishing data as CC0 is become easier now that
+website are becoming available to simplify publishing data. Two
+projects that can be mentioned in this context are FigShare
+(http://figshare.com/), where the data behind unpublished figures
+can be hosted, and Dryad (http://datadryad.org/) where data
+behind publications can be hosted. Initiatives like this make
+it possible to host small amounts of data, and those combinded
+are expected to become soon a substantial knowledge base.
 
@@
       }
-  \subsection*{Table 2 - Sample table title}
-    Large tables are attached as separate files but should
-    still be described here.
+  \subsection*{Table 2 - Open Data in chemistry.}
+    Overview of major open chemical data available under a license or waiver
+    compatible with the Panton Principles.
+    \par \mbox{}
+    \par
+    \mbox{
+\begin{tabular}{|l|l|l|}
+  %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
+  \hline Name & License/Waiver & Description \\ \hline
+  ChemPedia & CC0 & Crowd-sourced chemical names. Project discontinued. \\ \hline
+  CrystalEye     & PPDL & Crystal structures from primary literature. \\ \hline
+  ONS Solubility & CC0 & Solubility data for various solvents. \\ \hline
+\end{tabular}
+      }
 

### 2011-04-26 paper.tex
Correct spelling.

@@
 have opened the door for 
-an easy and seamless integration of FOS components into executable workflows 
+an easy and seamless integration of BO components into executable workflows 
 or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
@@
 path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
-visualization package, including full support for crystallography~\cite{Hanson2010},
+visualisation package, including full support for crystallography~\cite{Hanson2010},
 efficient delivery of surface data using fast solvent-excluded surface
@@
  display of molecular orbitals from standard basis set/coefficient data,
-the inclusion of dynamic minimization using the UFF force field, and
+the inclusion of dynamic minimisation using the UFF force field, and
 a full implementation of Daylight SMILES and SMART, with extensions to
@@
 It also provides some less common functionality, like matching
-tautomers and resonance substructures, enumeration of subgraps,
+tautomers and resonance substructures, enumeration of subgraphs,
 finding maximum common substructure of $N$ input structures, and
@@
 As such, there is a growing need to present chemical and scientific information in a web-based medium.
-Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in Javascript.
+Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in JavaScript.
 This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education.
 It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial.
-As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilize these new technologies to improve products and education.
+As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilise these new technologies to improve products and education.
 A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs.
@@
 CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
-Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharma- cophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
+Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
 CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
@@
 
-Open Source provides a unique opportunity for commercial organizations to work with the community.
-As opposed to traditional business models which focus on maximizing revenue through monetization of all aspects of software, working with the community allows a business to maximize profits through greater community adoption and popularity.
-This is an ideal solution for some commercial organizations because only they can afford the significant investment in funding and time that is required to develop a successful open source project.
+Open Source provides a unique opportunity for commercial organisations to work with the community.
+As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity.
+This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project.
 Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry.
-A goal of the Blue Obelisk is to reduce these challenges and bring commercial organizations and the community together to benefit science.
+A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.
 
@@
 This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities.
-On the other hand, commercial organizations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption.
-In return, that commercial organization gains recognition from that product and is the first choice for commercial support, consulting and custom development.
+On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption.
+In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development.
 This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.
@@
 The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects.
-Commercial organizations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.
+Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.
 
@@
 iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception.
-iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilize the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content.
-Other open source developers can benefit from this library freely and expect continued development, while commercial organizations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market.
+iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content.
+Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market.
 Of course, no one will need to waste time rewriting functionality that is already provided with quality.
@@
 Performance. There is a famous quote by Knuth that "Premature
-optimization is root of all evil (or something)". Given that compute
+optimisation is root of all evil (or something)". Given that compute
 time is cheap, ....
@@
 SMILES notation, which allows to define crystals structures, polymers,
-electron delocalization charges, molecule interations, and many other
+electron delocalisation charges, molecule interactions, and many other
 features absent in the initial SMILES specification. There has been
@@
 To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations.
-Descriptor implementations were initially available from CDK and JoeLIB, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
+Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
 
@@
 registration systems in general is the notion of structure
-standardization.  That is,  for a given input structure, multiple
+standardisation.  That is,  for a given input structure, multiple
 representations should be converted to one canonical form. 
-Structure canonicalization routines partially address this aspect,
+Structure canonicalisation routines partially address this aspect,
 converting multiple alternative topologies to a single canonical
-form. However, the problem of standardization is broader than just
-topological canonicalization. Features that must be considered include
+form. However, the problem of standardisation is broader than just
+topological canonicalisation. Features that must be considered include
 \begin{itemize}
-\item topological canonicalization
+\item topological canonicalisation
 \item handling of charges
-\item tautomer enumeration and canonicalization
-\item normalization of functional groups
+\item tautomer enumeration and canonicalisation
+\item normalisation of functional groups
 \end{itemize}
-Currently, most of the individual components of a `standardization
+Currently, most of the individual components of a `standardisation
 pipeline' can be implemented using BO tools. The larger problem is
-that there is no agreed upon list of steps for a standardization
-process. While some specifications have been published (e.g., Pubchem)
-and some standardization services and tools are available (Pubchem
+that there is no agreed upon list of steps for a standardisation
+process. While some specifications have been published (e.g., PubChem)
+and some standardisation services and tools are available (PubChem
 provides an online service to standardise molecules and the NCGC
 provides a stand alone tool) each group has their own set of rules. A
-common reference specification for standardization would be of immense
+common reference specification for standardisation would be of immense
 value in interoperability between structure repositories as well as
@@
 behind publications can be hosted. Initiatives like this make
-it possible to host small amounts of data, and those combinded
+it possible to host small amounts of data, and those combined
 are expected to become soon a substantial knowledge base.
@@
 \section*{Other areas of activity}
-Web services? XMPP services? (NIH, CDK and Chemspider)
+Web services? XMPP services? (NIH, CDK and ChemSpider)
 
@@
 that attempts to address this issue was the establishment in Apr 2010 of a
-question-and-answer (QA)
+question-and-answer (Q\&A)
 website related to Blue Obelisk projects and themes at
@@
 200 users have registered, many of whom had no previous involvement
-with the Blue Obelisk, showing that the QA website is a nice complement
+with the Blue Obelisk, showing that the Q\&A website is a nice complement
 to earlier existing channels of communication.

### 2011-04-26 paper.tex
Move OSRA and OPSIN into own section

@@
       \and
+         Dmitry Pavlov$^8$%
+         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
+      \and
+         Igor V Filippov\correspondingauthor$^{10}$%
+         \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
+      \and
          Jane E Doe\correspondingauthor$^2$%
          \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
-      \and
-         Dmitry Pavlov$^8$%
-         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
       }
@@
     \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
-    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden 
+    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
+    \iid(10)NIH, Somewhere in the US, USA
 }%
@@
 
+\subsubsection*{Converting chemical names and images to structures}
+
+The majority of chemical information is not stored in machine-readable
+formats, but rather as chemical names or depictions. The OSRA and OPSIN
+projects focus on extracting chemical information from these sources.
+Such software plays a particularly important role for datamining the
+chemical literature, including patents and theses.
+
+Optical Structure Recognition Application (OSRA) \cite{WebOSRA} was started
+in early 2007 with the goal to create first free and open source
+tool for extraction and conversion of molecular images into SMILES and
+SD files. From the very beginning the underlying philosophy was to integrate
+existing open source libraries and to avoid ``reinventing the wheel''
+wherever possible. OSRA relies on a variety of open source components:
+Open Babel for chemical format
+conversion and molecular property calculations, GraphicsMagick for image
+manipulation, Potrace for vectorization, GOCR and OCRAD for optical
+character recognition. The growing importance of image
+recognition technology can be seen in the fact that 
+only a few years ago there was only one widely available software
+package for chemical structure recognition -  CLiDE (commercially
+developed at Keymodule, Ltd), but today there are as many as seven
+available programs.
+
+The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC
+Nomenclature)~\cite{lowe_chemical_2011} focuses instead on interpreting chemical names.
+The chemical name is the oldest form of communication used to
+describe chemicals, predating
+even the knowledge of the atomic structure of compounds.
+Chemical names are abundant in the scientific
+literature and encode valuable structural information.
+Through successive books of
+recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993},
+IUPAC have tried to codify and to an extent standardise naming practices.
+OPSIN aims to make this abundance of
+chemical names machine readable by translating them to SMILES, CML or
+InChI. The program is based around the use of a regular grammar to
+guide tokenisation and parsing of chemical names, followed by
+step-wise application of nomenclature rules. OPSIN is able to offer
+fast and precise conversions for the majority of names using IUPAC
+organic nomenclature, and is available as a web service, Java
+library and standalone application for maximum interoperability.
+
   \subsubsection*{Collaboration and interoperability}
@@
 ?
-
-    \subsubsection*{IUPAC Nomenclature Conversion}
-The oldest form of communication used to describe chemicals, predating even the knowledge of the atomic structure of compounds, is the chemical name.
-Even now chemical names are abundant in the scientific literature and encode valuable structural information.
-The IUPAC have through successive books of recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993} aimed to codify and to an extent standardise naming practices.
-The latest provisional recommendations are freely available through the IUPAC's website~\cite{iupac_draft_2004}.
-
-The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC Nomenclature)~\cite{lowe_chemical_2011} aims to make this abundance of chemical names machine readable by translating them to SMILES, CML or InChI.
-The program is based around the use of a regular grammar to guide tokenisation and parsing of chemical names, followed by step-wise application of nomenclature rules.
-OPSIN is able to offer fast and precise conversions for the majority of names using IUPAC organic nomenclature.
-OPSIN is available as a web service, Java library and standalone application for maximum interoperability.
 

### 2011-04-26 paper.tex
Removed text on workflow systems from Background and moved to RemovedText.tex.

@@
 by biologists \cite{DeMatos:2009p3839}.
-Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
-have opened the door for 
-an easy and seamless integration of BO components into executable workflows 
-or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
-reported in 2010 \cite{Kuhn:2010p4001}.
-
-KNIME currently has a built-in collection of CDK-based and
-OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
-chemistry workflows. Another collection of chemistry nodes---based on
-the RDKit and its Java wrapper---appeared in 2010 and is available to
-KNIME users from the KNIME's official ``Community Updates'' site.
-Indigo nodes for KNIME are in preparation and will appear nearby the
-RDKit nodes by May 2011.
 
@@
 \begin{itemize}
-\item Open Source. One can use other people's code without further
+\item Open Source in Chemistry. One can use other people's code without further
 permission, including changing it for one's own use and distributing
 it again.
-\item Open Standards. One can find visible community mechanisms for
+\item Open Standards in Chemistry. One can find visible community mechanisms for
 protocols and communicating information. The mechanisms for creating
@@
 organisations, including various degrees of consent.
-\item Open Data. One can obtain all data in the public domain when
+\item Open Data in Chemistry. One can obtain all data in the public domain when
 wanted and reuse it for whatever purpose.

### 2011-04-29 paper.tex
Initial comments.

--- 2011-04-26 paper.tex
@@
      \and 
-	     Christoph Steinbeck$^6$%
-	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
-     \and 
          Noel M O'Boyle$^5$%
@@
       \and
-         Ola Spjuth$^9$%
-         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
-     \and 
-         Jonathan Alvarsson$^9$%
-         \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
-     \and 
-         Peter Murray-Rust$^7$%
-         \email{Peter Murray-Rust - pm286@cam.ac.uk}
-     \and 
-         Daniel M Lowe$^7$%
-         \email{Daniel Lowe - dl387@cam.ac.uk}
-     \and
-         Robert Hanson$^8$%
-         \email{Robert Hanson - hansonr@stolaf.edu}
-      \and
          Jane E Doe\correspondingauthor$^2$%
          \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
-      \and
-         Dmitry Pavlov$^8$%
-         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
       }
@@
     \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
-    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
-    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
-    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
-    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
-    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden 
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
+    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland
 }%
@@
         % Do not use inserted blank lines (ie \\) until main body of text.
-        \paragraph*{Background:} The Blue Obelisk movement was established in 2005 as a 
-		response to the lack of open data, open standards and
-		open source (ODOSOS) in chemistry. While other scientific disciplines
-		such as physics, biology and astronomy (to name a few) were embracing
-		new ways of doing science and reaping the benefits of community
-		efforts, there was little if any innovation in the field of chemistry
-		and scientific progress was actively hampered by the lack of access to
-		data and tools.
+        \paragraph*{Background:} Text for this section of the abstract.
 
-        \paragraph*{Results:} This
-		contribution looks back on the past 5 years and surveys progress and
-		remaining challenges in the areas of Open Source, Open Data and Open
-		Standards in chemistry.
+        \paragraph*{Results:} Text for this section of the abstract \ldots
 
-        \paragraph*{Conclusions:} Here we show that the Blue Obelisk has been very successful
-		in bringing together researchers and developers with common interests
-		in ODOSOS, leading to development of many useful resources freely
-		available to the chemistry community. But how best to engage with the
-		wider chemistry community outside of the Blue Obelisk remains an open
-		question.
+        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
 \end{abstract}
@@
 \section*{Background}
-The Blue Obelisk movement was established in 2005 at the
+The Blue Obelisk group was established in 2005 at the
 229\textsuperscript{th} National Meeting of the American Chemistry
@@
 data and tools.
-Since 2005 it has become evident that a good amount of development in open 
-chemical information and information is driven by the demands of neighbouring
-scientific fields. In many areas in biology, for example, the importance of 
-small molecules and their interactions and reactions in biological systems
-has been realised. In fact, one of the first free and open databases and ontologies 
-of small molecules was created as a resource about chemical structure and nomenclature
-by biologists \cite{DeMatos:2009p3839}.
-Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
-have opened the door for 
-an easy and seamless integration of BO components into executable workflows 
-or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
-reported in 2010 \cite{Kuhn:2010p4001}.
-
-KNIME currently has a built-in collection of CDK-based and
-OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
-chemistry workflows. Another collection of chemistry nodes---based on
-the RDKit and its Java wrapper---appeared in 2010 and is available to
-KNIME users from the KNIME's official ``Community Updates'' site.
-Indigo nodes for KNIME are in preparation and will appear nearby the
-RDKit nodes by May 2011.
 
@@
 and maintaining these standards cover a wide spectrum of human
-organisations, including various degrees of consent.
+organizations, including various degrees of consent.
 \item Open Data. One can obtain all data in the public domain when
@@
 
-The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
-later as a formal paper by Guha et al.\cite{Guha2006} in
+Note that while some may advocate also for Open Access to
+publications, the goals focus more on the availability of code (to
+reproduce results), standards (to exchange data), and the scientific
+data itself. All three of these goals stem from the basics of the
+scientific method for data sharing and reproducibilty.
+
+The Blue Obelisk was first described in Guha et al.\cite{Guha2006} in
 2006. This
@@
 
-Some project have also been adopted by other fields, including biology
-and crystallography. A good example
-here is Jmol, which has, in the past five years, gone through two major revisions
-as part of an accelerated development focus. Starting with Jmol 10.2, released in
-April of 2006, Jmol 11.0 added over 100 new features and was released in March of
-2007. Three years later, Jmol 12.0 followed with over 700 new features.
-Based on the outstanding work of Michael (Miguel) Howard and others, this development
-path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
-visualisation package, including full support for crystallography~\cite{Hanson2010},
-efficient delivery of surface data using fast solvent-excluded surface
-generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
- display of molecular orbitals from standard basis set/coefficient data,
-the inclusion of dynamic minimisation using the UFF force field, and
-a full implementation of Daylight SMILES and SMART, with extensions to
-conformational and biomolecular substructure searching in the form of Jmol
-BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
-Behind the scenes, Jmol has been essentially completely rewritten. A full
-mathematical expression evaluator was added, and the scripting language
-was folded into a versatile JavaScript-like command language, including a
-complete set of program flow commands, several new data types, and
-user-defined variables and functions. An exciting non-chemical application
-of Jmol has been its incorporation into the open source Sage
-Mathematics Project [http://www.sagemath.org/], where Jmol is being used
-to deliver three-dimensional depiction of mathematical functions.
-
 Two new Open Source cheminformatics toolkits have appeared since the
@@
 BSD License. RDKit continues to be actively developed and includes
-code donated by Novartis.
+code donated by Novartis. More recently, GGA Software Services (a
+contract programming company) released the Indigo toolkit
+\cite{WebIndigo} and associated software in 2009 under the GPL. [More
+text on Indigo here]
 
-More recently, GGA Software Services (a contract programming company)
-released the Indigo toolkit \cite{WebIndigo} and associated software
-in 2009 under the GPL. Indigo is written from scratch in C++ and has
-high-level wrappers for using it in C, Java, Python, or .NET
-environment. Like other toolkits, Indigo provides support for
-tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
-exact/substructure/SMARTS matching, fingerprint generation (all these
-both for molecules and reactions), and canonical SMILES computation.
-It also provides some less common functionality, like matching
-tautomers and resonance substructures, enumeration of subgraphs,
-finding maximum common substructure of $N$ input structures, and
-enumerating reaction products. Indigo also has its rendering plugin,
-which is used in a lot of commercial and open-source projects as the
-main 2D chemical rendering engine.
-
-While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented.
-As such, there is a growing need to present chemical and scientific information in a web-based medium.
-Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in JavaScript.
-This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education.
-It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial.
-As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilise these new technologies to improve products and education.
-A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs.
-The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception.
-ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra.
-Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices.
-It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets.
-iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry.
-This goal helps to make sure that the cost of education decreases while using the web to further spread science.
+One important aspect of the above
+toolkits is that they originated for the support of desktop
+and server-side applications. More recently, a number of workers have developed light-weight
+toolkits that are targeted towards web applications. One example of
+this form is the ChemDoodle library from iChemLabs. The focus of these
+toolkits is to support client-side cheminformatics functionality,
+allowing for rich, chemically-aware browser-based applications.
 
@@
 
-Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
-\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
-on the CDK and Open Babel, respectively. Bioclipse is an award-winning
-molecular workbench for life sciences with a dynamic plugin architecture, and
-which wraps cheminformatics functionality behind user-friendly interfaces and
-graphical editors. All functionality in Bioclipse is completely scriptable
-using an integrated high-level language called Bioclipse Scripting Language
-(BSL), which can be used for scripted research or to automate complex
-tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
-such as Brunn~\cite{Brunn}, which is a laboratory information system for
-microplate based high-throughput screening. In this manner open source projects
-empower other open source projects.
-
-Avogadro is a 3D molecular editor and viewer aimed at preparing
+Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
+\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such
+software, based on the CDK and Open Babel, respectively. Bioclipse is
+an award-winning `molecular workbench' for life sciences that ...,
+while Avogadro is a 3D molecular editor and viewer aimed at preparing
 and analysing computational chemistry calculations. An interesting
-aspect of both Avogadro and Bioclipse is that they share some developers
-with the underlying toolkits and this has driven the development of new
+aspect of both of these projects is that they share some developers
+with the underlying toolkit and this has driven the development of new
 features in the CDK and Open Babel.
-
-CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
-Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
-CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
-Taverna version 2.x.
-The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.
 
@@
 
-Open Source provides a unique opportunity for commercial organisations to work with the community.
-As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity.
-This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project.
-Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry.
-A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.
-
-Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license.
-With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources.
-Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development.
-Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it.
-These benefits allow a company to create a successful product with significantly reduced cost and effort.
-
-Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded.
-In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass.
-This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities.
-On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption.
-In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development.
-This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.
-
-In the scientific industry, it is incredibly difficult to start successful open projects.
-This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment.
-Because of this, there is an increased barrier to starting open projects in the scientific industry.
-The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects.
-Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.
-
-For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing.
-The license is chosen to best ensure the success of the project.
-iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception.
-iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content.
-Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market.
-Of course, no one will need to waste time rewriting functionality that is already provided with quality.
-Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
+Although the software developed by the Blue Obelisk is Open Source on
+principle, this does not rule out the establishment of a business
+based on open source chemistry software. Rather than supporting itself
+through sales, such a business generally relies on providing support,
+customisation and services around the software. The interaction
+between an open source project and a company is important, and works
+best where there is some form of `giving back' to the project (such as
+developer time, code contributions, or financial assistance with server costs).
 
@@
 uses Open Babel under-the-hood to manage its compound collections.
+
 Another more recent example occurred in July 2010 when Silicos, a
@@
 searching that provides extensive support for a variety of
-pharmacophore searches. Other tools released by them include Sieve for
-filtering by molecular property and XXX.
+pharmacophore searches. Other tools released by Silicos include Sieve for
+filtering by molecular property, Stripper for removing core scaffold
+structures from a molecule set, and Piramid for molecular alignment
+using shape determined by the Gaussian volumes as a descriptor.
 
@@
 support 2D to 3D conversion of structures (through energy
-minimisation). Similarly, efficient Jmol code for atom-atom rebonding
-has been ported to the CDK.
+minimisation).
 
@@
 
-In the area of cheminformatics toolkits, two of the existing toolkits
+In the area of cheminformatics tookits, two of the existing toolkits
 Open Babel and RDKit are planning to work together on a common
-underlying framework called MolCore. This project is still in the
-planning stage, but if it is a success it will mean that the two
+underlying framework called MolCore.\cite{WebMolCore} This project is still in the
+planning stage, but if it is a success it will mean that the the two
 libraries will be interoperable (while retaining their existing focus)
@@
 Performance. There is a famous quote by Knuth that "Premature
-optimisation is root of all evil (or something)". Given that compute
+optimization is root of all evil (or something)". Given that compute
 time is cheap, ....
 
-\section*{Chemical Structure Registration Systems, Databases}
-Databases? (XXX CS: A couple of database-related paragraphs could go here XXX)
-
-XXX the paragraph below on OrChem needs some work
-Registration, indexing and searching of chemical structures in relational databases is one of the core areas of cheminformatics. 
-A number of structure registration systems have been published in the last five years, exploiting the fact that 
-free cheminformatics toolkits such as OpenBabel and the CDK were available. 
-OrChem, for example, is an extension for the Oracle 11G database that adds registration and indexing of chemical structures to support fast substructure and similarity searching.
-The cheminformatics functionality is provided by the Chemistry Development Kit.
-OrChem provides similarity searching with response times in the order of seconds for databases with millions of compounds, depending on a given similarity cut-off.
-For substructure searching, it can make use of multiple processor cores on today's powerful database servers to provide fast response times in equally large data sets.
-OrChem is free software and can be redistributed and/or modified under the terms of the GNU Lesser General Public License as published by the Free Software Foundation.
-All software is available via http://orchem.sourceforge.net.
+Databases?
 
@@
 this standard identifier by the chemistry community.
-
-Also, Indigo is going to provide its own plugin with an interface to
-the InChI implementation in May 2011. This will enable developers to
-calculate InChI and InChIKey-s within Python and .NET projects (not to
-mention Java), eliminating the trouble of compiling and using binary
-modules directly.
 
@@
 
-Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
-SMILES notation, which allows to define crystals structures, polymers,
-electron delocalisation charges, molecule interactions, and many other
-features absent in the initial SMILES specification. There has been
-discussions about including parts of the CurlySMILES notation into
-OpenSMILES, especially polymers.
-
-\subsubsection*{CML}
-?
-
-    \subsubsection*{IUPAC Nomenclature Conversion}
-The oldest form of communication used to describe chemicals, predating even the knowledge of the atomic structure of compounds, is the chemical name.
-Even now chemical names are abundant in the scientific literature and encode valuable structural information.
-The IUPAC have through successive books of recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993} aimed to codify and to an extent standardise naming practices.
-The latest provisional recommendations are freely available through the IUPAC's website~\cite{iupac_draft_2004}.
-
-The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC Nomenclature)~\cite{lowe_chemical_2011} aims to make this abundance of chemical names machine readable by translating them to SMILES, CML or InChI.
-The program is based around the use of a regular grammar to guide tokenisation and parsing of chemical names, followed by step-wise application of nomenclature rules.
-OPSIN is able to offer fast and precise conversions for the majority of names using IUPAC organic nomenclature.
-OPSIN is available as a web service, Java library and standalone application for maximum interoperability.
-
-\subsubsection*{QSAR-ML}
-The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses.
-QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}.
-A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation.
-Hence, a dataset described in QSAR-ML is completely reproducible.
-To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations.
-Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
-
-\subsubsection*{Others...?}
+CML?
 
@@
 registration systems in general is the notion of structure
-standardisation.  That is,  for a given input structure, multiple
+standardization.  That is,  for a given input structure, multiple
 representations should be converted to one canonical form. 
-Structure canonicalisation routines partially address this aspect,
+Structure canonicalization routines partially address this aspect,
 converting multiple alternative topologies to a single canonical
-form. However, the problem of standardisation is broader than just
-topological canonicalisation. Features that must be considered include
+form. However, the problem of standardization is broader than just
+topological canonicalization. Features that must be considered include
 \begin{itemize}
-\item topological canonicalisation
+\item topological canonicalization
 \item handling of charges
-\item tautomer enumeration and canonicalisation
-\item normalisation of functional groups
+\item tautomer enumeration and canonicalization
+\item normalization of functional groups
 \end{itemize}
-Currently, most of the individual components of a `standardisation
+Currently, most of the individual components of a `standardization
 pipeline' can be implemented using BO tools. The larger problem is
-that there is no agreed upon list of steps for a standardisation
-process. While some specifications have been published (e.g., PubChem)
-and some standardisation services and tools are available (PubChem
+that there is no agreed upon list of steps for a standardization
+process. While some specifications have been published (e.g., Pubchem)
+and some standardization services and tools are available (Pubchem
 provides an online service to standardise molecules and the NCGC
 provides a stand alone tool) each group has their own set of rules. A
-common reference specification for standardisation would be of immense
+common reference specification for standardization would be of immense
 value in interoperability between structure repositories as well as
@@
 be licensed explicitly, and preferably under CC0 (Creative Commons `No
-Rights Reserved', also known as CCZero) \cite{WebCC0}. This license allows others to use the 
-data for any purpose whatsoever without any barriers. Other licenses
-compatible with the Panton Principles include the
-Open Data Commons Public Domain Dedication and Licence (PDDL), the
-Open Data Commons Attribution License, and the
-Open Data Commons Open Database License (ODbL) (see
-http://www.opendefinition.org/licenses/\#Data).
-
-Despite this positive news, little chemical data has become
-available from the traditional chemical fields of organic,
-inorganic, solid state chemistry. Table~2 lists a few notable
-exceptions: ChemPedia (a now discontinued crowd-sourcing project),
-CrystalEye (http://wwmm.ch.cam.ac.uk/crystaleye/),
-and the Open Notebook Science Solubility
-data~\cite{ONS2010}. There is also data available using licenses
-not compatible with the Panton Principles, but where the user
-is allows to modify and redistribute the data. A new data
-set in this category is the data from the ChEMBL database, 
-which is available under the Creative Commons Share-Alike
-Attribution license~\cite{Overington2009}.
-
-Importantly, publishing data as CC0 is become easier now that
-website are becoming available to simplify publishing data. Two
-projects that can be mentioned in this context are FigShare
-(http://figshare.com/), where the data behind unpublished figures
-can be hosted, and Dryad (http://datadryad.org/) where data
-behind publications can be hosted. Initiatives like this make
-it possible to host small amounts of data, and those combined
-are expected to become soon a substantial knowledge base.
+Rights Reserved') \cite{WebCC0}. This license allows others to use the data for any
+purpose whatsoever without any barriers.
 
@@
 \section*{Other areas of activity}
-Web services? XMPP services? (NIH, CDK and ChemSpider)
-
-\subsection*{Getting the message out}
-
-One recent development
-that attempts to address this issue was the establishment in Apr 2010 of a
-question-and-answer (Q\&A)
-website related to Blue Obelisk projects and themes at
-http://blueobelisk.shapado.com (see Figure~3). This is a website in the
-style of Stack Overflow that encourages high quality answers (and
-questions) through the use of a voting system. In the year since it
-was established, over
-200 users have registered, many of whom had no previous involvement
-with the Blue Obelisk, showing that the Q\&A website is a nice complement
-to earlier existing channels of communication.
+Web services? (NIH, CDK and Chemspider)
 
@@
 
+One recent development
+that attempts to address this issue was the establishment in Apr 2010 of a
+question-and-answer
+website related to Blue Obelisk projects and themes at
+http://blueobelisk.shapado.com. This is a website in the
+style of Stack Overflow that encourages high quality answers (and
+questions) through the use of a voting system. In the year since it
+was established, over
+200 users have registered, many of whom had no previous involvement
+with the Blue Obelisk.
+
 We hope to see this involvement between the Blue Obelisk and the wider
-community grow in the future. To this end, we encourage the reader to
+community grow in future. To this end, we encourage the reader to
 visit http://blueobelisk.org, send a message to our mailing list,
@@
 
-  \subsection*{Figure 3 - Screenshot of the Blue Obelisk eXchange Question
-    and Answer website.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
         \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
-        Chemistry Development Kit (CDK) \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
+        Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
         Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
@@
         RDKit & http://rdkit.org & Greg Landrum \\ \hline
-        ChemDoodle Web Components & http://web.chemdoodle.com & iChemLabs \\ \hline
-        \multicolumn{3}{|c|}{Integration} \\ \hline
-        CDK-Taverna \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
-        A3 & ..  & .  \\ \hline
         \multicolumn{3}{|c|}{Interconversion} \\ \hline
         OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
-        OPSIN & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
         A3 & ..  & .  \\ \hline
-        \multicolumn{3}{|c|}{Structure Databases} \\ \hline
-        OrChem \cite{RijnbeekS10} & http://orchem.sourceforge.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
-        MyChem  & http://XXX.sourceforge.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
-        PGChem  & http://XXX.sourceforge.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
-        Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
-
       \end{tabular}
       }
-  \subsection*{Table 2 - Open Data in chemistry.}
-    Overview of major open chemical data available under a license or waiver
-    compatible with the Panton Principles.
-    \par \mbox{}
-    \par
-    \mbox{
-\begin{tabular}{|l|l|l|}
-  %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-  \hline Name & License/Waiver & Description \\ \hline
-  ChemPedia & CC0 & Crowd-sourced chemical names. Project discontinued. \\ \hline
-  CrystalEye     & PPDL & Crystal structures from primary literature. \\ \hline
-  ONS Solubility & CC0 & Solubility data for various solvents. \\ \hline
-\end{tabular}
-      }
+  \subsection*{Table 2 - Sample table title}
+    Large tables are attached as separate files but should
+    still be described here.
 

### 2011-04-29 RemovedText.tex

Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
have opened the door for 
an easy and seamless integration of BO components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

KNIME currently has a built-in collection of CDK-based and
OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
chemistry workflows. Another collection of chemistry nodes---based on
the RDKit and its Java wrapper---appeared in 2010 and is available to
KNIME users from the KNIME's official ``Community Updates'' site.
Indigo nodes for KNIME are in preparation and will appear nearby the
RDKit nodes by May 2011.

### 2011-04-29 paper.tex

--- 2011-04-26 paper.tex
@@
 
+Note that while some may advocate also for Open Access to
+publications, the goals focus more on the availability of code (to
+reproduce results), standards (to exchange data), and the scientific
+data itself. All three of these goals stem from the basics of the
+scientific method for data sharing and reproducibilty.
+
 The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
@@
 Open Babel and RDKit are planning to work together on a common
-underlying framework called MolCore. This project is still in the
-planning stage, but if it is a success it will mean that the two
+underlying framework called MolCore.\cite{WebMolCore} This project is still in the
+planning stage, but if it is a success it will mean that the the two
 libraries will be interoperable (while retaining their existing focus)

### 2011-05-02 paper.tex
two sentences about what brunn does

@@
      \and
+         Geoffrey R Hutchison$^6$%
+         \email{Geoffrey R Hutchison - geoffh@pitt.edu}
+     \and 
          Ola Spjuth$^9$%
-         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+         \email{ola.spjuth@farmbio.uu.se}
      \and 
@@
     \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(10)NIH, Somewhere in the US, USA
+    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA\\
+    \iid(11)NIH, Somewhere in the US, USA
 }%
@@
 such as Brunn~\cite{Brunn}, which is a laboratory information system for
-microplate based high-throughput screening. In this manner open source projects
-empower other open source projects.
+microplate based high-throughput screening. Brunn provides a grahical interface
+for handling different plate layouts and dilution series. It can automaticly
+generate dose response curves and calculate IC$_{50}$-values. In this manner
+open source projects empower other open source projects.
 

### 2011-05-03 paper.tex
two sentences about what brunn does
Added a uses dep diagram

--- 2011-04-29 paper.tex
@@
 in ODOSOS, leading to development of many useful resources freely
-available to the chemistry community. But how best to engage with the
+available to the chemistry community. Figure 2 shows how the various
+Blue Obelisk projects collaborate. But how best to engage with the
 wider chemistry community outside of the Blue Obelisk remains an open
@@
 
-  \subsection*{Figure 2 - Sample figure title}
-      Figure legend text.
+  \subsection*{Figure 2 - Dependency diagram of Blue Obelisk projects.}
+      Each block represents a project. Square blocks show Open Data, ovals are Open Source,
+      and diamonds are Open Standards. Colors represent license: LGPL is green, GPL is orange,
+      and BSD is blue.
 

### 2011-05-10 paper.tex
Added a uses dep diagram
Added my name, fixed e-mails.
An initial stab at edits

@@
 
-\author{Charles A Darwin\correspondingauthor$^{1,2}$%
-       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
+\author{
+  Jonathan Alvarsson$^1$%
+  \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
   \and
-         Egon L Willighagen$^3$%
-         \email{Egon L Willighagen - egon.willighagen@ki.se}
+  Igor V Filippov\correspondingauthor$^2$%
+  \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
   \and
-         Rajarshi Guha$^4$%
+  Rajarshi Guha$^3$%
   \email{Rajarshi Guha - guhar@mail.nih.gov}
   \and
-	     Christoph Steinbeck$^6$%
-	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+  Robert Hanson$^4$%
+  \email{Robert Hanson - hansonr@stolaf.edu}
   \and
-         Noel M O'Boyle$^5$%
-         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+  Geoffrey R Hutchison$^5$%
+  \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-         Ola Spjuth$^9$%
-         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
-     \and 
-         Jonathan Alvarsson$^9$%
-         \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
+  Daniel M Lowe$^6$%
+  \email{Daniel Lowe - dl387@cam.ac.uk}
   \and
@@
   \and
-         Daniel M Lowe$^7$%
-         \email{Daniel Lowe - dl387@cam.ac.uk}
+  Noel M O'Boyle$^8$%
+  \email{Noel M O'Boyle - n.oboyle@ucc.ie}
   \and
-         Robert Hanson$^8$%
-         \email{Robert Hanson - hansonr@stolaf.edu}
-      \and
-         Dmitry Pavlov$^8$%
+  Dmitry Pavlov$^9$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
   \and
-         Igor V Filippov\correspondingauthor$^{10}$%
-         \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
+  Ola Spjuth$^1$%
+  \email{ola.spjuth@farmbio.uu.se}
   \and
-         Jane E Doe\correspondingauthor$^2$%
-         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
+  Christoph Steinbeck$^{10}$%
+  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+  \and
+  Egon L Willighagen$^{11}$%
+  \email{Egon L Willighagen - egon.willighagen@ki.se}
     }
@@
 \address{%
-    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
-        Waterloo Road, London, UK\\
-    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
-    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
@@
 of small molecules was created as a resource about chemical structure and nomenclature
-by biologists \cite{DeMatos:2009p3839}.
+by biologists.\cite{DeMatos:2009p3839}
 
@@
 \begin{itemize}
+\item Open Data in Chemistry. One can obtain all scientific data in the public domain when
+wanted and reuse it for whatever purpose.
 \item Open Source in Chemistry. One can use other people's code without further
@@
 organisations, including various degrees of consent.
-\item Open Data in Chemistry. One can obtain all data in the public domain when
-wanted and reuse it for whatever purpose.
 \end{itemize}
@@
 Note that while some may advocate also for Open Access to
-publications, the goals focus more on the availability of code (to
+publications, the Blue Obelisk goals (ODOSOS) focus more on the availability of code (to
 reproduce results), standards (to exchange data), and the scientific
-data itself. All three of these goals stem from the basics of the
-scientific method for data sharing and reproducibilty.
+data itself. All three of these goals stem from the fundamental
+tenants of the scientific method for data sharing and reproducibilty.
 
@@
 contribution looks back on the past 5 years and surveys progress and
-remaining challenges in the areas of Open Source, Open Data and Open
+remaining challenges in the areas of Open Data, Open Source, and Open
 Standards in chemistry.
@@
 and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
-continue to be actively developed. [Insert main focus of work on CDK
-in last 5 years] Since 2006, major new features of Open Babel include 3D structure
+continue to be actively developed.
+
+[Insert main focus of work on CDK in last 5 years]
+
+Since 2006, major new features of Open Babel include 3D structure
 generation and 2D structure-diagram generation, UFF and MMFF94
@@
 chemistry calculations. In addition, a major focus of Open Babel development
-has been to fix problems in the areas of stereochemistry, kekulisation
-and canonicalisation.
-
-Some project have also been adopted by other fields, including biology
-and crystallography. A good example
-here is Jmol, which has, in the past five years, gone through two major revisions
-as part of an accelerated development focus. Starting with Jmol 10.2, released in
-April of 2006, Jmol 11.0 added over 100 new features and was released in March of
-2007. Three years later, Jmol 12.0 followed with over 700 new features.
-Based on the outstanding work of Michael (Miguel) Howard and others, this development
-path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
-visualisation package, including full support for crystallography~\cite{Hanson2010},
-efficient delivery of surface data using fast solvent-excluded surface
-generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
- display of molecular orbitals from standard basis set/coefficient data,
-the inclusion of dynamic minimisation using the UFF force field, and
-a full implementation of Daylight SMILES and SMART, with extensions to
-conformational and biomolecular substructure searching in the form of Jmol
-BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
-Behind the scenes, Jmol has been essentially completely rewritten. A full
-mathematical expression evaluator was added, and the scripting language
-was folded into a versatile JavaScript-like command language, including a
-complete set of program flow commands, several new data types, and
-user-defined variables and functions. An exciting non-chemical application
-of Jmol has been its incorporation into the open source Sage
-Mathematics Project [http://www.sagemath.org/], where Jmol is being used
-to deliver three-dimensional depiction of mathematical functions.
+has been to provide for accurate conversion and representation in
+areas of stereochemistry, kekulisation, and canonicalisation. The
+project has also grown, in terms of new contributors, new support from
+commercial companies, and second-generation tools applying Open Babel
+to a variety of end-user applications, from molecular editors to
+chemical database systems.
 
@@
 main 2D chemical rendering engine.
-
-While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented.
-As such, there is a growing need to present chemical and scientific information in a web-based medium.
-Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in JavaScript.
-This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education.
-It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial.
-As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilise these new technologies to improve products and education.
-A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs.
-The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception.
-ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra.
-Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices.
-It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets.
-iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry.
-This goal helps to make sure that the cost of education decreases while using the web to further spread science.
 
@@
 Taverna version 2.x.
-The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.
+The combination of CDK-Taverna workers with the multitude of workers
+already published on myexperiment.org by the large and active user
+community of Taverna enables scientists to quickly build workflows to
+process a diverse set of data as typically found in today's systems
+biology scenarios.
 
@@
 
+\subsubsection*{Web applications}
+
+While desktop software has composed the majority of scientific tools
+since the computer was introduced, the internet continues to change
+how applications and content are distributed and presented. As such,
+there is a growing need to present chemical and scientific information
+in a web-based medium. Recently, a new version of the HTML
+specification, HTML5 \cite{html5}, defines a well-developed framework
+for creating native web applications in JavaScript. This native
+framework provides a significant benefit to scientists as the web is
+an open and free medium to distribute scientific knowledge, ideas and
+education. It also poses significant new problems, as the cost
+associated with learning and developing tools for new technologies is
+not trivial. As such, it is an important goal for the Blue Obelisk to
+provide such tools and encourage the community to utilise these new
+technologies to improve products and education. A specific example of
+this initiative is the ChemDoodle Web Components library
+\cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web
+Components library is open source under the GPL v3 license with a
+liberal HTML exception. ChemDoodle Web Components allow the scientist
+to present publication quality 2D and 3D graphics and animations for
+chemical structures, reactions and spectra. Beyond graphics, this tool
+provides a framework for user interaction to create dynamic
+applications through web browsers, desktop platforms and mobile
+devices such as the iPhone, iPad and Android devices. It is now easier
+than ever for scientists to create web-based scientific content,
+without having to master complicated technologies such as Java
+applets. iChemLabs is dedicated to funding, developing and supporting
+the library to ensure that the next generation of scientific
+applications is easily achievable by academia, government and
+industry. This goal helps to make sure that the cost of education
+decreases while using the web to further spread science.
+
+Some projects have also been adopted by other fields, including biology
+and crystallography. A good example
+here is Jmol, which has, in the past five years, gone through two major revisions
+as part of an accelerated development focus. Starting with Jmol 10.2, released in
+April of 2006, Jmol 11.0 added over 100 new features and was released in March of
+2007. Three years later, Jmol 12.0 followed with over 700 new features.
+Based on the outstanding work of Michael (Miguel) Howard and others, this development
+path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
+visualisation package, including full support for crystallography~\cite{Hanson2010},
+efficient delivery of surface data using fast solvent-excluded surface
+generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
+ display of molecular orbitals from standard basis set/coefficient data,
+the inclusion of dynamic minimisation using the UFF force field, and
+a full implementation of Daylight SMILES[ref] and SMARTS[ref], with extensions to
+conformational and biomolecular substructure searching in the form of Jmol
+BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
+Behind the scenes, Jmol has been essentially completely rewritten. A full
+mathematical expression evaluator was added, and the scripting language
+was folded into a versatile JavaScript-like command language, including a
+complete set of program flow commands, several new data types, and
+user-defined variables and functions. An exciting non-chemical application
+of Jmol has been its incorporation into the open source Sage
+Mathematics Project [http://www.sagemath.org/], where Jmol is being used
+to deliver three-dimensional depiction of mathematical functions.
+
 \subsubsection*{The business end}
+
+% We should cite Craig James
+% Care and Feeding of FIOSS: http://www.moonviewscientific.com/essays/software_lifecycle.htm
+% And, of course, the Warren/Matt Stahl debate:
+%Warren L. DeLano. "The case for open-source software in drug discovery" Drug Discovery Today (2005) 10(3) p. 213-217. DOI:10.1016/S1359-6446(04)03363-X
+%Matthew T. Stahl. "Open-source software: not quite endsville" Drug Discovery Today (2005) 10(3) p. 219-222. DOI:10.1016/S1359-6446(04)03364-1
 
@@
 Molecular Networks GmBH, that does XXX.
+% Really? I've only seen commercial software from them -GRH
 

### 2011-05-11 paper.tex
Some trivial corrections as my first commit.

@@
 Since 2005 it has become evident that a good amount of development in open
-chemical information and information is driven by the demands of neighbouring
+chemical information is driven by the demands of neighbouring
 scientific fields. In many areas in biology, for example, the importance of
@@
 
-Rather than releasing new tools, an alternative approach to provide an
+Rather than releasing new tools, an alternative approach is to provide an
 interface to existing tools. hBar Solutions has developed an online
@@
 projects, Jmol and Open Babel, as well as the open source quantum
-mechanics packages MPQC.
+mechanics package MPQC.
 
@@
 projects focus on extracting chemical information from these sources.
-Such software plays a particularly important role for datamining the
+Such software plays a particularly important role for data mining the
 chemical literature, including patents and theses.
@@
 Optical Structure Recognition Application (OSRA) \cite{WebOSRA} was started
-in early 2007 with the goal to create first free and open source
+in early 2007 with the goal to create the first free and open source
 tool for extraction and conversion of molecular images into SMILES and
@@
 recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993},
-IUPAC have tried to codify and to an extent standardise naming practices.
+IUPAC has tried to codify and to an extent standardise naming practices.
 OPSIN aims to make this abundance of
@@
 Performance. There is a famous quote by Knuth that "Premature
-optimisation is root of all evil (or something)". Given that compute
+optimisation is the root of all evil (or something)". Given that compute
 time is cheap, ....
@@
 QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}.
-A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation.
+A dataset in QSAR-ML includes the chemical structures (preferably described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation.
 Hence, a dataset described in QSAR-ML is completely reproducible.
@@
 not compatible with the Panton Principles, but where the user
-is allows to modify and redistribute the data. A new data
+is allowed to modify and redistribute the data. A new data
 set in this category is the data from the ChEMBL database,
@@
 
-Importantly, publishing data as CC0 is become easier now that
-website are becoming available to simplify publishing data. Two
+Importantly, publishing data as CC0 is becoming easier now that
+websites are becoming available to simplify publishing data. Two
 projects that can be mentioned in this context are FigShare

### 2011-05-12 paper.tex
Text on BO books

--- 2011-05-10 paper.tex
@@
 \section*{Other areas of activity}
-Web services? XMPP services? (NIH, CDK and ChemSpider)
 
-\subsection*{Getting the message out}
-
-One recent development
-that attempts to address this issue was the establishment in Apr 2010 of a
-question-and-answer (Q\&A)
-website related to Blue Obelisk projects and themes at
-http://blueobelisk.shapado.com (see Figure~3). This is a website in the
+While each Blue Obelisk project has its own website and point of
+contact (typically a mailing list), because of the breadth of BO
+projects it can be difficult for a newcomer to understand which of
+them, if any, can best address a particular problem. To address this
+issue, members of the Blue Obelisk established a Question \& Answer
+website at http://blueobelisk.shapado.com (see Figure~3). 
+This is a website in the
 style of Stack Overflow that encourages high quality answers (and
@@
 200 users have registered, many of whom had no previous involvement
-with the Blue Obelisk, showing that the Q\&A website is a nice complement
-to earlier existing channels of communication.
+with the Blue Obelisk, showing that the Q\&A website complements
+earlier existing channels of communication.
 
-Books (Angel, Egon, OB)
+The rise of self-publishing and print-on-demand services has meant
+that publishing a book is now as straightforward as uploading to an
+appropriate website. Unlike the traditional publishing route where
+books with projected low sales volume would be expensive,
+websites such as Lulu allow the sale of low-priced books on
+chemistry software, and books are now available for purchase
+on Jmol~\cite{JmolBook}, the Chemistry Development Kit~\cite{CDKBook}
+and Open Babel~\cite{OpenBabelBook}.
 

### 2011-05-12 paper.tex
Shorten indigo and brunn text, and add webref for brunn

@@
 company (since closed down), released RDKit \cite{WebRDKit} under the
-BSD License. RDKit continues to be actively developed and includes
-code donated by Novartis.
-
-More recently, GGA Software Services (a contract programming company)
+BSD License. This is a C++ library with Python and (more recently)
+Java bindings. RDKit is actively developed and includes
+code donated by Novartis. Recent developments include the Java
+bindings, as well as performance improvements for its database
+cartridge. More recently, GGA Software Services
+(a contract programming company)
 released the Indigo toolkit \cite{WebIndigo} and associated software
-in 2009 under the GPL. Indigo is written from scratch in C++ and has
-high-level wrappers for using it in C, Java, Python, or .NET
-environment. Like other toolkits, Indigo provides support for
+in 2009 under the GPL. Indigo is a C++ library with
+high-level wrappers in C, Java, Python, and the .NET
+environment. Like RDKit and other toolkits, Indigo provides support for
 tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
-exact/substructure/SMARTS matching, fingerprint generation (all these
-both for molecules and reactions), and canonical SMILES computation.
+exact/substructure/SMARTS matching, fingerprint generation, and
+canonical SMILES computation.
 It also provides some less common functionality, like matching
@@
 finding maximum common substructure of $N$ input structures, and
-enumerating reaction products. Indigo also has its rendering plugin,
-which is used in a lot of commercial and open-source projects as the
-main 2D chemical rendering engine.
+enumerating reaction products.
 
@@
 tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
-such as Brunn~\cite{Brunn}, which is a laboratory information system for
+such as Brunn~\cite{WebBrunn}, a laboratory information system for
 microplate based high-throughput screening. Brunn provides a grahical interface
-for handling different plate layouts and dilution series. It can automaticly
-generate dose response curves and calculate IC$_{50}$-values. In this manner
-open source projects empower other open source projects.
+for handling different plate layouts and dilution series and can automaticly
+generate dose response curves and calculate IC$_{50}$-values.
 

### 2011-05-12 paper.tex
Combine bioclipse and avogadro paragraphs

@@
 on the CDK and Open Babel, respectively. Bioclipse is an award-winning
-molecular workbench for life sciences with a dynamic plugin architecture, and
+molecular workbench for life sciences
 which wraps cheminformatics functionality behind user-friendly interfaces and
-graphical editors. All functionality in Bioclipse is completely scriptable
-using an integrated high-level language called Bioclipse Scripting Language
-(BSL), which can be used for scripted research or to automate complex
-tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
-such as Brunn~\cite{WebBrunn}, a laboratory information system for
-microplate based high-throughput screening. Brunn provides a grahical interface
-for handling different plate layouts and dilution series and can automaticly
-generate dose response curves and calculate IC$_{50}$-values.
-
-Avogadro is a 3D molecular editor and viewer aimed at preparing
-and analysing computational chemistry calculations. An interesting
+graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
+preparing and analysing computational chemistry calculations. Both
+projects have are designed to be extended or scripted by users through
+a plugin architective and the provision of scripting support (using
+Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
+of Avogadro).
+An interesting
 aspect of both Avogadro and Bioclipse is that they share some developers
@@
 features in the CDK and Open Babel.
+
+Bioclipse in turn acts as base for other software
+such as Brunn~\cite{WebBrunn}, a laboratory information system for
+microplate based high-throughput screening. Brunn provides a graphical interface
+for handling different plate layouts and dilution series and can automaticly
+generate dose response curves and calculate IC$_{50}$-values.
 

### 2011-05-12 paper.tex
Tidy up section on 2nd gen tools

@@
 manipulating chemical structures, and can focus on providing
-additional functionality and ease-of-use.
+additional functionality and ease-of-use. Although a wide range of
+chemistry software has been built using Blue Obelisk
+components (see for example, the ``Related Software'' link on the Open
+Babel website, or ``Software using CDK'' at the CDK website), in this
+section we focus on second-generation tools which themselves have been
+developed by members of the Blue Obelisk.
 
@@
 molecular workbench for life sciences
-which wraps cheminformatics functionality behind user-friendly interfaces and
+that wraps cheminformatics functionality behind user-friendly interfaces and
 graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
 preparing and analysing computational chemistry calculations. Both
-projects have are designed to be extended or scripted by users through
-a plugin architective and the provision of scripting support (using
+projects are designed to be extended or scripted by users through
+the provision of a plugin architective and scripting support (using
 Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
@@
 
-CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
-Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc.
-CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to
-Taverna version 2.x.
-The combination of CDK-Taverna workers with the multitude of workers
-already published on myexperiment.org by the large and active user
-community of Taverna enables scientists to quickly build workflows to
-process a diverse set of data as typically found in today's systems
-biology scenarios.
-
-Some other recent projects that build on Blue Obelisk software include
-AMBIT (a GUI that facilitates registration of chemicals for the REACH
-EU directive on toxicity, based on the CDK), etc.
+A final example of 2nd-generation Blue Obelisk software is the 
+AMBIT2~\cite{WebAMBIT} software, a GUI that facilitates registration
+of chemicals for the REACH EU directive on toxicity, and which is
+based on the CDK.
 
@@
 create a depiction using RDKit.
+
+Another way through which interoperability of Blue Obelisk projects
+has been promoted and developed is through integration into
+workflow software such as Taverna~\cite{Hull:2006p60} and
+KNIME~\cite{WebKNIME} (both open source).
+Such software makes it easy to automate recurring
+tasks, and to combine analyses or data from a variety of different software
+and web services.
+A combination of the Chemistry Development Kit and Taverna, for instance, was
+reported in 2010~\cite{Kuhn:2010p4001}. 
+In the case of KNIME, it comes with built-in basic collection of CDK-based and
+Open Babel-based nodes, while other nodes for the RDKit and Indigo are
+available from KNIME's ``Community Updates'' site.
 

### 2011-05-12 paper.tex
Use publication format for PDF. Looks much nicer.
added Brunn reference

--- 2011-05-12 paper.tex
@@
 Bioclipse in turn acts as base for other software
-such as Brunn~\cite{WebBrunn}, a laboratory information system for
+such as Brunn~\cite{Alvarsson:2011fk}, a laboratory information system for
 microplate based high-throughput screening. Brunn provides a graphical interface

### 2011-05-12 paper.tex

--- 2011-05-12 paper.tex
@@
 %Review style settings
-\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 

### 2011-05-12 paper.tex
Added my name and contact.
Start tidying up web app section

@@
 Bioclipse in turn acts as base for other software
-such as Brunn~\cite{Alvarsson:2011fk}, a laboratory information system for
+such as Brunn~\cite{WebBrunn}, a laboratory information system for
 microplate based high-throughput screening. Brunn provides a graphical interface
@@
 since the computer was introduced, the internet continues to change
-how applications and content are distributed and presented. As such,
-there is a growing need to present chemical and scientific information
-in a web-based medium. Recently, a new version of the HTML
-specification, HTML5 \cite{html5}, defines a well-developed framework
-for creating native web applications in JavaScript. This native
-framework provides a significant benefit to scientists as the web is
+how applications and content are distributed and presented. The web
+presents new opportunities for scientists as it
 an open and free medium to distribute scientific knowledge, ideas and
-education. It also poses significant new problems, as the cost
-associated with learning and developing tools for new technologies is
-not trivial. As such, it is an important goal for the Blue Obelisk to
-provide such tools and encourage the community to utilise these new
-technologies to improve products and education. A specific example of
+education. Web applications are software that runs within the browser,
+typically implemented in Java or JavaScript.
+Recently, a new version of the HTML
+specification, HTML5, defines a well-developed framework
+for creating native web applications in JavaScript and opens up
+new possibilities for visualising chemical data.
+
+Jmol, the interactive 3D molecular viewer, is one of the most widely used
+chemistry applets, and indeed has
+seen widespread use in other fields such as biology and 
+even mathematics (it is used for 3D depiction of mathematical
+functions in the Sage Mathematics Projects~\cite{WebSage}). It is implemented
+in Java, and has gone from being a ``Rasmol/Chime'' replacement to a fully fledged molecular
+visualisation package, including full support for crystallography~\cite{Hanson2010},
+display of molecular orbitals from standard basis set/coefficient data,
+the inclusion of dynamic minimisation using the UFF force field, and
+a full implementation of Daylight SMILES and SMARTS, with extensions to
+conformational and biomolecular substructure searching (Jmol
+BioSMARTS).
+
+A specific example of
 this initiative is the ChemDoodle Web Components library
@@
 decreases while using the web to further spread science.
-
-Some projects have also been adopted by other fields, including biology
-and crystallography. A good example
-here is Jmol, which has, in the past five years, gone through two major revisions
-as part of an accelerated development focus. Starting with Jmol 10.2, released in
-April of 2006, Jmol 11.0 added over 100 new features and was released in March of
-2007. Three years later, Jmol 12.0 followed with over 700 new features.
-Based on the outstanding work of Michael (Miguel) Howard and others, this development
-path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
-visualisation package, including full support for crystallography~\cite{Hanson2010},
-efficient delivery of surface data using fast solvent-excluded surface
-generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
- display of molecular orbitals from standard basis set/coefficient data,
-the inclusion of dynamic minimisation using the UFF force field, and
-a full implementation of Daylight SMILES[ref] and SMARTS[ref], with extensions to
-conformational and biomolecular substructure searching in the form of Jmol
-BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
-Behind the scenes, Jmol has been essentially completely rewritten. A full
-mathematical expression evaluator was added, and the scripting language
-was folded into a versatile JavaScript-like command language, including a
-complete set of program flow commands, several new data types, and
-user-defined variables and functions. An exciting non-chemical application
-of Jmol has been its incorporation into the open source Sage
-Mathematics Project [http://www.sagemath.org/], where Jmol is being used
-to deliver three-dimensional depiction of mathematical functions.
 

### 2011-05-12 paper.tex

@@
   \and
+  Kevin J Theisen$^{12}$%
+  \email{Kevin J Theisen - kevin@ichemlabs.com}
+  \and
   Egon L Willighagen$^{11}$%
@@
     \iid(11)NIH, Somewhere in the US, USA
+    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA
 }%

### 2011-05-12 paper.tex
Improved readability of OPSIN section. Changed OPSIN associated from CML-Xom to CML Removed orphaned reference

@@
 
-The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC
+OPSIN (Open Parser for Systematic IUPAC
 Nomenclature)~\cite{lowe_chemical_2011} focuses instead on interpreting chemical names.
@@
 guide tokenisation and parsing of chemical names, followed by
-step-wise application of nomenclature rules. OPSIN is able to offer
+step-wise application of nomenclature rules. It is able to offer
 fast and precise conversions for the majority of names using IUPAC

### 2011-05-13 paper.tex
Fixed a few author-affiliation links

@@
   \and
-  Igor V Filippov\correspondingauthor$^2$%
-  \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
+  Igor V Filippov$^2$%
+  \email{Igor V Filippov - igorf@helix.nih.gov}
   \and
@@
   \and
-  Noel M O'Boyle$^8$%
-  \email{Noel M O'Boyle - n.oboyle@ucc.ie}
+  Noel M O'Boyle\correspondingauthor$^8$%
+  \email{Noel M O'Boyle\correspondingauthor - n.oboyle@ucc.ie}
   \and
@@
 \address{%
-   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+   \iid(11)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
@@
     \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
-    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
+    \iid(7)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
     \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
-    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
+    \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
     \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA\\
-    \iid(11)NIH, Somewhere in the US, USA
+    \iid(3)NIH, Somewhere in the US, USA\\
     \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA
@@
 \section*{Open Source}
-  \subsection*{Progress}
 
-\subsubsection*{Cheminformatics toolkits}
+\subsection*{Cheminformatics toolkits}
 
@@
 
-\subsubsection*{Second-generation tools}
+\subsection*{Second-generation tools}
 
@@
 
-\subsubsection*{Web applications}
+\subsection*{Web applications}
 
@@
 
-A specific example of
-this initiative is the ChemDoodle Web Components library
-\cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web
-Components library is open source under the GPL v3 license with a
-liberal HTML exception. ChemDoodle Web Components allow the scientist
+In 2009, iChemLabs released the ChemDoodle Web Components
+library~\cite{ChemDoodleWeb} under the GPL v3 license (with a
+liberal HTML exception). This library is completely implemented in JavaScript
+and uses HTML5 to allow the scientist
 to present publication quality 2D and 3D graphics and animations for
@@
 applications through web browsers, desktop platforms and mobile
-devices such as the iPhone, iPad and Android devices. It is now easier
-than ever for scientists to create web-based scientific content,
-without having to master complicated technologies such as Java
-applets. iChemLabs is dedicated to funding, developing and supporting
-the library to ensure that the next generation of scientific
-applications is easily achievable by academia, government and
-industry. This goal helps to make sure that the cost of education
-decreases while using the web to further spread science.
+devices such as the iPhone, iPad and Android devices.
 
-\subsubsection*{The business end}
+\subsection*{The business end}
 
@@
 
-\subsubsection*{Converting chemical names and images to structures}
+\subsection*{Converting chemical names and images to structures}
 
@@
 
-  \subsubsection*{Collaboration and interoperability}
+  \subsection*{Collaboration and interoperability}
 
@@
 
-Accuracy. Very often software work at the 95\% to 98\% level. The
-variety of chemical structures is such that with a large enough
-dataset, 'unusual' structures are always found which may be mishandled
-by software. Given that much of the development of open source
-software is unfunded, and relies heavily on developer motivation, it
-is understandable that working on the final N\% of problem structures
-(which may require substantial work) is not the most exciting of
-tasks. Putting a bounty system in place may be useful for these
-situations where a researcher needs a problem to be fixed because it
-affects a particular dataset of interest.
+(Say something here about benchmarks to measure accuracy. Clear examples
+ of performance on open datasets are required. Otherwise there is
+ nothing to counter anecdotal evidence or FUD spread by others.)
 
-Performance. There is a famous quote by Knuth that "Premature
-optimisation is the root of all evil (or something)". Given that compute
-time is cheap, ....
+(Better engagement with industry...make it clear what and how industry
+ members can engage with projects.)
 
@@
 \section*{Open Standards}
-  \subsection*{Progress}
+  \subsection*{InChI}
 
@@
 
-    \subsubsection*{OpenSMILES}
+    \subsection*{OpenSMILES}
 
@@
 
-\subsubsection*{CML}
+\subsection*{CML}
 ?
 
-\subsubsection*{QSAR-ML}
+\subsection*{QSAR-ML}
 The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses.
@@
 Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
-
-\subsubsection*{Others...?}
-
-Pistoia Alliance (and others?) should be mentioned at this point
 
@@
 \section*{Open Data}
-  \subsection*{Progress}
 
@@
 are expected to become soon a substantial knowledge base.
-
-  \subsection*{Remaining challenges}
 

### 2011-05-13 paper.tex
Edited paper.tex via GitHub
Tidy up end of web apps

--- 2011-05-12 paper.tex
@@
 applications through web browsers, desktop platforms and mobile
-devices such as the iPhone, iPad and Android devices. It is now easier
-than ever for scientists to create web-based scientific content,
-without having to master complicated technologies such as Java
-applets. iChemLabs is dedicated to funding, developing and supporting
-the library to ensure that the next generation of scientific
-applications is easily achievable by academia, government and
-industry. This goal helps to make sure that the cost of education
-decreases while using the web to further spread science.
+devices such as the iPhone, iPad and Android devices.
 

### 2011-05-13 paper.tex
Change subsub sections to sub sections

--- 2011-05-13 paper.tex
@@
   \and
-  Igor V Filippov$^2$%
-  \email{Igor V Filippov - igorf@helix.nih.gov}
+  Igor V Filippov\correspondingauthor$^2$%
+  \email{Igor V Filippov\correspondingauthor - igorf@helix.nih.gov}
   \and
@@
   \and
-  Noel M O'Boyle\correspondingauthor$^8$%
-  \email{Noel M O'Boyle\correspondingauthor - n.oboyle@ucc.ie}
+  Noel M O'Boyle$^8$%
+  \email{Noel M O'Boyle - n.oboyle@ucc.ie}
   \and
@@
 \address{%
-   \iid(11)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
@@
     \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
-    \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
+    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
     \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA\\
-    \iid(3)NIH, Somewhere in the US, USA\\
+    \iid(11)NIH, Somewhere in the US, USA
     \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA

### 2011-05-13 paper.tex
Cleaned up the Business End section

--- 2011-05-13 paper.tex
@@
 
-Open Source provides a unique opportunity for commercial organisations to work with the community.
-As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity.
-This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project.
-Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry.
-A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.
+Open Source provides a unique opportunity for commercial organisations to work with the
+cheminformatics community. Traditional business models rely on monetisation the source
+code, causing every company to repeat the development by other companies, often combined
+with a free (gratis) model for people working at academic institutes, to increase adoption
+and allow contributions from academics. This solution defines the IP on the software
+return on investment, but has the downside of investment losses due to duplication of
+software and method development, which becomes visible when proprietary companies
+merge.
 
-Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license.
-With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources.
-Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development.
-Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it.
-These benefits allow a company to create a successful product with significantly reduced cost and effort.
+The Blue Obelisk community, however, takes advantage that much of the investments needed
+for development are either payed by academic institutes and funding schemes, and by
+volunteers investing time and effort who get full access to the source code in return.
+In fact, all partners get full access up front and the Open Source licending ensures
+that they will have access any time in the future. As such, it functions as a social
+contract between everyone to arrange the immediate return on investment. Effectively,
+this approach shares the burden of the high investment in having to develop cheminformatics
+software from scratch, allowing researchers and commercial partners alike to focus
+on their core business, rather than the development of prerequisites. As such, the rich
+collection of Open Source cheminformatics tools provided by the Blue Obelisk
+greatly reduces investment up front for new companies in the cheminformatics market.
 
-Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded.
-In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass.
-This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities.
-On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption.
-In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development.
-This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.
+The use of Open Standards allows everyone to select those Blue Obelisk components
+they find most useful, as they can easily replace one component with another providing
+the same functionality, taking advantage that they use the same standards for,
+for example, data exchange. This way, licensing issues are becoming a marginal
+problem, allowing companies to select a license appropriate for their business
+model. This too, allows a company to create a successful product with significantly
+reduced cost and effort.
 
-In the scientific industry, it is incredibly difficult to start successful open projects.
-This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment.
-Because of this, there is an increased barrier to starting open projects in the scientific industry.
-The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects.
-Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.
+At the time of writing there are many commercial companies development chemistry
+solutions around Open Source cheminformatics components provided by the Blue Obelisk
+community. Examples of such companies include IChemLabs, IdeaConsult, Wingu, Silicos,
+GenettaSoft, hBar, and Inkspot Science. Some of these merely use components, but several
+actively contribute back to the Blue Obelisk project they use, or donate new
+Open Source cheminformatics projects to the community.
 
-For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing.
-The license is chosen to best ensure the success of the project.
-iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception.
-iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content.
-Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market.
-Of course, no one will need to waste time rewriting functionality that is already provided with quality.
-Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
+For example, iChemLabs released the ChemDoodle Web Components library under the GPL v3
+license, based on the upcoming HTML5 Open Standard. It allows making web and mobile
+interfaces for chemical content. The project is already being adopted by others,
+including iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by
+Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
 
-An example of such an interaction was the donation by eMolecules, Inc.
-to Open Babel of code for the canonicalisation of molecules and
-fragments (Nov 2006). eMolecules is an online vendor of chemicals that
-uses Open Babel under-the-hood to manage its compound collections.
+Silicos has release several Open Source utilities based on
+OpenBabel, such as Pharao, a tool for pharmacophore searching,
+Sieve for filtering molecular structure by molecular property,
+Stripper for removing core scafold structures from a molecule
+set, and Piramid for molecular alignment using shape determined
+by the Gaussian volumes as a descriptor. Additionally,
+contributions have been made to the OpenBabel project itself.
 
-Another more recent example occurred in July 2010 when Silicos, a
-Belgian company that provides services
-in the area of cheminformatics, released several command line
-applications based on Open Babel as well as donating code to the
-project. For example, the Pharao
-tool released by them is a comprehensive solution for pharmacophore
-searching that provides extensive support for a variety of
-pharmacophore searches. Other tools released by Silicos include Sieve for
-filtering by molecular property, Stripper for removing core scaffold
-structures from a molecule set, and Piramid for molecular alignment
-using shape determined by the Gaussian volumes as a descriptor.
-
-Another example of Open Source
-tools originating from commercial groups is the ChemCraft tool from
-Molecular Networks GmBH, that does XXX.
-% Really? I've only seen commercial software from them -GRH
-
-Rather than releasing new tools, an alternative approach is to provide an
-interface to existing tools. hBar Solutions has developed an online
-portal, hBar Lab, for managing and performing computational chemistry
-calculations in the cloud. To do so it leverages two Blue Obelisk
-projects, Jmol and Open Babel, as well as the open source quantum
-mechanics package MPQC.
+Other companies use Blue Obelisk components and contribute patches,
+smaller and larger. For example, IXELES donated the isomorphism
+code in the CDK, eMolecules donated canonicalisation code to
+OpenBabel, and AstraZeneca contributed code to the CDK for
+signatures. This is just a very minor selection, and the author
+is encouraged to contact the individual Blue Obelisk projects
+for an elaborate list.
 

### 2011-05-13 paper.tex
Tidy up the text on InChI and add ref for JNI-InChI

--- 2011-05-13 paper.tex
@@
 compilations. Although its development predates the Blue Obelisk,
-software such as Open Babel has included InChI support since 2005.
+software such as Open Babel has included InChI support since 2005,
+and support for InChI in Indigo is due in 2011.
+
 Since the official InChI implementation is in C, it is difficult to
 access from the other widely used language for cheminformatics
-toolkits, Java. The Blue Obelisk project JNI-InChI has been set up to
+toolkits, Java. The Blue Obelisk project JNI-InChI~\cite{WebJNIInChI}
+was established in 2006 to
 solve this problem by using the Java Native Interface to link the

### 2011-05-13 paper.tex
Added subsection about computational chemistry and info/cite to cclib.

--- 2011-05-13 paper.tex
@@
 based on the CDK.
+
+\subsection*{Facilitating chemical computations}
+
+An area where cheminformatic tools can make a signficant impact is supporting
+chemical computations and interpreting their results.
+This is especially true for electronic structure calculations since many
+quantum chemistry methods already have a long tradition in the scientific
+community -- most prominently the Hartree-Fock method and density functional theory,
+as well as more accurate variations of these.
+Many methods have been implemented and reimplemented in a variety of software
+packages over the last 30 years.
+When a computational chemist performs calculations using several programs,
+he/she will find that the format of their input files and outputs
+differ enourmously.
+Sometimes even different version of the same program implement different input
+syntax and output formats.
+Furthermore, some of these programs provide additional algorithms such as
+populations analyses, which in general are applicable to all packages.
+
+Obtaining electronic structure is often the first step in a research pipeline,
+after which data can be further analyzed or visualized.
+Preparing inputs and some basic analyses, usually with the
+goal of visualization, are often a part of graphical molecular editors.
+Avogadro for example is capable of generating input files for several
+electronic structure packages, and visualizes molecular orbitals.
+
+The Blue Obelisk project  cclib\cite{cclib} is dedicated to parsing output
+from computational chemistry programs.
+It's main goal is to read as many different kinds of results from as
+many different packages as possible.
+After parsing, cclib provides all results in standardized form as attributes
+of a python object Python.
+This means that if a developer implements an algorithm based on such an API,
+they can feed it output files from any of the supported computational software.
+The current version of cclib (1.0.1) supports 8 different computational chemistry
+codes and extracts over 30 different calculations results.
 
@@
         Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
-
+        cclib  & http://cclib.sourceforge.net & Python library for parsing and interpreting computational chemistry results \\ \hline
         \end{tabular}

### 2011-05-13 paper.tex
Move database section to before collaboration (so that we can talk about collaboration between databases *after* we have introduced them).

--- 2011-05-13 paper.tex
@@
 
+\subsection*{Chemical Databases}
+
+Registration, indexing and searching of chemical structures in
+relational databases is one of the core areas of cheminformatics.
+A number of structure registration systems have been published in the last five years, exploiting the fact that
+free cheminformatics toolkits such as OpenBabel and the CDK were available.
+OrChem, for example, is an extension for the Oracle 11G database that
+adds registration and indexing of chemical structures to support fast
+substructure and similarity searching. The cheminformatics
+functionality is provided by the Chemistry Development Kit. OrChem
+provides similarity searching with response times in the order of
+seconds for databases with millions of compounds, depending on a given
+similarity cut-off. For substructure searching, it can make use of
+multiple processor cores on today's powerful database servers to
+provide fast response times in equally large data sets.
+OrChem is free software and can be redistributed and/or modified under
+the terms of the GNU Lesser General Public License as published by the
+Free Software Foundation. All software is available via
+http://orchem.sourceforge.net.
+
   \subsection*{Collaboration and interoperability}
@@
  members can engage with projects.)
-
-\section*{Chemical Structure Registration Systems, Databases}
-Databases? (XXX CS: A couple of database-related paragraphs could go here XXX)
-
-XXX the paragraph below on OrChem needs some work
-Registration, indexing and searching of chemical structures in relational databases is one of the core areas of cheminformatics.
-A number of structure registration systems have been published in the last five years, exploiting the fact that
-free cheminformatics toolkits such as OpenBabel and the CDK were available.
-OrChem, for example, is an extension for the Oracle 11G database that adds registration and indexing of chemical structures to support fast substructure and similarity searching.
-The cheminformatics functionality is provided by the Chemistry Development Kit.
-OrChem provides similarity searching with response times in the order of seconds for databases with millions of compounds, depending on a given similarity cut-off.
-For substructure searching, it can make use of multiple processor cores on today's powerful database servers to provide fast response times in equally large data sets.
-OrChem is free software and can be redistributed and/or modified under the terms of the GNU Lesser General Public License as published by the Free Software Foundation.
-All software is available via http://orchem.sourceforge.net.
 

### 2011-05-13 paper.tex
Tidy up comp chem section

--- 2011-05-13 paper.tex
@@
 based on the CDK.
+
+\subsection*{Computational chemistry analysis}
+
+Another area where the Blue Obelisk has had a signficant impact in the
+past five years is in supporting
+quantum chemistry calculations and in interpreting their results.
+Electronic structure calculations have a long tradition in the
+chemistry community and a variety of programs exist, mostly
+proprietary software with the notable exception of NWChem.
+However, since each program uses different input formats, and the
+results are output differently even varying between different versions
+of the software, preparing calculations and extracting the results in
+an automated way is problematic.
+
+Avogadro has already been mentioned as a GUI for preparing
+computational chemistry calculations. It uses Open Babel to read the
+output of several electronic structure packages, and also to write the
+input files for each. Jmol (see below) can also depict computational
+chemistry results including molecular orbitals.
+
+In 2006, the Blue Obelisk project cclib~\cite{cclib} was established
+with the goal of parsing the output from computational chemistry
+programs and presenting it in a standard way so that further analyses
+could be carried out independently of the quantum package used to
+obtain the results. cclib is a Python library, and
+the current version (version 1.0.1)
+supports 8 different computational chemistry codes and extracts over
+30 different calculated attributes. A related BO project,
+GaussSum~\cite{WebGaussSum},
+is a GUI that builds on top of cclib to allow predicted properties to
+be compared to experimental properties (for example, the calculated
+vibrational frequencies and the IR spectrum).
+
+[Few lines on Quixote here]
 

### 2011-05-13 paper.tex
Fix formatting of some references

@@
 Open Babel and RDKit are planning to work together on a common
-underlying framework called MolCore.\cite{WebMolCore} This project is still in the
+underlying framework called MolCore~\cite{WebMolCore}. This project is still in the
 planning stage, but if it is a success it will mean that the the two

### 2011-05-13 paper.tex
Modified cclib section by clarifying GaussSum (and adding text about it's other useful features!); Added text for QMForge.

@@
 supports 8 different computational chemistry codes and extracts over
-30 different calculated attributes. A related BO project,
-GaussSum~\cite{WebGaussSum},
-is a GUI that builds on top of cclib to allow predicted properties to
-be compared to experimental properties (for example, the calculated
-vibrational frequencies and the IR spectrum).
+30 different calculated attributes. Two related BO projects build upon 
+cclib. GaussSum~\cite{WebGaussSum},
+is a GUI that can monitors the progress of SCF and geometry convergences, 
+and can plot predicted UV/Vis absorption and infrared spectra from 
+appropriate logfiles containing energies and oscillator strengths for easy 
+comparison to experimental data. QMForge~\cite{WebQMForge} provides 
+a GUI for various electronic structure analyses such as Frenking's charge 
+decomposition analysis~\cite{Frenking} and Mulliken or C-squared analyses
+on user-defined molecular fragments. QMForge also provides a rudamentary
+Cartesian coordinate editor allowing molecular structures to be saved via OpenBabel.
 

### 2011-05-13 paper.tex
Added myself and affiliation.
Added my name, email, and affiliation

@@
   \and
+  Karol M Langner$^{13}$%
+  \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
+  \and
   Daniel M Lowe$^6$%
@@
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+  \and
+  Adam L Tenderholt$^{14}$%
+  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
@@
     \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA
+    \iid(13)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands
+    \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195
 }%

### 2011-05-13 paper.tex
Updated many website, references, formatting.

@@
 
-\title{Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on}
+\title{Open Data, Open Source and Open Standards in chemistry: \\ The Blue Obelisk five years on}
 
@@
   \and
-  Rajarshi Guha$^3$%
+  Rajarshi Guha$^2$%
   \email{Rajarshi Guha - guhar@mail.nih.gov}
   \and
-  Robert Hanson$^4$%
+  Robert Hanson$^3$%
   \email{Robert Hanson - hansonr@stolaf.edu}
   \and
-  Geoffrey R Hutchison$^5$%
+  Geoffrey R Hutchison$^4$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Karol M Langner$^{13}$%
+ Karol M Langner$^5$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
@@
   \and
-  Adam L Tenderholt$^{14}$%
+ Adam L Tenderholt$^{11}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
@@
   \and
-  Egon L Willighagen$^{11}$%
+  Egon L Willighagen$^{13}$%
   \email{Egon L Willighagen - egon.willighagen@ki.se}
@@
 \address{%
-   \iid(11)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
+    \iid(2)National Institute of Health, Somewhere in the US, USA\\
     \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
@@
    \iid(3)NIH, Somewhere in the US, USA\\
-    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA
-    \iid(13)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands
+    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ    08854, USA \\
+    \iid(13)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden,    The Netherlands \\
     \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195
@@
 		contribution looks back on the past 5 years and surveys progress and
-		remaining challenges in the areas of Open Source, Open Data and Open
-		Standards in chemistry.
+		remaining challenges in the areas of Open Data, Open Standards, and
+    Open Source in chemistry.
 
@@
 \begin{itemize}
-\item Open Data in Chemistry. One can obtain all scientific data in the public domain when
+\item {\bf Open Data in Chemistry.} One can obtain all scientific data in the public domain when
 wanted and reuse it for whatever purpose.
-\item Open Source in Chemistry. One can use other people's code without further
-permission, including changing it for one's own use and distributing
-it again.
-\item Open Standards in Chemistry. One can find visible community mechanisms for
+\item {\bf Open Standards in Chemistry.} One can find visible community mechanisms for
 protocols and communicating information. The mechanisms for creating
@@
 organisations, including various degrees of consent.
+\item {\bf Open Source in Chemistry.} One can use other people's code without further
+permission, including changing it for one's own use and distributing
+it again.
 \end{itemize}
@@
 Note that while some may advocate also for Open Access to
-publications, the Blue Obelisk goals (ODOSOS) focus more on the availability of code (to
-reproduce results), standards (to exchange data), and the scientific
-data itself. All three of these goals stem from the fundamental
-tenants of the scientific method for data sharing and reproducibilty.
+publications, the Blue Obelisk goals (ODOSOS) focus more on the
+availability of the underlying scientific data, standards (to exchange
+data), and code (to reproduce results). All three of these goals stem
+from the fundamental tenants of the scientific method for data sharing
+and reproducibilty.
 
@@
 later as a formal paper by Guha et al.\cite{Guha2006} in
-2006. This
-contribution looks back on the past 5 years and surveys progress and
-remaining challenges in the areas of Open Data, Open Source, and Open
-Standards in chemistry.
+2006. This contribution looks back on the past 5 years and surveys
+progress and remaining challenges in the areas of Open Data, Open
+Source, and Open Standards in chemistry.
 
@@
 components (see for example, the ``Related Software'' link on the Open
-Babel website, or ``Software using CDK'' at the CDK website), in this
+Babel website,\cite{WebOBRelated} listing over 40 projects as of this writing, 
+or ``Software using CDK'' at the CDK website), in this
 section we focus on second-generation tools which themselves have been
@@
 Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
-of Avogadro).
-An interesting
-aspect of both Avogadro and Bioclipse is that they share some developers
-with the underlying toolkits and this has driven the development of new
-features in the CDK and Open Babel.
+of Avogadro). An interesting aspect of both Avogadro and Bioclipse is
+that they share some developers with the underlying toolkits and this
+has driven the development of new features in the CDK and Open Babel.
 
-Bioclipse in turn acts as base for other software
+Both products in turn act as extensible platforms for other
+software. Bioclipse, for example is used by software
 such as Brunn~\cite{Alvarsson:2011fk}, a laboratory information system for
@@
 for handling different plate layouts and dilution series and can automaticly
-generate dose response curves and calculate IC$_{50}$-values.
+generate dose response curves and calculate IC$_{50}$-values. Avogadro
+is used by XtalOpt~\cite{WebXtalOpt}\cite{Lonie2011}, an evolutionary
+algorithm for crystal structure prediction. XtalOpt provides a
+graphical interface using Avogadro and submits calculations using a
+range of solid-state simulation software to predict stable polymorphs.
 
@@
 % Care and Feeding of FIOSS: http://www.moonviewscientific.com/essays/software_lifecycle.htm
-% And, of course, the Warren/Matt Stahl debate:
-%Warren L. DeLano. "The case for open-source software in drug discovery" Drug Discovery Today (2005) 10(3) p. 213-217. DOI:10.1016/S1359-6446(04)03363-X
-%Matthew T. Stahl. "Open-source software: not quite endsville" Drug Discovery Today (2005) 10(3) p. 219-222. DOI:10.1016/S1359-6446(04)03364-1
 
 Open Source provides a unique opportunity for commercial organisations to work with the
-cheminformatics community. Traditional business models rely on monetisation the source
+cheminformatics community. Traditional business models rely on monetisation of source
 code, causing every company to repeat the development by other companies, often combined
@@
 software and method development, which becomes visible when proprietary companies
-merge.
+merge. Some authors have rationalized that in the chemistry field, the
+data has value, and few contributors are available to volunteer time
+to improve codes.\cite{Stahl:2005fk} If true, this would hamper
+adoption of Open Source and Open Data in chemistry, and greatly slow
+the growth of projects such as those in the Blue Obelisk.
 
@@
 collection of Open Source cheminformatics tools provided by the Blue Obelisk
-greatly reduces investment up front for new companies in the cheminformatics market.
+greatly reduces investment up front for new companies in the
+cheminformatics market. Such advantages have also been noted in the
+drug discovery field.\cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}
 
@@
 free cheminformatics toolkits such as OpenBabel and the CDK were available.
-OrChem, for example, is an extension for the Oracle 11G database that
+OrChem,\cite{WebOrChem} for example, is an open source extension for the Oracle 11G database that
 adds registration and indexing of chemical structures to support fast
 substructure and similarity searching. The cheminformatics
-functionality is provided by the Chemistry Development Kit. OrChem
+functionality is provided by the CDK. OrChem
 provides similarity searching with response times in the order of
@@
 provide fast response times in equally large data sets.
-OrChem is free software and can be redistributed and/or modified under
-the terms of the GNU Lesser General Public License as published by the
-Free Software Foundation. All software is available via
-http://orchem.sourceforge.net.
 
@@
 were looking to add support for energy minimisation, rather than
-implement a forcefield from scratch they ported the UFF forcefield
+implement a forcefield from scratch they ported the UFF forcefield\cite{Rappe:1992um}
 implementation from Open Babel to Jmol. This code has allowed Jmol to
@@
 the ChemiSQL project. This brought together the developers of several
-open source chemistry database cartridges (PgChem, MyChem, OrChem and
-more recently Bingo) with a view to making their database APIs more
+open source chemistry database cartridges (PgChem,\cite{WebPgChem} MyChem,\cite{WebMyChem} OrChem\cite{WebOrChem} and
+more recently Bingo\cite{WebBingo}) with a view to making their database APIs more
 similar and collaborating on benchmark datasets for assessing
@@
 A considerable stumbling block in advocating the release of scientific
-data as Open Data has been how exactly to define Open. A major step
+data as Open Data has been how exactly to define ``Open.'' A major step
 forward was the launch in 2010 of the Panton Principles for Open Data
@@
 Open Data Commons Attribution License, and the
-Open Data Commons Open Database License (ODbL) (see
-http://www.opendefinition.org/licenses/\#Data).
+Open Data Commons Open Database License (ODbL).\cite{WebOpenData}
 
@@
 exceptions: ChemPedia (a now discontinued crowd-sourcing project),
-CrystalEye (http://wwmm.ch.cam.ac.uk/crystaleye/),
+CrystalEye,\cite{WebCrystalEye}
 and the Open Notebook Science Solubility
@@
 websites are becoming available to simplify publishing data. Two
-projects that can be mentioned in this context are FigShare
-(http://figshare.com/), where the data behind unpublished figures
-can be hosted, and Dryad (http://datadryad.org/) where data
-behind publications can be hosted. Initiatives like this make
-it possible to host small amounts of data, and those combined
-are expected to become soon a substantial knowledge base.
+projects that can be mentioned in this context are
+FigShare\cite{WebFigShare}, where the data behind unpublished figures
+can be hosted, and Dryad\cite{WebDryad} where data behind publications
+can be hosted. Initiatives like this make it possible to host small
+amounts of data, and those combined are expected to become soon a
+substantial knowledge base.
 
@@
 issue, members of the Blue Obelisk established a Question \& Answer
-website at http://blueobelisk.shapado.com (see Figure~3). 
-This is a website in the
-style of Stack Overflow that encourages high quality answers (and
+website\cite{WebBOShapado} (see Figure~3).  This is a website in the
+style of Stack Overflow\cite{WebStackOverflow} that encourages high quality answers (and
 questions) through the use of a voting system. In the year since it
-was established, over
-200 users have registered, many of whom had no previous involvement
-with the Blue Obelisk, showing that the Q\&A website complements
-earlier existing channels of communication.
+was established, over 200 users have registered, many of whom had no
+previous involvement with the Blue Obelisk, showing that the Q\&A
+website complements earlier existing channels of communication.
 
@@
 books with projected low sales volume would be expensive,
-websites such as Lulu allow the sale of low-priced books on
+websites such as Lulu\cite{WebLulu} allow the sale of low-priced books on
 chemistry software, and books are now available for purchase
@@
 community grow in the future. To this end, we encourage the reader to
-visit http://blueobelisk.org, send a message to our mailing list,
+visit the Blue Obelisk website\cite{WebBlueObelisk}, send a message to our mailing list,
 investigate related projects or read our blogs.
@@
     (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
+      %% GeoffH - I think we should do this as multi-line, with the
+      %% website below the Name
     \par \mbox{}
@@
      \begin{tabular}{|c|c|c|}
-        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        \hline Name & Website & Description? or Lead Developer? \\ \hline
-        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
-        Chemistry Development Kit (CDK) \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
-        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
-        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
-        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
-        RDKit & http://rdkit.org & Greg Landrum \\ \hline
-        ChemDoodle Web Components & http://web.chemdoodle.com & iChemLabs \\ \hline
-        \multicolumn{3}{|c|}{Integration} \\ \hline
-        CDK-Taverna \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
-        A3 & ..  & .  \\ \hline
-        \multicolumn{3}{|c|}{Interconversion} \\ \hline
-        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
-        OPSIN & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
-        A3 & ..  & .  \\ \hline
-        \multicolumn{3}{|c|}{Structure Databases} \\ \hline
-        OrChem \cite{RijnbeekS10} & http://orchem.sourceforge.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
-        MyChem  & http://XXX.sourceforge.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
-        PGChem  & http://XXX.sourceforge.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
-        Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
-        cclib  & http://cclib.sourceforge.net & Python library for parsing and interpreting computational chemistry results \\ \hline
+       \hline \textbf{Name} & \textbf{Website} & \textbf{Description} \\ \hline
+        \multicolumn{3}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
+        \textbf{Chemistry Development Kit (CDK)} \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
+        \textbf{Cinfony} & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
+        \textbf{Indigo} & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
+        \textbf{Open Babel} & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
+        \textbf{RDKit} & http://rdkit.org & Greg Landrum \\ \hline
+        \multicolumn{3}{|c|}{\textbf{Second Generation}} \\ \hline
+        \textbf{Avogadro} & http://avogadro.openmolecules.net & XXX \\ \hline
+        \multicolumn{3}{|c|}{\textbf{Web Applications}} \\ \hline
+        \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com & iChemLabs \\ \hline
+        \textbf{Jmol} & http://jmol.org & Hanson \\ \hline
+       \multicolumn{3}{|c|}{\textbf{Integration}} \\ \hline
+        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\        \hline
+       \textbf{A3} & ..  & .  \\ \hline
+        \multicolumn{3}{|c|}{\textbf{Interconversion}} \\ \hline
+        \textbf{OSRA} & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
+        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
+        \textbf{A3} & ..  & .  \\ \hline
+        \multicolumn{3}{|c|}{\textbf{Structure Databases}} \\ \hline
+        \textbf{Bingo}  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
+        \textbf{MyChem}  & http://XXX.sf.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
+        \textbf{OrChem} \cite{RijnbeekS10} & http://orchem.sf.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
+       \textbf{PGChem}  & http://XXX.sf.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
+       \multicolumn{3}{|c|}{\textbf{Other}} \\ \hline
+        \textbf{cclib}  & http://cclib.sf.net & Python library for parsing and interpreting computational chemistry results \\
+        \hline
+%% GaussSum, QMForge
        \end{tabular}
@@
     \mbox{
-\begin{tabular}{|l|l|l|}
+\begin{tabular}{|l|c|l|}
   %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-  \hline Name & License/Waiver & Description \\ \hline
-  ChemPedia & CC0 & Crowd-sourced chemical names. Project discontinued. \\ \hline
-  CrystalEye     & PPDL & Crystal structures from primary literature. \\ \hline
-  ONS Solubility & CC0 & Solubility data for various solvents. \\ \hline
+  \hline \textbf{Name} & \textbf{License/Waiver} & \textbf{Description} \\ \hline
+  \textbf{ChemPedia} & CC0 & Crowd-sourced chemical names. Project discontinued. \\ \hline
+  \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature. \\ \hline
+  \textbf{ONS Solubility} & CC0 & Solubility data for various solvents. \\ \hline
 \end{tabular}

### 2011-05-14 paper.tex
Fixed the order of affiliations (I hope). Two affiliations remain missing.

@@
 \address{%
-   \iid(1)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
-    \iid(2)National Institute of Health, Somewhere in the US, USA\\
-    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
-    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
-    \iid(7)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
-    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
     \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA\\
-   \iid(3)NIH, Somewhere in the US, USA\\
+    \iid(2)National Institute of Health, Somewhere in the US, USA \\
+    \iid(3)... \\
+    \iid(4)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(5)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
+    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(7)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
+    \iid(8)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
+    \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(10)... \\
+    \iid(11)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
     \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(13)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden,    The Netherlands \\
-    \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195
+    \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
 }%

### 2011-05-14 paper.tex
Added reference for NWChem.

@@
 chemistry community and a variety of programs exist, mostly
-proprietary software with the notable exception of NWChem.
+proprietary software with the notable exception of NWChem.\cite{NWChem}
 However, since each program uses different input formats, and the

### 2011-05-14 paper.tex
Various minor edits pertaining to cclib.

--- 2011-05-14 paper.tex
@@
 chemistry community and a variety of programs exist, mostly
-proprietary software with the notable exception of NWChem.
+proprietary software with the notable exception of NWChem.\cite{NWChem}
 However, since each program uses different input formats, and the
 results are output differently even varying between different versions
-of the software, preparing calculations and extracting the results in
-an automated way is problematic.
+of the software, preparing calculations and automatically extracting
+the results is problematic.
 
-Avogadro has already been mentioned as a GUI for preparing
-computational chemistry calculations. It uses Open Babel to read the
-output of several electronic structure packages, and also to write the
-input files for each. Jmol (see below) can also depict computational
-chemistry results including molecular orbitals.
+Avogadro has already been mentioned as a GUI for preparing calculations.
+It uses Open Babel to read the output of several electronic structure
+packages, and also to write the input files for each. Jmol (see below)
+can also depict computational chemistry results including molecular orbitals.
 
@@
 programs and presenting it in a standard way so that further analyses
-could be carried out independently of the quantum package used to
-obtain the results. cclib is a Python library, and
-the current version (version 1.0.1)
+could be carried out independently of the quantum package used.
+cclib is a Python library, and the current version (version 1.0.1)
 supports 8 different computational chemistry codes and extracts over

### 2011-05-15 paper.tex
Added my details and affiliation, reordered list.

@@
   \and
-  Geoffrey R Hutchison$^4$%
+  Marcus D Hanwell$^{4}$%
+  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
+  \and
+  Geoffrey R Hutchison$^5$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Karol M Langner$^5$%
+  Karol M Langner$^6$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
   \and
-  Daniel M Lowe$^6$%
+  Daniel M Lowe$^7$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
   \and
-  Peter Murray-Rust$^7$%
+  Peter Murray-Rust$^8$%
   \email{Peter Murray-Rust - pm286@cam.ac.uk}
   \and
-  Noel M O'Boyle\correspondingauthor$^8$%
+  Noel M O'Boyle\correspondingauthor$^9$%
   \email{Noel M O'Boyle\correspondingauthor - n.oboyle@ucc.ie}
   \and
-  Dmitry Pavlov$^9$%
+  Dmitry Pavlov$^{10}$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
@@
   \and
-  Christoph Steinbeck$^{10}$%
+  Christoph Steinbeck$^{11}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{11}$%
+  Adam L Tenderholt$^{12}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{12}$%
+  Kevin J Theisen$^{13}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
   \and
-  Egon L Willighagen$^{13}$%
+  Egon L Willighagen$^{14}$%
   \email{Egon L Willighagen - egon.willighagen@ki.se}
@@
     \iid(3)... \\
-    \iid(4)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(5)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
-    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(7)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
-    \iid(8)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
-    \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(10)... \\
-    \iid(11)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
-    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+    \iid(4)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(5)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(6)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
+    \iid(7)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(8)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
+    \iid(9)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
+    \iid(10)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(11)... \\
+    \iid(12)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
+    \iid(13)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
+    \iid(14)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden

### 2011-05-15 paper.tex
Added details about Kalzium to intro.
Kalzium was the first user of the Avogadro library, and was pivotal in
both providing one of the authors with early funding and pushing the
development of robust APIs in a second application visualizing chemical
structure.

@@
 microplate based high-throughput screening. Brunn provides a graphical interface
-for handling different plate layouts and dilution series and can automaticly
+for handling different plate layouts and dilution series and can automatically
 generate dose response curves and calculate IC$_{50}$-values. Avogadro
-is used by XtalOpt~\cite{WebXtalOpt}\cite{Lonie2011}, an evolutionary
+is used by Kalzium~\cite{WebKalzium}, a periodic table and chemical editor in KDE,
+and XtalOpt~\cite{WebXtalOpt}\cite{Lonie2011}, an evolutionary
 algorithm for crystal structure prediction. XtalOpt provides a

### 2011-05-15 paper.tex
Corrected some of the text on Avogadro and comp. chem.
Also, added some details about the recently split out OpenQube. Whilst
still in a very early stage, I think that it should also be considered a
Blue Obelisk project, and it is being actively worked on and was
incubated by the Avogadro project.

@@
 It uses Open Babel to read the output of several electronic structure
-packages, and also to write the input files for each. Jmol (see below)
+packages. Avogadro generates input files on the fly in response to user
+input on forms, as well as allowing inline editing of the files before
+they are saved to disk. It also features intuitive syntax highlighting
+for GAMESS input files, allowing expert users to easily spot mistakes
+before saving an input file to disk.
+
+In addition to this, significant development of new parsing routines took
+place in an Avogadro plugin to read in basis sets, and electronic structure
+output in order to calculate molecular orbital and electron density grids.
+This code was written to be parallel, using desktop shared memory parallelism
+and high level APIs in order to significantly speed up analysis. Most of this
+code was recently separated from the plugin, and released as a BSD licensed
+library, that is now used by the latest version of Avogadro, called OpenQube.
+Jmol (see below)
 can also depict computational chemistry results including molecular orbitals.

### 2011-05-15 paper.tex
Added my affiliation

--- 2011-05-15 paper.tex
@@
   \and
-  Marcus D Hanwell$^{4}$%
-  \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
-  \and
-  Geoffrey R Hutchison$^5$%
+  Geoffrey R Hutchison$^4$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Karol M Langner$^6$%
+  Karol M Langner$^5$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
   \and
-  Daniel M Lowe$^7$%
+  Daniel M Lowe$^6$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
   \and
-  Peter Murray-Rust$^8$%
+  Peter Murray-Rust$^7$%
   \email{Peter Murray-Rust - pm286@cam.ac.uk}
   \and
-  Noel M O'Boyle\correspondingauthor$^9$%
+  Noel M O'Boyle\correspondingauthor$^8$%
   \email{Noel M O'Boyle\correspondingauthor - n.oboyle@ucc.ie}
   \and
-  Dmitry Pavlov$^{10}$%
+  Dmitry Pavlov$^9$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
@@
   \and
-  Christoph Steinbeck$^{11}$%
+  Christoph Steinbeck$^{10}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{12}$%
+  Adam L Tenderholt$^{11}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{13}$%
+  Kevin J Theisen$^{12}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
   \and
-  Egon L Willighagen$^{14}$%
+  Egon L Willighagen$^{13}$%
   \email{Egon L Willighagen - egon.willighagen@ki.se}
@@
     \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(2)National Institute of Health, Somewhere in the US, USA \\
+    \iid(2)NIH Center for Translational Therapeutics, 9800 Medical  Center Drive, Rockville, MD 20878 \\
     \iid(3)... \\
-    \iid(4)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(5)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(6)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
-    \iid(7)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(8)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
-    \iid(9)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
-    \iid(10)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(11)... \\
-    \iid(12)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
-    \iid(13)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(14)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+    \iid(4)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(5)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
+    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(7)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
+    \iid(8)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
+    \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(10)... \\
+    \iid(11)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
+    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
+    \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden

### 2011-05-15 paper.tex
Added text on CDK devs and some associated refs

--- 2011-05-15 paper.tex
@@
     \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(2)National Institute of Health, Somewhere in the US, USA \\
+    \iid(2)NIH Center for Translational Therapeutics, 9800 Medical  Center Drive, Rockville, MD 20878 \\
     \iid(3)... \\
@@
 
-[Insert main focus of work on CDK in last 5 years]
+The CDK project has been under regular development over the last five
+years. Several feature have been implemented ranging from core
+components such as an extensible SMARTS matching system, new graph (\&
+subgraph) isomorphism method \cite{smsd} to more application oriented
+components such as 3D pharmacophore searching and matching, a variety
+of structural key and hashed fingerprints. In addition, there have
+been a number of second generation tools developed on top of the CDK
+such as OrChem \cite{orchem} and PaDEL \cite{padel}. In addition to
+the use of the CDK in various applications, it has been deployed in
+the form of web services \cite{Dong:2007aa} and has formed the basis
+of a variety of web applications.
 

### 2011-05-18 paper.tex
Expanded JNI-InChI section

@@
 for chemical substances designed to enable linking of diverse data
-compilations. Although its development predates the Blue Obelisk,
-software such as Open Babel has included InChI support since 2005,
-and support for InChI in Indigo is due in 2011.
+compilations. Prior to the development of the InChI identifier chemical
+information systems and databases used a wide variety of (generally
+proprietary) identifiers, greatly limiting their interoperability.
+Although its development predates the Blue Obelisk, software such as Open
+Babel has included InChI support since 2005, and support for InChI in
+Indigo is due in 2011.
 
@@
 access from the other widely used language for cheminformatics
-toolkits, Java. The Blue Obelisk project JNI-InChI~\cite{WebJNIInChI}
-was established in 2006 to
-solve this problem by using the Java Native Interface to link the
-InChI binary to Java. In this way, it promotes the wider adoption of
+toolkits, Java. Early attempts to generate InChI identifiers from
+within Java involved programatically launching the InChI executable
+and capturing the output, an approach that was found to be fairly
+unreliable and broke the 'write once, run anywhere' philosophy
+of Java.  The Blue Obelisk project JNI-InChI~\cite{WebJNIInChI}
+was established in 2006 to solve this problem by using the Java Native
+Interface framework to provide transparent access to the InChI
+library from within Java, and other Java Virtual Machine (JVM) based
+languages, supporting the wider adoption of
 this standard identifier by the chemistry community.
+
+The Java Native Interface framework provides a mechanism for code
+running inside the JVM to place calls to libraries written in languages
+such as C, C++ and Fortran, and compiled into native, machine specific,
+code. JNI-InChI provides a thin C wrapper, with corresponding Java code,
+around the IUPAC InChI library, exposing the InChI library's functionality
+to the JVM.  To overcome the need to have the correct InChI library pre-installed
+on a system, JNI-InChI comes with a variety of precompiled native binaries and
+automatically extracts and deploys the correct one for the detected operating
+system and architecture. The JNI-InChI library comes with native binaries
+supporting a range of operating systems and architectures; the current version
+has binaries for 32- and 64-bit Windows, Linux and Solaris, 64-bit FreeBSD
+and 64-bit Intel-based Mac OS X - a number of which are not supported by the
+original IUPAC distribution of InChI.
+
+The JNI-InChI project has matured to support the full range of
+functionality of the InChI C library: structure-to-InChI, InChI-to-structure,
+AuxInfo-to-structure, InChIKey generation, and InChI and InChIKey validation.
+JNI-InChI provides the InChI functionality for a number of Open Source projects,
+including the Chemistry Development Kit, Bioclipse and CMLXOM/JUMBO, and is
+also used by commercial applications and internally in a number of companies.
+Through its widespread use and Open Source development model, a number of
+issues in earlier versions of the software have been identified and resolved,
+and JNI-InChI now offers a robust, and defacto standard tool for working with
+InChIs in the JVM.
 

### 2011-05-23 paper.tex
Added Sam Adams to author list; Fixed Daniel Lowe's affiliation
spelling (s/scafold/scaffold/)

--- 2011-05-15 paper.tex
@@
 Sieve for filtering molecular structure by molecular property,
-Stripper for removing core scafold structures from a molecule
+Stripper for removing core scaffold structures from a molecule
 set, and Piramid for molecular alignment using shape determined

### 2011-05-27 paper.tex
Corrected capitalization of iChemLabs.

--- 2011-05-18 paper.tex
@@
 \author{
+  Samuel E Adams$^8$%
+  \email{Samuel E Adams - sea36@cam.ac.uk}
+  \and
   Jonathan Alvarsson$^1$%
@@
   \and
-  Daniel M Lowe$^7$%
+  Daniel M Lowe$^8$%
  \email{Daniel Lowe - dl387@cam.ac.uk}
@@
 solutions around Open Source cheminformatics components provided by the Blue Obelisk
-community. Examples of such companies include IChemLabs, IdeaConsult, Wingu, Silicos,
+community. Examples of such companies include iChemLabs, IdeaConsult, Wingu, Silicos,
 GenettaSoft, hBar, and Inkspot Science. Some of these merely use components, but several

### 2011-05-29 paper.tex
Added sections from PMR

@@
 %%
+
+\section*{Scope}
+The Blue Obelisk covers many areas of chemistry and chemical resources
+used by neighbouring disciplines ({\it e.g.} biochemistry, materials
+science). Many of the efforts relate to cheminformatics (the scope of
+this journal) and we believe that many of the publications in {\it J.
+Cheminf.} could be completely carried out using Blue Obelisk resources
+and other OS chemical tools. The importance of this is that for the
+first time it would allow reviewers, editors and readers to validate
+assertions in the journal and also to re-run and re-analyse parts of
+the calculation.
+
+However, Blue Obelisk software and data is used outside
+cheminformatics and certainly in the five main areas that, for
+example, chemical markup language (CML) supports:
+
+\begin{itemize}
+\item{1.}{\bf Molecules:} This is probably the largest area for Blue
+Obelisk software and data, and is reflected by many programs that
+visualise, transform, convert formats and calculate properties. It is
+almost certain that any file format in use can be processed by Blue
+Obelisk software and properties can be calculated for most (organic
+compounds).
+\item{2.}{\bf Reactions:} Blue Obelisk software can describe the
+semantics of reactions and provide atom-atom matching and analyse
+stoichiometric balance in reactions.
+\item{3.}{\bf Computational chemistry:} Blue Obelisk software can
+interpret many of the current output files from calculations and
+create input for jobs. The Quixote project (see later and elsewhere in
+this issue) shows that Open Source approaches based on Blue Obelisk
+resources and principles are increasing the availability and
+re-usability of computational chemistry.
+\item{4.}{\bf Spectra:} 1-D spectra (NMR, IR, UV etc.) are fully
+supported in Blue Obelisk offerings for conversion and display. There
+is a limited amount of spectral analysis but the software gives a
+platform on which it should be straightforward to develop spectral
+annotation and manipulation. Currently the Blue Obelisk does not
+support much in multi-dimensional NMR and multi-equipment spectra
+({\it e.g.} GC-MS).
+\item{5.}{\bf Crystallography:} The Blue Obelisk software supports the
+bi-directional processing of crystal structure files (CIF) and also
+solid-state calculations such as plane-waves with periodic boundary
+conditions. There is considerable support for the visualisation of
+both periodic and aperiodic condensed objects.
+\end{itemize}
+
+Many of the current operations in installing and running chemical
+computations and using the data are integration and customisation
+rather than fundamental algorithms. It is very difficult to create
+universal platforms that can be distributed and run by a wide range of
+different users, and in general, the Blue Obelisk deliberately does
+not address these. Our approach is to produce components that can be
+embedded in chemists. environments, including databases and workflows.
+We believe that a chemical laboratory with reasonable access to common
+software engineering should be able to build customised applications
+using Blue Obelisk components and standard infrastructure such as
+workflows (KNIME or Taverna) and databases (MySQL, CouchDB etc.).
+Where the Blue Obelisk itself produces data resources (NMRShiftDB,
+CrystalEye, Quixote etc.) they are normally done with Open components
+so that the community can, if necessary, replicate them but not
+necessarily without additional effort.
+
+Much of the impetus behind Blue Obelisk software is to create an
+environment for chemical computation (including cheminformatics) where
+all of the components, data, specifications, semantics, ontology and
+software are Openly visible and discussable. The largest current uses
+by the general chemical community are in authoring, visualisation and
+cheminformatics calculations but we anticipate that this will shortly
+extend into mainstream computational chemistry and solid-state.
+Although many of the authors are employed as research scientists,
+there are also several people who contribute in their spare time and
+we anticipate an increasing value and use of the Blue Obelisk in
+education at all levels.
+
 \section*{Open Source}
@@
 \section*{Open Standards}
+
+\subsection*{Chemical Markup Language, CML}
+
+Chemical Markup Language (CML) is discussed in several places in this
+issue, and a brief summary here re-iterates that it is designed
+primarily to create a validatable semantic representation for chemical
+objects. The five main areas (molecules, reactions, computational
+chemistry, spectra and solid-state (see above)) have now all been
+extensively deployed and tested. CML can therefore be used as a
+reference for input and output for Blue Obelisk software and a means
+of representing data in Blue Obelisk resources.
+
+CML, being an XML application, can inter-operate with other markup
+languages and in particular XHTML, SVG, MathML, docx and more
+specialised applications such as UnitsML and GML (geosciences). We
+believe that it would be possible using these languages to encode
+large parts of, say, first year chemistry text books in XML.
+Similarly, it is possible to create compound documents with word
+processing or spreadsheet software that has inter-operating text,
+graphics and chemistry (as in Chem4Word). Being a markup language, CML
+is designed for re-purposing, including styling, and therefore a
+mixture of these languages can be used for chemical catalogues,
+general publications, logbooks and many other types of document in the
+scientific process.
+
+CML describes much of its semantics through conventions and
+dictionaries, and the emerging ecosystem (especially in computational
+chemistry) is available as a semantic resource for many of the
+applications and specifications in this article.
+
   \subsection*{InChI}
@@
 OpenSMILES, especially polymers.
-
-\subsection*{CML}
-?
 

### 2011-05-31 paper.tex
Update to table1 + bib from PMR

@@
        \hline \textbf{Name} & \textbf{Website} & \textbf{Description} \\ \hline
+        \multicolumn{3}{|c|}{\textbf{CML Tools}} \\ \hline
+		\textbf{JUMBO & CMLXOM} & http://sourceforge.net/projects/cml/ https://bitbucket.org/wwmm/cmlxom/ & Peter Murray-Rust et al. \\ \hline
 		\multicolumn{3}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
@@
         \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
+		\textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ & Sam Adams et al. \\ \hline
         \textbf{A3} & ..  & .  \\ \hline
@@
         \textbf{OPSIN} & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
+		\textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ & Peter Murray-Rust et al. \\ \hline
+		\textbf{CIFXOM} & https://bitbucket.org/wwmm/cifxom/ & Peter Murray-Rust et al. \cite{DayEtAl2011} \\ \hline
         \textbf{A3} & ..  & .  \\ \hline
@@
         \textbf{PGChem}  & http://XXX.sf.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
+ 	    \textbf{Chempound (Chem#)} & https://bitbucket.org/chempound & Sam Adams et al. \\ \hline
+	   \multicolumn{3}{|c|}{\textbf{Text mining}} \\ \hline
+	    \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ & Peter Murray-Rust et al. \\ \hline
+		\textbf{ChemicalTagger} & http://chemicaltagger.ch.cam.ac.uk/ & Lezan Hawizy et al. \cite{HawizyEtAl2011} \\ \hline
        \multicolumn{3}{|c|}{\textbf{Other}} \\ \hline
@@
   \textbf{ONS Solubility} & CC0 & Solubility data for various solvents. \\ \hline
+  \textbf{NMRShiftDB} & ? & Database of NMR spectra of organic compounds. \\ \hline
 \end{tabular}

### 2011-05-31 paper.tex
Restored unix line endings
Fix compile issues

@@
         \multicolumn{3}{|c|}{\textbf{CML Tools}} \\ \hline
-		\textbf{JUMBO & CMLXOM} & http://sourceforge.net/projects/cml/ https://bitbucket.org/wwmm/cmlxom/ & Peter Murray-Rust et al. \\ \hline
+		\textbf{JUMBO \& CMLXOM} & http://sourceforge.net/projects/cml/ https://bitbucket.org/wwmm/cmlxom/ & Peter Murray-Rust et al. \\ \hline
 		\multicolumn{3}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
@@
         \textbf{PGChem}  & http://XXX.sf.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
- 	    \textbf{Chempound (Chem#)} & https://bitbucket.org/chempound & Sam Adams et al. \\ \hline
+ 	    \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound & Sam Adams et al. \\ \hline
 	   \multicolumn{3}{|c|}{\textbf{Text mining}} \\ \hline

### 2011-05-31 paper.tex
More updates from PMR

@@
 
-[Few lines on Quixote here]
+The Quixote project epitomises the full use of the Blue Obelisk
+software and is described in detail in a sibling
+article. Here we observe that it is possible to convert legacy files
+of all sorts into semantic chemistry and extract
+those parts which are suitable for input to computational chemistry
+programs. This chemistry is then combined with
+generic concepts of computational chemistry ({\it e.g.} strategy,
+machine resources, timing, accuracy etc.) into the
+legacy inputs for a wide range of programs. Quixote itself follows
+Blue Obelisk principles in that it does not manage
+the submission and monitoring of jobs but resumes action when the jobs
+have been completed, and then applies a range
+of parsing and transformation tools to create standardised semantic
+chemical content. A major feature of Quixote is
+that it requires all concepts to validate against dictionaries and the
+process of parsing files necessarily generates
+communally-agreed dictionaries, which represent an important step
+forward in the Open specifications for Blue Obelisk.
+When widely-deployed, Quixote will advertise the value of Open
+community standards for semantics to the world.
+
+The Quixote project is not dependent on any particular technology,
+other than the representation of computational
+chemistry in CML and the management of semantics through CML
+dictionaries. At present, we use JUMBO-Converters for most
+ of the semantic conversion, Lensfield2 for the workflow and Chempound
+(chem\#) to store and disseminate the results.
 
@@
 for an elaborate list.
+
+In May 2011, the EBI ran an industry-oriented workshop (MIOSS -
+Molecular Informatics Open Source Software). This explored the role of
+Open Source in industrial laboratories and companies, and several of
+the presenters are among the authors of this paper.
+The meeting identified that Open Source was extremely valuable to
+industry, not just because it is 'free as in beer'
+but because it allows the validation of source code, data and
+computational procedures. A phrase from the meeting
+summed it up: "The ice is beginning to melt", signifiying that we can
+expect a rapid increase in industry's interest
+in Open Source.
+
+Some of the discussion was on business models. There are difficulties
+for software for which there is no formal
+transaction, other than downloasding and agreeing to license terms.
+COmpanies are concerened about training and support
+and in some cases, product liability. One anecdote was of a company
+which wished to donate money to an Open Source
+project but could not find a mechanism to do so.
+
+There is a considerable amount of contribution-in-kind, both from
+enhancements to software and also completely new
+systems and toolkits, and companies are finding it easier to create
+mechanisms for releasing Open Source software
+without violating confidentiality or incurring liability.
 
@@
 
-\subsection*{Chemical Databases}
+\subsection*{Chemical database software}
 
@@
 
+Besides the traditional and proven relational databse approach, with
+added chemical features ('cartridges'), there is growing
+interest in tools and approaches based on the web philosophy and
+practice. Several groups are experimenting with RDF
+on the assumption that generic high-performance solutions are
+appearing. RDF allows everything to be described by
+URIs (data, molecules, dictionaries, relations). The Chempound system,
+as deployed in
+Quixote and elsewhere, is an RDF-based approach to chemical structures
+and compounds and their properties. For small
+to medium-sized collections (such as an individual's calculations or
+literature retrieval), there are many RDF tools
+(e.g. SIMILE, Apache Jena) which can operate in machine memory and provide
+the flexibility that RDF offers. For larger
+systems, it is unclear whtether complete RDF solutions (e.g. Virtuoso)
+will be satisfactory or whether a hybrid system
+based on name-value pairs (e.g. CouchDB, MongoDB) will be sufficient.
+
   \subsection*{Collaboration and interoperability}
@@
 available from KNIME's ``Community Updates'' site.
-
-  \subsection*{Remaining challenges}
-
-(Say something here about benchmarks to measure accuracy. Clear examples
- of performance on open datasets are required. Otherwise there is
- nothing to counter anecdotal evidence or FUD spread by others.)
-
-(Better engagement with industry...make it clear what and how industry
- members can engage with projects.)
 
@@
 substantial knowledge base.
+
+\subsection*{The Blue Obelisk Data Repository, BODR;
+http://sourceforge.net/projects/bodr/}
+
+From the beginning, the Blue Obelisk created a repository of key data,
+particularly that which would be used in
+algorithms and where there was a need to ensure that the values were
+standard between codes. Examples are atomic
+masses and conversion between physical constants. In principle, this
+material can be copied between sites such as IUPAC and
+wikipedia and continued practice in this area should lead to a
+enhancement in the quality of community reference data.
+
+\subsection*{NMRShiftDB}
+
+NMRShiftDB represents one of the earliest (2002?) resources for Open
+community-contributed data. Groups which
+measure NMR spectra or extract it from the literature contribute to
+NMRShiftDB which provides an Open resource where
+entries can be searched by chemical structure or properties
+(especially peaks). It is, not surprisingly, difficult to
+extract large amounts of altruistic contribution (as happens in
+wikipedia) but it is increasingly possible to link
+data capture with data publication. For example, the Blue Obelisk has
+enough software that it is possible to create
+a seamless chain for converting NMR structures in-house into
+NMRShiftDB entries. If and when the chemistry community
+encourages or requires semantic publication of spectra, rather than
+PDFs, it would be possible to populate NMRShiftDB rapidaly
+along the the lines of CrystalEye.
+
+\subsection*{CrystalEye}
+
+CrystalEye represents the cost-effective extraction of data from the
+literature where this is published both
+Openly and semantically. The cost is extremely low and software can be
+run every night, updating the entries
+(currently at about 250,000 structures). CrystalEye serves as a model
+for a high-value, high-quality Open data
+resource, including the licensing of each component as
+Panton-compatible Open data.
 

### 2011-06-02 paper.tex
Half way through proofread

@@
         \paragraph*{Background:} The Blue Obelisk movement was established in 2005 as a
-		response to the lack of open data, open standards and
-		open source (ODOSOS) in chemistry. While other scientific disciplines
-		such as physics, biology and astronomy (to name a few) were embracing
-		new ways of doing science and reaping the benefits of community
-		efforts, there was little if any innovation in the field of chemistry
-		and scientific progress was actively hampered by the lack of access to
-		data and tools.
+		response to the lack of Open Data, Open Standards and
+		Open Source (ODOSOS) in chemistry. It aims to
+make it easier to carry out chemistry research by
+promoting interoperability between chemistry software, encouraging cooperation
+between Open Source developers, and developing community resources and
+Open Standards.
 
@@
 
-        \paragraph*{Conclusions:} Here we show that the Blue Obelisk has been very successful
+        \paragraph*{Conclusions:} We show that the Blue Obelisk has been very successful
 		in bringing together researchers and developers with common interests
@@
 229\textsuperscript{th} National Meeting of the American Chemistry
-Society as a response to the lack of open data, open standards and
-open source (ODOSOS) in chemistry. While other scientific disciplines
+Society as a response to the lack of Open Data, Open Standards and
+Open Source (ODOSOS) in chemistry. While other scientific disciplines
 such as physics, biology and astronomy (to name a few) were embracing
@@
 Cheminf.} could be completely carried out using Blue Obelisk resources
-and other OS chemical tools. The importance of this is that for the
+and other Open Source chemical tools. The importance of this is that for the
 first time it would allow reviewers, editors and readers to validate
@@
 cheminformatics and certainly in the five main areas that, for
-example, chemical markup language (CML) supports:
+example, Chemical Markup Language (CML) \cite{murray-rust_chemical_1999} supports:
 
@@
 visualise, transform, convert formats and calculate properties. It is
-almost certain that any file format in use can be processed by Blue
-Obelisk software and properties can be calculated for most (organic
+almost certain that any file format currently in use can be processed by Blue
+Obelisk software and that properties can be calculated for most (organic
 compounds).
@@
 interpret many of the current output files from calculations and
-create input for jobs. The Quixote project (see later and elsewhere in
+create input for jobs. The Quixote project (see below and elsewhere in
 this issue) shows that Open Source approaches based on Blue Obelisk
@@
 platform on which it should be straightforward to develop spectral
-annotation and manipulation. Currently the Blue Obelisk does not
-support much in multi-dimensional NMR and multi-equipment spectra
+annotation and manipulation. However, currently the Blue Obelisk lacks
+support for multi-dimensional NMR and multi-equipment spectra
 ({\it e.g.} GC-MS).
@@
 not address these. Our approach is to produce components that can be
-embedded in chemists. environments, including databases and workflows.
+embedded in many environments, from stand-alone applications to web
+applications, databases and workflows.
 We believe that a chemical laboratory with reasonable access to common
-software engineering should be able to build customised applications
+software engineering techniques should be able to build customised applications
 using Blue Obelisk components and standard infrastructure such as
@@
 CrystalEye, Quixote etc.) they are normally done with Open components
-so that the community can, if necessary, replicate them but not
-necessarily without additional effort.
+so that the community can, if necessary, replicate them.
 
@@
 were the Chemistry Development Kit (CDK)
-\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpenBabel},
+\cite{Steinbeck2003, Steinbeck2006}, Open Babel \cite{WebOpen Babel},
 and JOELib \cite{WebJOELib}. Of these, both the CDK and Open Babel
@@
 years. Several feature have been implemented ranging from core
-components such as an extensible SMARTS matching system, new graph (\&
-subgraph) isomorphism method \cite{smsd} to more application oriented
-components such as 3D pharmacophore searching and matching, a variety
-of structural key and hashed fingerprints. In addition, there have
+components such as an extensible SMARTS matching system and a new graph
+(and
+subgraph) isomorphism method \cite{smsd}, to more application oriented
+components such as 3D pharmacophore searching and matching, and a variety
+of structural-key and hashed fingerprints. In addition, there have
 been a number of second generation tools developed on top of the CDK
-such as OrChem \cite{orchem} and PaDEL \cite{padel}. In addition to
-the use of the CDK in various applications, it has been deployed in
+(see below). As well as the use of the CDK in various tools, it has been deployed in
 the form of web services \cite{Dong:2007aa} and has formed the basis
@@
 bindings, as well as performance improvements for its database
-cartridge. More recently, GGA Software Services
+cartridge.
+
+More recently, GGA Software Services
 (a contract programming company)
@@
 
-A final example of 2nd-generation Blue Obelisk software is the 
+A final example of second-generation Blue Obelisk software is the 
 AMBIT2~\cite{WebAMBIT} software, a GUI that facilitates registration
@@
 chemistry community and a variety of programs exist, mostly
-proprietary software with the notable exception of NWChem.\cite{NWChem}
+proprietary software with the notable exception of
+NWChem~\cite{NWChem}.
 However, since each program uses different input formats, and the
-results are output differently even varying between different versions
-of the software, preparing calculations and automatically extracting
+the output formats vary widely (sometimes even varying between different versions
+of the same software), preparing calculations and automatically extracting
 the results is problematic.
@@
 In addition to this, significant development of new parsing routines took
-place in an Avogadro plugin to read in basis sets, and electronic structure
+place in an Avogadro plugin to read in basis sets and electronic structure
 output in order to calculate molecular orbital and electron density grids.
@@
 code was recently separated from the plugin, and released as a BSD licensed
-library, that is now used by the latest version of Avogadro, called OpenQube.
+library, OpenQube, which is now used by the latest version of
+Avogadro.
 Jmol (see below)
@@
 on user-defined molecular fragments. QMForge also provides a rudamentary
-Cartesian coordinate editor allowing molecular structures to be saved via OpenBabel.
+Cartesian coordinate editor allowing molecular structures to be saved via Open Babel.
 
@@
 software and is described in detail in a sibling
-article. Here we observe that it is possible to convert legacy files
+article~\cite{QuixoteSameIssue}. Here we observe that it is possible
+to convert legacy chemistry file formats
 of all sorts into semantic chemistry and extract
@@
 software and method development, which becomes visible when proprietary companies
-merge. Some authors have rationalized that in the chemistry field, the
+merge. Some authors have rationalised that in the chemistry field, the
 data has value, and few contributors are available to volunteer time
@@
 Silicos has release several Open Source utilities based on
-OpenBabel, such as Pharao, a tool for pharmacophore searching,
+Open Babel, such as Pharao, a tool for pharmacophore searching,
 Sieve for filtering molecular structure by molecular property,
@@
 by the Gaussian volumes as a descriptor. Additionally,
-contributions have been made to the OpenBabel project itself.
+contributions have been made to the Open Babel project itself.
 
 Other companies use Blue Obelisk components and contribute patches,
-smaller and larger. For example, IXELES donated the isomorphism
+smaller and larger. For example, IXELIS donated the isomorphism
 code in the CDK, eMolecules donated canonicalisation code to
-OpenBabel, and AstraZeneca contributed code to the CDK for
+Open Babel, and AstraZeneca contributed code to the CDK for
 signatures. This is just a very minor selection, and the author
@@
 conversion and molecular property calculations, GraphicsMagick for image
-manipulation, Potrace for vectorization, GOCR and OCRAD for optical
+manipulation, Potrace for vectorisation, GOCR and OCRAD for optical
 character recognition. The growing importance of image
@@
 A number of structure registration systems have been published in the last five years, exploiting the fact that
-free cheminformatics toolkits such as OpenBabel and the CDK were available.
+free cheminformatics toolkits such as Open Babel and the CDK were available.
 OrChem,\cite{WebOrChem} for example, is an open source extension for the Oracle 11G database that
@@
 on Jmol~\cite{JmolBook}, the Chemistry Development Kit~\cite{CDKBook}
-and Open Babel~\cite{OpenBabelBook}.
+and Open Babel~\cite{Open BabelBook}.
 

### 2011-06-03 paper.tex
Proofreading

--- 2011-05-31 paper.tex
@@
 chemistry in CML and the management of semantics through CML
-dictionaries. At present, we use JUMBO-Converters for most
- of the semantic conversion, Lensfield2 for the workflow and Chempound
-(chem\#) to store and disseminate the results.
+dictionaries. At present, we use JUMBO-Converters~\cite{JUMBO-Convertors} for most
+ of the semantic conversion, Lensfield2~\cite{Lensfield2} for the workflow and Chempound
+(chem\#) \cite{Chempound} to store and disseminate the results.
 
@@
 cheminformatics community. Traditional business models rely on monetisation of source
-code, causing every company to repeat the development by other companies, often combined
+code, causing companies to repeat work done by other companies. This
+model is sometimes combined
 with a free (gratis) model for people working at academic institutes, to increase adoption
-and allow contributions from academics. This solution defines the IP on the software
+and encourage contributions from academics. This solution defines the IP on the software
 return on investment, but has the downside of investment losses due to duplication of
 software and method development, which becomes visible when proprietary companies
-merge. Some authors have rationalized that in the chemistry field, the
-data has value, and few contributors are available to volunteer time
-to improve codes.\cite{Stahl:2005fk} If true, this would hamper
+merge. Some authors have argued hat in the chemistry field
+few contributors are available to volunteer time
+to improve codes and IP considerations may prevent contributions from
+industry~\cite{Stahl:2005fk}. If true, this would hamper
 adoption of Open Source and Open Data in chemistry, and greatly slow
@@
 
-The Blue Obelisk community, however, takes advantage that much of the investments needed
-for development are either payed by academic institutes and funding schemes, and by
-volunteers investing time and effort who get full access to the source code in return.
-In fact, all partners get full access up front and the Open Source licending ensures
-that they will have access any time in the future. As such, it functions as a social
+The Blue Obelisk community, however, takes advantage of the fact that much of the investment needed
+for development is either paid for by academic institutes and funding
+schemes, or by
+volunteers investing time and effort. In return, contributors get full
+access to the source code, and the Open Source licensing ensures
+that they will have access any time in the future. In this way, the
+license functions as a social
 contract between everyone to arrange the immediate return on investment. Effectively,
@@
 software from scratch, allowing researchers and commercial partners alike to focus
-on their core business, rather than the development of prerequisites. As such, the rich
-collection of Open Source cheminformatics tools provided by the Blue Obelisk
+on their core business, rather than the development of prerequisites.
+In the case of the Blue Obelisk, the rich
+collection of Open Source cheminformatics tools provided
 greatly reduces investment up front for new companies in the
 cheminformatics market. Such advantages have also been noted in the
-drug discovery field.\cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}
+drug discovery
+field \cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}.
 
@@
 interfaces for chemical content. The project is already being adopted by others,
-including iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by
-Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.
+including iBabel~\cite{iBabel}, ChemSpotlight~\cite{chemspotlight} and the RSC ChemSpider~\cite{chemspider_chemdoodle}.
 

### 2011-06-03 paper.tex

--- 2011-06-02 paper.tex
@@
 cheminformatics community. Traditional business models rely on monetisation of source
-code, causing every company to repeat the development by other companies, often combined
+code, causing companies to repeat work done by other companies. This
+model is sometimes combined
 with a free (gratis) model for people working at academic institutes, to increase adoption
-and allow contributions from academics. This solution defines the IP on the software
+and encourage contributions from academics. This solution defines the IP on the software
 return on investment, but has the downside of investment losses due to duplication of
 software and method development, which becomes visible when proprietary companies
-merge. Some authors have rationalised that in the chemistry field, the
-data has value, and few contributors are available to volunteer time
-to improve codes.\cite{Stahl:2005fk} If true, this would hamper
+merge. Some authors have argued that in the chemistry field
+few contributors are available to volunteer time
+to improve codes and IP considerations may prevent contributions from
+industry~\cite{Stahl:2005fk}. If true, this would hamper
 adoption of Open Source and Open Data in chemistry, and greatly slow

### 2011-06-03 paper.tex
Proofreading

@@
 with a free (gratis) model for people working at academic institutes, to increase adoption
-and encourage contributions from academics. This solution defines the IP on the software
-return on investment, but has the downside of investment losses due to duplication of
-software and method development, which becomes visible when proprietary companies
+and encourage contributions from academics. This solution defines the return on investment as the IP on the software,
+but has the downside of investment losses due to duplication of
+software and method development, which become visible when proprietary companies
 merge. Some authors have argued that in the chemistry field
@@
 license functions as a social
-contract between everyone to arrange the immediate return on investment. Effectively,
+contract between everyone to arrange an immediate return on investment. Effectively,
 this approach shares the burden of the high investment in having to develop cheminformatics
@@
 is encouraged to contact the individual Blue Obelisk projects
-for an elaborate list.
-
-In May 2011, the EBI ran an industry-oriented workshop (MIOSS -
-Molecular Informatics Open Source Software). This explored the role of
-Open Source in industrial laboratories and companies, and several of
-the presenters are among the authors of this paper.
-The meeting identified that Open Source was extremely valuable to
-industry, not just because it is 'free as in beer'
+for a detailed list.
+
+In May 2011, a Wellcome Trust Workshop on
+Molecular Informatics Open Source Software (MIOSS) explored the role of
+Open Source in industrial laboratories and companies as well as
+academia (several of
+the presenters are among the authors of this paper).
+The meeting identified that Open Source software was extremely valuable to
+industry not just because it is available for free,
 but because it allows the validation of source code, data and
-computational procedures. A phrase from the meeting
-summed it up: "The ice is beginning to melt", signifiying that we can
+computational procedures. Some 
+of the discussion was on business models or other ways to maintain 
+development of Open Source software on which a business relied.
+Companies are concerned about training and support
+and, in some cases, product liability.  There are difficulties
+for software for which there is no formal
+transaction other than downloading and agreeing to license terms.
+One anecdote concerned a company
+that wished to donate money to an Open Source
+project but could not find a mechanism to do so.
+
+Industry participants also pointed out that 
+there is a considerable amount of
+contribution-in-kind from industry, both from
+enhancements to software and also the development of completely new
+software and toolkits. Companies are now finding it easier to create
+mechanisms for releasing Open Source software
+without violating confidentiality or incurring liability.
+A phrase from the meeting
+summed it up: ``The ice is beginning to melt'', signifiying that we can
 expect a rapid increase in industry's interest
 in Open Source.
-
-Some of the discussion was on business models. There are difficulties
-for software for which there is no formal
-transaction, other than downloasding and agreeing to license terms.
-COmpanies are concerened about training and support
-and in some cases, product liability. One anecdote was of a company
-which wished to donate money to an Open Source
-project but could not find a mechanism to do so.
-
-There is a considerable amount of contribution-in-kind, both from
-enhancements to software and also completely new
-systems and toolkits, and companies are finding it easier to create
-mechanisms for releasing Open Source software
-without violating confidentiality or incurring liability.
 
@@
 A number of structure registration systems have been published in the last five years, exploiting the fact that
-free cheminformatics toolkits such as Open Babel and the CDK were available.
-OrChem,\cite{WebOrChem} for example, is an open source extension for the Oracle 11G database that
+Open Source cheminformatics toolkits such as Open Babel and the CDK
+are available.
+OrChem~\cite{RijnbeekS10}, for example, is an open source extension for the Oracle 11G database that
 adds registration and indexing of chemical structures to support fast
@@
 
-Besides the traditional and proven relational databse approach, with
-added chemical features ('cartridges'), there is growing
+Besides the traditional and proven relational database approach with
+added chemical features (`cartridges'), there is growing
 interest in tools and approaches based on the web philosophy and
-practice. Several groups are experimenting with RDF
-on the assumption that generic high-performance solutions are
-appearing. RDF allows everything to be described by
-URIs (data, molecules, dictionaries, relations). The Chempound system,
+practice. Several groups are experimenting with 
+the Resource Description Framework (RDF) language
+on the assumption that generic high-performance solutions will appear.
+RDF allows everything to be described by
+URIs (data, molecules, dictionaries, relations). The Chempound
+system~\cite{Chempound},
 as deployed in
@@
 
-One of the effects of the Blue Obelisk has been to bring developers
+One of the successes of the Blue Obelisk has been to bring developers
 together from different Open Source chemistry projects so that they
@@
 were looking to add support for energy minimisation, rather than
-implement a forcefield from scratch they ported the UFF forcefield\cite{Rappe:1992um}
-implementation from Open Babel to Jmol. This code has allowed Jmol to
+implement a forcefield from scratch they ported the UFF forcefield~\cite{Rappe:1992um}
+implementation from Open Babel to Jmol. This code enables Jmol to
 support 2D to 3D conversion of structures (through energy
-minimisation). Similarly, efficient Jmol code for atom-atom rebonding
-has been ported to the CDK.
+minimisation). In a similar manner, efficient Jmol code for atom-atom rebonding
+has been ported to the CDK. Figure 2 shows the collaborative nature of
+software developed in the Blue Obelisk, as one project builds on
+functionality provided by another project.
 
@@
 the ChemiSQL project. This brought together the developers of several
-open source chemistry database cartridges (PgChem,\cite{WebPgChem} MyChem,\cite{WebMyChem} OrChem\cite{WebOrChem} and
-more recently Bingo\cite{WebBingo}) with a view to making their database APIs more
+open source chemistry database cartridges (PgChem~\cite{WebPgChem},
+MyChem~\cite{WebMyChem}, OrChem~\cite{RijnbeekS10} and
+more recently Bingo~\cite{WebBingo}) with a view to making their database APIs more
 similar and collaborating on benchmark datasets for assessing
@@
 toolkits \cite{OBoyleCinfony2008}.
-It is a Python library that enables Open Babel, the CDK, and RDKit to
+It is a Python library that enables Open Babel, the CDK, and RDKit
+(and shortly, Indigo and OPSIN) to
 be used using the same API; this makes it easy, for example, to read a
@@
 
-Chemical Markup Language (CML) is discussed in several places in this
+Chemical Markup Language (CML) is discussed in several articles in this
 issue, and a brief summary here re-iterates that it is designed
@@
 Similarly, it is possible to create compound documents with word
-processing or spreadsheet software that has inter-operating text,
+processing or spreadsheet software that have inter-operating text,
 graphics and chemistry (as in Chem4Word). Being a markup language, CML
@@
 Interface framework to provide transparent access to the InChI
-library from within Java, and other Java Virtual Machine (JVM) based
+library from within Java and other Java Virtual Machine (JVM) based
 languages, supporting the wider adoption of
@@
 The Java Native Interface framework provides a mechanism for code
-running inside the JVM to place calls to libraries written in languages
+running inside the JVM, to place calls to libraries written in languages
 such as C, C++ and Fortran, and compiled into native, machine specific,
@@
 issues in earlier versions of the software have been identified and resolved,
-and JNI-InChI now offers a robust, and defacto standard tool for working with
+and JNI-InChI now offers a robust tool for working with
 InChIs in the JVM.
@@
 One of the most widely used ways to store chemical structures is the
-SMILES format (or SMILES string). This is a linear notation depicted
+SMILES format (or SMILES string). This is a linear notation developed
 by Daylight Information Systems that describes the connection table
@@
 convenient to manipulate (e.g. to include in spreadsheets, or copy
-from a Wikipedia article).
+from a web page).
 
@@
 
-Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
-SMILES notation, which allows to define crystals structures, polymers,
-electron delocalisation charges, molecule interactions, and many other
-features absent in the initial SMILES specification. There has been
-discussions about including parts of the CurlySMILES notation into
-OpenSMILES, especially polymers.
-
 \subsection*{QSAR-ML}
-The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses.
-QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}.
-A dataset in QSAR-ML includes the chemical structures (preferably described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation.
-Hence, a dataset described in QSAR-ML is completely reproducible.
-To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations.
-Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.
+The field of QSAR has long been hampered by the lack of open
+standards, which makes it difficult to share and reproduce descriptor
+calculations and analyses. QSAR-ML was recently proposed as an open
+standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset
+in QSAR-ML includes the chemical structures (preferably described in
+CML) with InChI to protect integrity, chemical descriptors linked to
+the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values,
+units, and versioned descriptor implementations to allow for
+integrating several descriptor software in the same calculation.
+Hence, a dataset described in QSAR-ML is completely reproducible. To
+allow for easy setup of QSAR-ML compliant datasets, a plugin for
+Bioclipse was created with a graphical interface for
+setting up QSAR datasets and performing calculations. Descriptor
+implementations are available from the CDK and JOELib, as well
+as via remote web services such as XMPP~\cite{Wagener:2009uq}.
 
@@
 process. While some specifications have been published (e.g., PubChem)
-and some standardisation services and tools are available (PubChem
-provides an online service to standardise molecules and the NCGC
-provides a stand alone tool) each group has their own set of rules. A
+and some standardisation services and tools are available (for
+example, PubChem
+provides an online service to standardise
+molecules~\cite{WebPubChemStandardizer})
+each group has their own set of rules. A
 common reference specification for standardisation would be of immense
@@
 exceptions: ChemPedia (a now discontinued crowd-sourcing project),
-CrystalEye,\cite{WebCrystalEye}
+CrystalEye~\cite{WebCrystalEye},
 and the Open Notebook Science Solubility
@@
 
-\subsection*{The Blue Obelisk Data Repository, BODR;
-http://sourceforge.net/projects/bodr/}
-
-From the beginning, the Blue Obelisk created a repository of key data,
-particularly that which would be used in
-algorithms and where there was a need to ensure that the values were
-standard between codes. Examples are atomic
-masses and conversion between physical constants. In principle, this
-material can be copied between sites such as IUPAC and
-wikipedia and continued practice in this area should lead to a
-enhancement in the quality of community reference data.
+\subsection*{The Blue Obelisk Data Repository (BODR)}
+
+The Blue Obelisk has created a repository of key chemical
+data in a machine-readable format~\cite{BODR}.
+The BODR focuses on data that is commonly required for
+chemistry software, and where there is a need to ensure that values
+are standard between codes. Examples are atomic
+masses and conversions between physical constants. These data
+can be used by others for any purpose (for example, for entry into
+Wikipedia or use in in-house software), and should lead
+to an enhancement in the quality of community reference data.
 
@@
 
-NMRShiftDB represents one of the earliest (2002?) resources for Open
-community-contributed data. Groups which
-measure NMR spectra or extract it from the literature contribute to
+NMRShiftDB represents one of the earliest resources for Open
+community-contributed data (first released in 2003). Research groups that 
+measure NMR spectra or extract it from the literature can contribute that
+information to
 NMRShiftDB which provides an Open resource where
 entries can be searched by chemical structure or properties
-(especially peaks). It is, not surprisingly, difficult to
-extract large amounts of altruistic contribution (as happens in
-wikipedia) but it is increasingly possible to link
+(especially peaks). Although it is difficult to
+encourage large amounts of altruistic contribution (as happens with
+Wikipedia), an alternative possible source of data could come from
+linking
 data capture with data publication. For example, the Blue Obelisk has
@@
 NMRShiftDB entries. If and when the chemistry community
-encourages or requires semantic publication of spectra, rather than
-PDFs, it would be possible to populate NMRShiftDB rapidaly
-along the the lines of CrystalEye.
+encourages or requires semantic publication of spectra rather than
+PDFs, it would be possible to populate NMRShiftDB rapidly
+along the the lines of CrystalEye (see below).
 
@@
 
-CrystalEye represents the cost-effective extraction of data from the
+CrystalEye~\cite{WebCrystalEye} is an example of cost-effective extraction of data from the
 literature where this is published both
-Openly and semantically. The cost is extremely low and software can be
-run every night, updating the entries
-(currently at about 250,000 structures). CrystalEye serves as a model
+Openly and semantically. Software extracts Openly-published crystal
+structures from a variety of scholarly journals, processes
+them and then makes them available through a web interface.
+It currently contains about 250,000 structures. CrystalEye serves as a model
 for a high-value, high-quality Open data
@@
 in ODOSOS, leading to development of many useful resources freely
-available to the chemistry community. Figure 2 shows how the various
-Blue Obelisk projects collaborate. But how best to engage with the
+available to the chemistry community. But how best to engage with the
 wider chemistry community outside of the Blue Obelisk remains an open
@@
 \section*{Figures}
-  \subsection*{Figure 1 - Blue Obelisk logo}
-
-  \subsection*{Figure 2 - Dependency diagram of Blue Obelisk projects.}
+
+  \subsection*{Figure 1 - Dependency diagram of some Blue Obelisk projects.}
       Each block represents a project. Square blocks show Open Data, ovals are Open Source,
-      and diamonds are Open Standards. Colors represent license: LGPL is green, GPL is orange,
-      and BSD is blue.
+      and diamonds are Open Standards. %%Colors represent license: LGPL is green, GPL is orange,
+      %%and BSD is blue.
 

### 2011-06-03 paper.tex
Spellcheck

@@
 from the fundamental tenants of the scientific method for data sharing
-and reproducibilty.
+and reproducibility.
 
@@
 projects are designed to be extended or scripted by users through
-the provision of a plugin architective and scripting support (using
+the provision of a plugin architecture and scripting support (using
 Bioclipse Scripting Language~\cite{Bioclipse2}, or Python in the case
@@
 
-Another area where the Blue Obelisk has had a signficant impact in the
+Another area where the Blue Obelisk has had a significant impact in the
 past five years is in supporting
@@
 decomposition analysis~\cite{Frenking} and Mulliken or C-squared analyses
-on user-defined molecular fragments. QMForge also provides a rudamentary
+on user-defined molecular fragments. QMForge also provides a rudimentary
 Cartesian coordinate editor allowing molecular structures to be saved via Open Babel.
@@
 A phrase from the meeting
-summed it up: ``The ice is beginning to melt'', signifiying that we can
+summed it up: ``The ice is beginning to melt'', signifying that we can
 expect a rapid increase in industry's interest
@@
 the flexibility that RDF offers. For larger
-systems, it is unclear whtether complete RDF solutions (e.g. Virtuoso)
+systems, it is unclear whether complete RDF solutions (e.g. Virtuoso)
 will be satisfactory or whether a hybrid system

### 2011-06-03 paper.tex
Finish figures

@@
 \cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
-on the CDK and Open Babel, respectively. Bioclipse is an award-winning
+on the CDK and Open Babel, respectively. Bioclipse (Figure~1) is an award-winning
 molecular workbench for life sciences
 that wraps cheminformatics functionality behind user-friendly interfaces and
-graphical editors while Avogadro is a 3D molecular editor and viewer aimed at
+graphical editors while Avogadro (Figure~2) is a 3D molecular editor and viewer aimed at
 preparing and analysing computational chemistry calculations. Both
@@
 and uses HTML5 to allow the scientist
-to present publication quality 2D and 3D graphics and animations for
+to present publication quality 2D and 3D graphics (see Figure~3) and animations for
 chemical structures, reactions and spectra. Beyond graphics, this tool
@@
 minimisation). In a similar manner, efficient Jmol code for atom-atom rebonding
-has been ported to the CDK. Figure 2 shows the collaborative nature of
+has been ported to the CDK. Figure~4 shows the collaborative nature of
 software developed in the Blue Obelisk, as one project builds on
@@
 issue, members of the Blue Obelisk established a Question \& Answer
-website\cite{WebBOShapado} (see Figure~3).  This is a website in the
+website\cite{WebBOShapado} (see Figure~5).  This is a website in the
 style of Stack Overflow\cite{WebStackOverflow} that encourages high quality answers (and
@@
 
-  \subsection*{Figure 1 - Dependency diagram of some Blue Obelisk projects.}
+ \subsection*{Figure 1 - Screenshot of Bioclipse using Jmol to
+visualise a molecular surface}
+
+ \subsection*{Figure 2 - Screenshot of Avogadro showing a depiction of
+a carbon nanotube}
+
+ \subsection*{Figure 3 - Screenshot of the MolGrabber 3D demo from
+ChemDoodle Web Components}
+
+  \subsection*{Figure 4 - Dependency diagram of some Blue Obelisk projects.}
 Each block represents a project. Square blocks show Open Data, ovals are Open Source,
-      and diamonds are Open Standards. %%Colors represent license: LGPL is green, GPL is orange,
-      %%and BSD is blue.
-
-  \subsection*{Figure 3 - Screenshot of the Blue Obelisk eXchange Question
+and diamonds are Open Standards.
+
+  \subsection*{Figure 5 - Screenshot of the Blue Obelisk eXchange Question
     and Answer website.}

### 2011-06-04 paper.tex
added references for NMRShiftDB and tried to clarify license

@@
 
-NMRShiftDB represents one of the earliest resources for Open
+NMRShiftDB \cite{nmrshiftdb, Steinbeck:2004do, Steinbeck2004} represents one of the earliest resources for Open
 community-contributed data (first released in 2003). Research groups that 
@@
   \textbf{ONS Solubility} & CC0 & Solubility data for various solvents. \\ \hline
-  \textbf{NMRShiftDB} & ? & Database of NMR spectra of organic compounds. \\ \hline
+  \textbf{NMRShiftDB} & Public Domain & Database of NMR spectra of organic compounds. \\ \hline
 \end{tabular}

### 2011-06-04 paper.tex
added reference for CDK-Taverna website in table
Fixed affiliation Chris

@@
     \iid(6)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
-    \iid(7)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
     \iid(8)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
@@
     \iid(10)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(11)... \\
+    \iid(11)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
     \iid(12)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
@@
        \multicolumn{3}{|c|}{\textbf{Integration}} \\ \hline
-        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
+        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdktaverna.wordpress.com & Christoph Steinbeck, Workflow \\ \hline
 		\textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ & Sam Adams et al. \\ \hline

### 2011-06-04 paper.tex
Proper enumerate environment

@@
 
-\begin{itemize}
-\item{1.}{\bf Molecules:} This is probably the largest area for Blue
+\begin{enumerate}
+\item {\bf Molecules:} This is probably the largest area for Blue
 Obelisk software and data, and is reflected by many programs that
@@
 compounds).
-\item{2.}{\bf Reactions:} Blue Obelisk software can describe the
+\item {\bf Reactions:} Blue Obelisk software can describe the
 semantics of reactions and provide atom-atom matching and analyse
 stoichiometric balance in reactions.
-\item{3.}{\bf Computational chemistry:} Blue Obelisk software can
+\item {\bf Computational chemistry:} Blue Obelisk software can
 interpret many of the current output files from calculations and
@@
 re-usability of computational chemistry.
-\item{4.}{\bf Spectra:} 1-D spectra (NMR, IR, UV etc.) are fully
+\item {\bf Spectra:} 1-D spectra (NMR, IR, UV etc.) are fully
 supported in Blue Obelisk offerings for conversion and display. There
@@
 ({\it e.g.} GC-MS).
-\item{5.}{\bf Crystallography:} The Blue Obelisk software supports the
+\item {\bf Crystallography:} The Blue Obelisk software supports the
 bi-directional processing of crystal structure files (CIF) and also
@@
 both periodic and aperiodic condensed objects.
-\end{itemize}
+\end{enumerate}
 

### 2011-06-04 paper.tex
Better AMBIT reference

@@
 A final example of second-generation Blue Obelisk software is the 
-AMBIT2~\cite{WebAMBIT} software, a GUI that facilitates registration
+AMBIT2~\cite{Jeliazkova2011} software, a GUI that facilitates registration
 of chemicals for the REACH EU directive on toxicity, and which is

### 2011-06-07 paper.tex
Fix line endings
Tidy affiliations and edit author order.

@@
 \author{
-  Samuel E Adams$^8$%
+  Noel M O'Boyle\correspondingauthor$^1$%
+  \email{Noel M O'Boyle\correspondingauthor - baoilleach@gmail.com}
+  \and
+  Samuel E Adams$^2$%
   \email{Samuel E Adams - sea36@cam.ac.uk}
   \and
-  Jonathan Alvarsson$^1$%
+  Jonathan Alvarsson$^3$%
   \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
   \and
-  Igor V Filippov$^2$%
+  Igor V Filippov$^4$%
   \email{Igor V Filippov - igorf@helix.nih.gov}
   \and
-  Rajarshi Guha$^2$%
+  Rajarshi Guha$^4$%
   \email{Rajarshi Guha - guhar@mail.nih.gov}
   \and
-  Robert Hanson$^3$%
+  Robert Hanson$^5$%
   \email{Robert Hanson - hansonr@stolaf.edu}
   \and
-  Marcus D Hanwell$^{4}$%
+  Marcus D Hanwell$^{6}$%
   \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
   \and
-  Geoffrey R Hutchison$^5$%
+  Geoffrey R Hutchison$^7$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Karol M Langner$^6$%
+  Karol M Langner$^8$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
   \and
-  Daniel M Lowe$^8$%
+  Daniel M Lowe$^2$%
   \email{Daniel Lowe - dl387@cam.ac.uk}
   \and
-  Peter Murray-Rust$^8$%
+  Dmitry Pavlov$^{9}$%
+  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
+  \and
+  Ola Spjuth$^3$%
+  \email{ola.spjuth@farmbio.uu.se}
+  \and
+  Christoph Steinbeck$^{10}$%
+  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
+  \and
+  Adam L Tenderholt$^{11}$%
+  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
+  \and
+  Kevin J Theisen$^{12}$%
+  \email{Kevin J Theisen - kevin@ichemlabs.com}
+  \and
+  Egon L Willighagen$^{13}$%
+  \email{Egon L Willighagen - egon.willighagen@ki.se}
+  \and
+  Peter Murray-Rust$^2$%
   \email{Peter Murray-Rust - pm286@cam.ac.uk}
-  \and
-  Noel M O'Boyle\correspondingauthor$^9$%
-  \email{Noel M O'Boyle\correspondingauthor - n.oboyle@ucc.ie}
-  \and
-  Dmitry Pavlov$^{10}$%
-  \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
-  \and
-  Ola Spjuth$^1$%
-  \email{ola.spjuth@farmbio.uu.se}
-  \and
-  Christoph Steinbeck$^{11}$%
-  \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
-  \and
-  Adam L Tenderholt$^{12}$%
-  \email{Adam L Tenderholt - adamlt82@u.washington.edu}
-  \and
-  Kevin J Theisen$^{13}$%
-  \email{Kevin J Theisen - kevin@ichemlabs.com}
-  \and
-  Egon L Willighagen$^{14}$%
-  \email{Egon L Willighagen - egon.willighagen@ki.se}
 }
@@
 \address{%
-    \iid(1)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(2)NIH Center for Translational Therapeutics, 9800 Medical  Center Drive, Rockville, MD 20878 \\
-    \iid(3)... \\
-    \iid(4)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(5)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(6)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
-    \iid(8)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
-    \iid(9)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
-    \iid(10)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(11)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(12)Department of Chemistry, University of Washington, Seattle, WA 98195 \\
-    \iid(13)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(14)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+    \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
+    \iid(2)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
+    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
+    \iid(4)NIH Center for Translational Therapeutics, 9800 Medical  Center Drive, Rockville, MD 20878, USA \\
+    \iid(5)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
+    \iid(6)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(7)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(8)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
+    \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(10)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(11)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
+    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
+    \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden

### 2011-06-07 paper.tex
Authors contributions and acknowledgements. Remove additional files.

@@
 \section*{Authors contributions}
-   Charles Darwin did all the work. The others stole the glory.
+   The inital layout of the manuscript grew from discussions between
+NMOB, RG and ELW. All of the authors contributed sections to this
+article, and approved the final manuscript.
 
@@
   \ifthenelse{\boolean{publ}}{\small}{}
-  Thanks to everyone.
+  NMOB is supported by a Health Research Board Career Development
+Fellowship (PD/2009/13).
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                               %%
-%% Additional Files              %%
-%%                               %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
-
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
-
 \end{bmcformat}

### 2011-06-07 paper.tex
Put in my "official" affiliation.
Tidy up references. A little more to go.

@@
 science). Many of the efforts relate to cheminformatics (the scope of
-this journal) and we believe that many of the publications in {\it J.
-Cheminf.} could be completely carried out using Blue Obelisk resources
+this journal) and we believe that many of the publications in Journal of
+Cheminformatics could be completely carried out using Blue Obelisk resources
 and other Open Source chemical tools. The importance of this is that for the
@@
 The Quixote project epitomises the full use of the Blue Obelisk
-software and is described in detail in a sibling
-article~\cite{QuixoteSameIssue}. Here we observe that it is possible
+software and is described in detail in another article
+in this issue. Here we observe that it is possible
 to convert legacy chemistry file formats
@@
 
-NMRShiftDB \cite{nmrshiftdb, Steinbeck:2004do, Steinbeck2004} represents one of the earliest resources for Open
+NMRShiftDB \cite{nmrshiftdb, Steinbeck2004} represents one of the earliest resources for Open
 community-contributed data (first released in 2003). Research groups that 

### 2011-06-07 paper.tex
Finish with references

@@
     \iid(7)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(8)Gorlaeus Laboratories, Leiden University, 2333 CC Leiden, The Netherlands \\
+    \iid(8)Leiden Institute of Chemistry, Universiteit Leiden, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
     \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
@@
 chemistry in CML and the management of semantics through CML
-dictionaries. At present, we use JUMBO-Converters~\cite{JUMBO-Convertors} for most
+dictionaries. At present, we use JUMBO-Converters~\cite{JUMBO-Converters} for most
  of the semantic conversion, Lensfield2~\cite{Lensfield2} for the workflow and Chempound

### 2011-06-08 paper.tex
proofreading
"several" and "have" hints that "feature" should be "features"

@@
 The CDK project has been under regular development over the last five
-years. Several feature have been implemented ranging from core
+years. Several features have been implemented ranging from core
 components such as an extensible SMARTS matching system and a new graph

### 2011-06-08 paper.tex
proofreading
adding missing "is"
Tidy some refs

--- 2011-06-07 paper.tex
@@
 later as a formal paper by Guha et al.\cite{Guha2006} in
-2006. This contribution looks back on the past 5 years and surveys
+2006. Its home on the web is at http://blueobelisk.org.
+This contribution looks back on the past 5 years and surveys
 progress and remaining challenges in the areas of Open Data, Open
@@
 is used by Kalzium~\cite{WebKalzium}, a periodic table and chemical editor in KDE,
-and XtalOpt~\cite{WebXtalOpt}\cite{Lonie2011}, an evolutionary
+and XtalOpt~\cite{WebXtalOpt, Lonie2011}, an evolutionary
 algorithm for crystal structure prediction. XtalOpt provides a
@@
 drug discovery
-field \cite{DeLano:2005uq}\cite{Munos:2006vn}\cite{Geldenhuys:2006kx}.
+field \cite{DeLano:2005uq, Munos:2006vn, Geldenhuys:2006kx}.
 

### 2011-06-08 paper.tex

@@
 how applications and content are distributed and presented. The web
-presents new opportunities for scientists as it
+presents new opportunities for scientists as it is
 an open and free medium to distribute scientific knowledge, ideas and

### 2011-06-08 paper.tex
Start tidying the tables. Reducing to two column or I'll never finish with it.

@@
 \usepackage[utf8]{inputenc} %unicode support
+\usepackage{lscape} % Landscape orientation for tables
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
@@
 \section*{Tables}
+%%\begin{landscape}
   \subsection*{Table 1 - Blue Obelisk Open Source software projects}
@@
      \begin{tabular}{|c|c|c|}
-       \hline \textbf{Name} & \textbf{Website} & \textbf{Description} \\ \hline
-        \multicolumn{3}{|c|}{\textbf{CML Tools}} \\ \hline
-		\textbf{JUMBO \& CMLXOM} & http://sourceforge.net/projects/cml/ https://bitbucket.org/wwmm/cmlxom/ & Peter Murray-Rust et al. \\ \hline
-		\multicolumn{3}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
-        \textbf{Chemistry Development Kit (CDK)} \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
-        \textbf{Cinfony} & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
-        \textbf{Indigo} & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
-        \textbf{Open Babel} & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
-        \textbf{RDKit} & http://rdkit.org & Greg Landrum \\ \hline
-        \multicolumn{3}{|c|}{\textbf{Second Generation}} \\ \hline
-        \textbf{Avogadro} & http://avogadro.openmolecules.net & XXX \\ \hline
-        \multicolumn{3}{|c|}{\textbf{Web Applications}} \\ \hline
-        \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com & iChemLabs \\ \hline
-        \textbf{Jmol} & http://jmol.org & Hanson \\ \hline
-       \multicolumn{3}{|c|}{\textbf{Integration}} \\ \hline
-        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdktaverna.wordpress.com & Christoph Steinbeck, Workflow \\ \hline
-		\textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ & Sam Adams et al. \\ \hline
-        \textbf{A3} & ..  & .  \\ \hline
-        \multicolumn{3}{|c|}{\textbf{Interconversion}} \\ \hline
-        \textbf{OSRA} & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
-        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
-		\textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ & Peter Murray-Rust et al. \\ \hline
-		\textbf{CIFXOM} & https://bitbucket.org/wwmm/cifxom/ & Peter Murray-Rust et al. \cite{DayEtAl2011} \\ \hline
-        \textbf{A3} & ..  & .  \\ \hline
-        \multicolumn{3}{|c|}{\textbf{Structure Databases}} \\ \hline
-        \textbf{Bingo}  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline
-        \textbf{MyChem}  & http://XXX.sf.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
-        \textbf{OrChem} \cite{RijnbeekS10} & http://orchem.sf.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
-        \textbf{PGChem}  & http://XXX.sf.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
- 	    \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound & Sam Adams et al. \\ \hline
-	   \multicolumn{3}{|c|}{\textbf{Text mining}} \\ \hline
-	    \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ & Peter Murray-Rust et al. \\ \hline
-		\textbf{ChemicalTagger} & http://chemicaltagger.ch.cam.ac.uk/ & Lezan Hawizy et al. \cite{HawizyEtAl2011} \\ \hline
-       \multicolumn{3}{|c|}{\textbf{Other}} \\ \hline
-        \textbf{cclib}  & http://cclib.sf.net & Python library for parsing and interpreting computational chemistry results \\
+       \hline \textbf{Name} & \textbf{Website}  \\ \hline
+        \multicolumn{2}{|c|}{\textbf{CML Tools}} \\ \hline
+		\textbf{JUMBO} & http://sourceforge.net/projects/cml/ \\ \hline
+		\textbf{CMLXOM} & https://bitbucket.org/wwmm/cmlxom/ \\ \hline
+		\multicolumn{2}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
+        \textbf{Chemistry Development Kit (CDK)} \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net \\ \hline
+        \textbf{Cinfony} & http://cinfony.googlecode.com \\ \hline
+        \textbf{Indigo} & http://ggasoftware.com/opensource/indigo \\ \hline
+        \textbf{Open Babel} & http://openbabel.org \\ \hline
+        \textbf{RDKit} & http://rdkit.org \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Second Generation}} \\ \hline
+        \textbf{Avogadro} & http://avogadro.openmolecules.net \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Web Applications}} \\ \hline
+        \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com \\ \hline
+        \textbf{Jmol} & http://jmol.org \\ \hline
+       \multicolumn{2}{|c|}{\textbf{Integration}} \\ \hline
+        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdktaverna.wordpress.com \\ \hline
+		\textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Interconversion}} \\ \hline
+        \textbf{OSRA} & http://osra.sf.net \\ \hline
+        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk \\ \hline
+		\textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ \\ \hline
+		\textbf{CIFXOM} \cite{DayEtAl2011} & https://bitbucket.org/wwmm/cifxom/  \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Structure Databases}} \\ \hline
+        \textbf{Bingo}  & http://XXX.XXX.XXX \\ \hline
+        \textbf{MyChem}  & http://XXX.sf.net \\ \hline
+        \textbf{OrChem} \cite{RijnbeekS10} & http://orchem.sf.net \\ \hline
+        \textbf{PGChem}  & http://XXX.sf.net \\ \hline
+ 	    \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
+	   \multicolumn{2}{|c|}{\textbf{Text mining}} \\ \hline
+	    \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ \\ \hline
+		\textbf{ChemicalTagger} \cite{HawizyEtAl2011} & http://chemicaltagger.ch.cam.ac.uk/ \\ \hline
+       \multicolumn{2}{|c|}{\textbf{Other}} \\ \hline
+        \textbf{cclib}  & http://cclib.sf.net \\
         \hline
@@
       }
+%%\end{landscape}
+
   \subsection*{Table 2 - Open Data in chemistry.}

### 2011-06-08 paper.tex
More table action.

@@
 \usepackage[utf8]{inputenc} %unicode support
-\usepackage{lscape} % Landscape orientation for tables
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
@@
 \section*{Tables}
-%%\begin{landscape}
-  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
-    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
-      %% GeoffH - I think we should do this as multi-line, with the
-      %% website below the Name
+  \subsection*{Table 1 - Blue Obelisk Open Source Software projects}
     \par \mbox{}
@@
         \multicolumn{2}{|c|}{\textbf{CML Tools}} \\ \hline
+        \textbf{CMLXOM} & https://bitbucket.org/wwmm/cmlxom/ \\ \hline
         \textbf{JUMBO} & http://sourceforge.net/projects/cml/ \\ \hline
-		\textbf{CMLXOM} & https://bitbucket.org/wwmm/cmlxom/ \\ \hline
         \multicolumn{2}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
@@
         \textbf{RDKit} & http://rdkit.org \\ \hline
-        \multicolumn{2}{|c|}{\textbf{Second Generation}} \\ \hline
-        \textbf{Avogadro} & http://avogadro.openmolecules.net \\ \hline
         \multicolumn{2}{|c|}{\textbf{Web Applications}} \\ \hline
@@
         \multicolumn{2}{|c|}{\textbf{Integration}} \\ \hline
+        \textbf{Bioclipse} & http://www.bioclipse.net \\ \hline
         \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdktaverna.wordpress.com \\ \hline
@@
         \multicolumn{2}{|c|}{\textbf{Interconversion}} \\ \hline
+        \textbf{CIFXOM} \cite{DayEtAl2011} & https://bitbucket.org/wwmm/cifxom/  \\ \hline
+        \textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ \\ \hline
+        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk \\ \hline
         \textbf{OSRA} & http://osra.sf.net \\ \hline
-        \textbf{OPSIN} & http://opsin.ch.cam.ac.uk \\ \hline
-		\textbf{JUMBO-Converters} & https://bitbucket.org/wwmm/jumbo-converters/ \\ \hline
-		\textbf{CIFXOM} \cite{DayEtAl2011} & https://bitbucket.org/wwmm/cifxom/  \\ \hline
         \multicolumn{2}{|c|}{\textbf{Structure Databases}} \\ \hline
-        \textbf{Bingo}  & http://XXX.XXX.XXX \\ \hline
-        \textbf{MyChem}  & http://XXX.sf.net \\ \hline
+        \textbf{Bingo}  & http://ggasoftware.com/opensource/bingo \\ \hline
+        \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
+        \textbf{MyChem}  & http://mychem.sf.net \\ \hline
         \textbf{OrChem} \cite{RijnbeekS10} & http://orchem.sf.net \\ \hline
-        \textbf{PGChem}  & http://XXX.sf.net \\ \hline
- 	    \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
+        \textbf{pgchem}  & http://pgfoundry.org/projects/pgchem/ \\ \hline
         \multicolumn{2}{|c|}{\textbf{Text mining}} \\ \hline
+        \textbf{ChemicalTagger} \cite{HawizyEtAl2011} & http://chemicaltagger.ch.cam.ac.uk/ \\ \hline
         \textbf{OSCAR4} & https://bitbucket.org/wwmm/oscar4/ \\ \hline
-		\textbf{ChemicalTagger} \cite{HawizyEtAl2011} & http://chemicaltagger.ch.cam.ac.uk/ \\ \hline
-       \multicolumn{2}{|c|}{\textbf{Other}} \\ \hline
-        \textbf{cclib}  & http://cclib.sf.net \\
+        \multicolumn{2}{|c|}{\textbf{Computational Chemistry}} \\ \hline
+        \textbf{Avogadro} & http://avogadro.openmolecules.net \\ \hline
+        \textbf{cclib}  & http://cclib.sf.net \\ \hline
+        \textbf{GaussSum}  & http://gausssum.sf.net \\ \hline
+        \textbf{QMForge}  & http://qmforge.sf.net \\ \hline
         \hline
-%% GaussSum, QMForge
        \end{tabular}
       }
-%%\end{landscape}
 

### 2011-06-09 paper.tex
Added gChem paragraph, updated Chempedia information and links. Minor typo correction.

@@
 
+\subsection*{Web services}
+
+Web services offer a potentially powerful application of Open Source software,
+particularly when combined with Open Data. For example, Metamolecular~\cite{Metamolecular}Metamolecular have released
+gChem~\cite{gChem}, an Open Source library for Google Spreadsheets that enables
+non-programmers to build and maintain chemically-aware spreadsheet documents
+that can be easily shared and published. gChem makes use of the National Cancer Institute's
+Chemical Identifier Resolver~\cite{ChemicalIdentifierResolver} to interconvert
+substance names, database identifiers, SMILES and InChI codes, and to generate
+embedded chemical structure images.
+
 \subsection*{The business end}
@@
 community. Examples of such companies include iChemLabs, IdeaConsult, Wingu, Silicos,
-GenettaSoft, hBar, and Inkspot Science. Some of these merely use components, but several
+GenettaSoft, hBar, Metamolecular, and Inkspot Science. Some of these merely use components, but several
 actively contribute back to the Blue Obelisk project they use, or donate new
@@
 
-Silicos has release several Open Source utilities based on
+Silicos has released several Open Source utilities based on
 Open Babel, such as Pharao, a tool for pharmacophore searching,
@@
 code in the CDK, eMolecules donated canonicalisation code to
-Open Babel, and AstraZeneca contributed code to the CDK for
-signatures. This is just a very minor selection, and the author
+Open Babel, Metamolecular improved the extensibility and unit testing suite of OPSIN,
+and AstraZeneca contributed code to the CDK for
+signatures. This is just a very minor selection, and the reader
 is encouraged to contact the individual Blue Obelisk projects
@@
 inorganic, solid state chemistry. Table~2 lists a few notable
-exceptions: ChemPedia (a now discontinued crowd-sourcing project),
+exceptions: Metamolecular's Chempedia~\cite{Chempedia} (a discontinued substance registration
+and naming service whose raw data can still be accessed),
 CrystalEye~\cite{WebCrystalEye},

### 2011-06-09 paper.tex
Add Apodaca author information, add Resolver url.

@@
   \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
+  \and
+  Richard L Apodaca$^{14}$%
+  \email{Richard L Apodaca - rapodaca@metamolecular.com}
   \and
@@
     \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
+    \iid(14)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA
 }%

### 2011-06-09 paper.tex
Integrating Rich's text, and minor changes

@@
 
-\subsection*{Web services}
-
 Web services offer a potentially powerful application of Open Source software,
-particularly when combined with Open Data. For example, Metamolecular~\cite{Metamolecular}Metamolecular have released
+particularly when combined with Open Data. For example, Metamolecular~\cite{Metamolecular}have released
 gChem~\cite{gChem}, an Open Source library for Google Spreadsheets that enables
@@
 community. Examples of such companies include iChemLabs, IdeaConsult, Wingu, Silicos,
-GenettaSoft, hBar, Metamolecular, and Inkspot Science. Some of these merely use components, but several
+GenettaSoft, eMolecules, hBar, Metamolecular, and Inkspot Science. Some of these merely use components, but several
 actively contribute back to the Blue Obelisk project they use, or donate new

### 2011-06-09 paper.tex
Add Andy Lang and JC Bradley as authors

@@
   \and
-  Richard L Apodaca$^{14}$%
+  Richard L Apodaca$^{4}$%
   \email{Richard L Apodaca - rapodaca@metamolecular.com}
   \and
-  Igor V Filippov$^4$%
+  Jean-Claude Bradley$^5$%
+  \email{Jean-Claude Bradley - bradlejc@drexel.edu}
+  \and
+  Igor V Filippov$^6$%
   \email{Igor V Filippov - igorf@helix.nih.gov}
   \and
-  Rajarshi Guha$^4$%
+  Rajarshi Guha$^6$%
   \email{Rajarshi Guha - guhar@mail.nih.gov}
   \and
-  Robert Hanson$^5$%
+  Robert Hanson$^7$%
   \email{Robert Hanson - hansonr@stolaf.edu}
   \and
-  Marcus D Hanwell$^{6}$%
+  Marcus D Hanwell$^{8}$%
   \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
   \and
-  Geoffrey R Hutchison$^7$%
+  Geoffrey R Hutchison$^9$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Karol M Langner$^8$%
+  Andrew SID Lang$^{10}$%
+  \email{Andrew SID Lang - alang@oru.edu}
+  \and
+  Karol M Langner$^{11}$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
@@
   \and
-  Dmitry Pavlov$^{9}$%
+  Dmitry Pavlov$^{12}$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
@@
   Ola Spjuth$^3$%
-  \email{ola.spjuth@farmbio.uu.se}
-  \and
-  Christoph Steinbeck$^{10}$%
+  \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
+  \and
+  Christoph Steinbeck$^{13}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{11}$%
+  Adam L Tenderholt$^{14}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{12}$%
+  Kevin J Theisen$^{15}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
   \and
-  Egon L Willighagen$^{13}$%
+  Egon L Willighagen$^{16}$%
   \email{Egon L Willighagen - egon.willighagen@ki.se}
@@
     \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(4)NIH Center for Translational Therapeutics, 9800 Medical  Center Drive, Rockville, MD 20878, USA \\
-    \iid(5)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
-    \iid(6)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(7)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(8)Leiden Institute of Chemistry, Universiteit Leiden, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
-    \iid(9)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(10)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(11)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
-    \iid(12)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(13)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
-    \iid(14)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA
+    \iid(4)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
+    \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
+    \iid(6)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
+    \iid(7)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
+    \iid(8)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(9)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(10)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
+    \iid(11)Leiden Institute of Chemistry, Universiteit Leiden, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
+    \iid(12)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(13)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
+    \iid(15)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
+    \iid(16)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
 }%

### 2011-06-09 paper.tex
Done for today.

@@
 substantial knowledge base.
+
+\subsection*{Reaction Attempts}
+Although there are existing databases that allow for searching reactions, those
+using Open Data are harder to find. The Reaction Attempts database
+\cite{ReactionAttempts}, to which anyone can
+submit reaction attempts data, consists mainly of reaction information
+abstracted from Open Notebooks in organic chemistry, such as the
+UsefulChem project from the Bradley group
+\cite{UsefulChemReactionAttempts}
+and the notebooks from the Todd group \cite{UsefulChemTodd}.
+Key information from each experiment is abstracted manually, with the
+only required information consisting of the ChemSpider IDs of the
+reactants and the product targeted in the experiment; and a link to
+the laboratory notebook page.  Since the database reflects all data
+from the notebooks, many experiments will be short of completely
+optimized and includes experiments in progress, ambiguous results and
+failed runs.  Unlike most reaction databases that only identify
+experiments successfully reported in the literature, Reaction Attempts
+is meant to be used to easily discover what researchers are trying to
+accomplish thus potentially forging collaborations more quickly~\cite{UsefulChemSocialNetworks}
+or providing useful knowledge about what doesn't work.
+
+The database can be accessed using
+a web-based Reaction Attempts Explorer
+\cite{ReactionAttemptsExplorer}.
+The explorer allows you to find reaction attempts either by reactant
+or product name, via a series of dynamic dropdowns, or by performing a
+substructure search. Found reactions are shown graphically as schemes
+with accompanying structure images and text with links to the lab
+notebook pages, the ChemSpider pages for each compound, and a separate
+web service, The Optimal Solvent Selector
+[http://usefulchem.blogspot.com/2010/08/reaction-attempts-solvent-selector.html,
+that predicts the solubility of the reactants and product in 72
+solvents and sorts by theoretical maximum yield.  The Reaction
+Attempts Explorer allows researchers to easily find patterns in
+reactions that have already been performed and since the data are open
+and results are reported across all research groups, intersections are
+easily discovered and possible Open Collaboration opportunities are
+easily found \cite{ekins_collaboration_2011}.
 

### 2011-06-10 paper.tex
Added text from Andy Lang and JC Bradley

--- 2011-06-09 paper.tex
@@
 substantial knowledge base.
+
+\subsection*{Reaction Attempts}
+Although there are existing databases that allow for searching reactions, those
+using Open Data are harder to find. The Reaction Attempts database
+\cite{ReactionAttempts}, to which anyone can
+submit reaction attempts data, consists mainly of reaction information
+abstracted from Open Notebooks in organic chemistry, such as the
+UsefulChem project from the Bradley group
+\cite{UsefulChemReactionAttempts}
+and the notebooks from the Todd group \cite{UsefulChemTodd}.
+Key information from each experiment is abstracted manually, with the
+only required information consisting of the ChemSpider IDs of the
+reactants and the product targeted in the experiment; and a link to
+the laboratory notebook page. Information in the database can be
+searched and accessed using the web-based Reaction Attempts Explorer
+\cite{ReactionAttemptsExplorer}.
+
+Since the database reflects all data
+from the notebooks, it includes experiments in progress, ambiguous results and
+failed runs. Unlike most reaction databases that only identify
+experiments successfully reported in the literature, 
+the Reaction
+Attempts Explorer allows researchers to easily find patterns in
+reactions that have already been performed, 
+and since the data are open
+and results are reported across all research groups, intersections are
+easily discovered and possible Open Collaboration opportunities are
+easily found \cite{UsefulChemSocialNetworks, ekins_collaboration_2011}.
+
+\subsection*{Non-Aqueous Solubility}
+
+Although the aqueous solubility of many common organic compounds is
+generally available, quantitative reports of non-aqueous solubility
+are more difficult to find.
+Such information can be valuable for selecting solvents for
+reactions, re-crystallization and related processes. In 2008, the Open
+Notebook Science Solubility Challenge was launched for the purpose of
+measuring non-aqueous solubility of organic compounds, reporting all
+the details of the experiments in an Open Notebook and recording the
+results as Open Data in a centralized
+database~\cite{ONS2010, BeautifulData_2009}.
+This crowdsourcing project was also supported by Submeta,
+Sigma-Aldrich, Nature Publishing Group and the Royal Society of
+Chemistry. The database currently holds 1932 total measurements and
+1428 averaged solute/solvent measurements all of which are available under
+a CC0 license.  Several web services and
+feeds are available to filter and re-use the dataset~\cite{SolubilityServices}.
+In particular, models have been developed for the prediction of
+non-aqueous solubility in 72 different
+solvents~\cite{UCSolubilityPrediction}
+using the method of Abraham et al~\cite{AbrahamSolubility} with
+descriptors calculated by
+the Chemistry Development Kit. These models are available online and
+will be refined as more solubility data is collected.
 

### 2011-06-10 paper.tex
Edits to Tables 1 and 2.

@@
 
-Silicos has released several Open Source utilities based on
+Silicos has released several Open Source utilities~\cite{SilicosDownloads} based on
 Open Babel, such as Pharao, a tool for pharmacophore searching,
@@
 \section*{Tables}
-  \subsection*{Table 1 - Blue Obelisk Open Source Software projects}
+  \subsection*{Table 1 - Blue Obelisk Open Source Software projects listed in the text}
     \par \mbox{}
@@
         \multicolumn{2}{|c|}{\textbf{Cheminformatics Toolkits}} \\ \hline
-        \textbf{Chemistry Development Kit (CDK)} \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net \\ \hline
+        \textbf{Chemistry Development Kit (CDK)} & http://cdk.sf.net \\ \hline
         \textbf{Cinfony} & http://cinfony.googlecode.com \\ \hline
         \textbf{Indigo} & http://ggasoftware.com/opensource/indigo \\ \hline
+        \textbf{JOELib} & http://sf.net/projects/joelib \\ \hline
         \textbf{Open Babel} & http://openbabel.org \\ \hline
@@
         \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com \\ \hline
+        \textbf{gChem} & http://metamolecular.com/gchem/ \\ \hline
         \textbf{Jmol} & http://jmol.org \\ \hline
@@
         \textbf{Bioclipse} & http://www.bioclipse.net \\ \hline
-        \textbf{CDK-Taverna} \cite{Kuhn:2010p4001} & http://cdktaverna.wordpress.com \\ \hline
+        \textbf{CDK-Taverna} & http://cdktaverna.wordpress.com \\ \hline
         \textbf{Lensfield2} & https://bitbucket.org/sea36/lensfield2/ \\ \hline
@@
         \textbf{MyChem}  & http://mychem.sf.net \\ \hline
-        \textbf{OrChem} \cite{RijnbeekS10} & http://orchem.sf.net \\ \hline
+        \textbf{OrChem} & http://orchem.sf.net \\ \hline
         \textbf{pgchem}  & http://pgfoundry.org/projects/pgchem/ \\ \hline
@@
         \textbf{QMForge}  & http://qmforge.sf.net \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Computational Drug Design}} \\ \hline
+        \textbf{Confab}~\cite{Confab} & http://confab.googlecode.com \\ \hline
+        \textbf{Pharao} & http://silicos.be/download \\ \hline
+        \textbf{Piramid} & http://silicos.be/download \\ \hline
+        \textbf{Sieve} & http://silicos.be/download \\ \hline
+        \textbf{Stripper} & http://silicos.be/download \\ \hline
+        \multicolumn{2}{|c|}{\textbf{Other Applications}} \\ \hline
+        \textbf{AMBIT2}  & http://ambit.sf.net \\ \hline
+        \textbf{Brunn}  & http://brunn.sf.net \\ \hline
+        \textbf{XtalOpt}  & http://xtalopt.openmolecules.net \\ \hline
         \hline
@@
   \hline \textbf{Name} & \textbf{License/Waiver} & \textbf{Description} \\ \hline
-  \textbf{ChemPedia} & CC0 & Crowd-sourced chemical names. Project discontinued. \\ \hline
-  \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature. \\ \hline
-  \textbf{ONS Solubility} & CC0 & Solubility data for various solvents. \\ \hline
-  \textbf{NMRShiftDB} & Public Domain & Database of NMR spectra of organic compounds. \\ \hline
+  \textbf{ChemPedia} & CC0 & Crowd-sourced chemical names (project discontinued) \\ \hline
+  \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature \\ \hline
+  \textbf{NMRShiftDB} & Public Domain & Database of NMR spectra of organic compounds \\ \hline
+  \textbf{ONS Solubility} & CC0 & Solubility data for various solvents \\ \hline
+  \textbf{Reaction Attempts} & CC0 & Data on successful and unsuccessful reactions \\ \hline
 \end{tabular}

### 2011-06-10 paper.tex
Add a link to Table 1 in the text

@@
 
+The development of Open Source software has been one of the most
+successful of the Blue Obelisk's activities. The following sections
+describe recent work in this area, and Table 1
+provides an overview of the projects
+discussed and where to find them online.
+
 \subsection*{Cheminformatics toolkits}
@@
 \section*{Tables}
-  \subsection*{Table 1 - Blue Obelisk Open Source Software projects listed in the text}
+  \subsection*{Table 1 - Blue Obelisk Open Source Software projects discussed in the text}
     \par \mbox{}

### 2011-06-13 paper.tex
Fix typos, etc. pointed out by co-authors.

@@
   \and
-  Robert Hanson$^7$%
-  \email{Robert Hanson - hansonr@stolaf.edu}
+  Robert M Hanson$^7$%
+  \email{Robert M Hanson - hansonr@stolaf.edu}
   \and
@@
 chemistry community and a variety of programs exist, mostly
-proprietary software with the notable exception of
-NWChem~\cite{NWChem}.
+proprietary software but with an increasing number of open source codes.
 However, since each program uses different input formats, and the

### 2011-06-13 paper.tex
NMRShiftDB is not public domain; it's GNU FDL
Added a practical use case of three BO products: Oscar[3], Bioclipse, and NMRShiftDB.org
Remove early refs to BO projects.

@@
 using Blue Obelisk components and standard infrastructure such as
-workflows (KNIME or Taverna) and databases (MySQL, CouchDB etc.).
-Where the Blue Obelisk itself produces data resources (NMRShiftDB,
-CrystalEye, Quixote etc.) they are normally done with Open components
+workflows and databases.
+Where the Blue Obelisk itself produces data resources
+they are normally done with Open components
 so that the community can, if necessary, replicate them.
@@
   \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature \\ \hline
-  \textbf{NMRShiftDB} & Public Domain & Database of NMR spectra of organic compounds \\ \hline
   \textbf{ONS Solubility} & CC0 & Solubility data for various solvents \\ \hline

### 2011-06-13 paper.tex
Remove use of abbreviation for Blue Obelisk

@@
 supports 8 different computational chemistry codes and extracts over
-30 different calculated attributes. Two related BO projects build upon 
+30 different calculated attributes. Two related Blue Obelisk projects build upon 
 cclib. GaussSum~\cite{WebGaussSum},
@@
 Currently, most of the individual components of a `standardisation
-pipeline' can be implemented using BO tools. The larger problem is
+pipeline' can be implemented using Blue Obelisk tools. The larger problem is
 that there is no agreed upon list of steps for a standardisation
@@
 PDFs, it would be possible to populate NMRShiftDB rapidly
-along the the lines of CrystalEye (see below).
+along the the lines of CrystalEye (see below). A similar approach
+has been demonstrated earlier using the Blue Obelisk components
+Oscar and Bioclipse using text mining approaches~\cite{NMRExtraction}.
 
@@
 While each Blue Obelisk project has its own website and point of
-contact (typically a mailing list), because of the breadth of BO
+contact (typically a mailing list), because of the breadth of Blue Obelisk
 projects it can be difficult for a newcomer to understand which of
@@
 then an attempt must be made to reach beyond the subscribers to the
-BO mailing list and blogs of members.
+Blue Obelisk mailing list and blogs of members.
 

### 2011-06-13 paper.tex
Fix typo reported by Kevin Thiesen

@@
 
-At the time of writing there are many commercial companies development chemistry
+At the time of writing there are many commercial companies developing chemistry
 solutions around Open Source cheminformatics components provided by the Blue Obelisk

### 2011-06-15 paper.tex
Minor corrections

@@
   \and
-  Rajarshi Guha$^6$%
+  Rajarshi Guha$^7$%
   \email{Rajarshi Guha - guhar@mail.nih.gov}
   \and
-  Robert M Hanson$^7$%
+  Robert M Hanson$^8$%
   \email{Robert M Hanson - hansonr@stolaf.edu}
   \and
-  Marcus D Hanwell$^{8}$%
+  Marcus D Hanwell$^{9}$%
   \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
   \and
-  Geoffrey R Hutchison$^9$%
+  Geoffrey R Hutchison$^{10}$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Andrew SID Lang$^{10}$%
+  Andrew SID Lang$^{11}$%
   \email{Andrew SID Lang - alang@oru.edu}
   \and
-  Karol M Langner$^{11}$%
+  Karol M Langner$^{12}$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
@@
   \and
-  Dmitry Pavlov$^{12}$%
+  Dmitry Pavlov$^{13}$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
@@
   \and
-  Christoph Steinbeck$^{13}$%
+  Christoph Steinbeck$^{14}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{14}$%
+  Adam L Tenderholt$^{15}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{15}$%
+  Kevin J Theisen$^{16}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
   \and
-  Egon L Willighagen$^{16}$%
+  Egon L Willighagen$^{17}$%
   \email{Egon L Willighagen - egon.willighagen@ki.se}
@@
     \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
-    \iid(6)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
-    \iid(7)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
-    \iid(8)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(9)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(10)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
-    \iid(11)Leiden Institute of Chemistry, Universiteit Leiden, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
-    \iid(12)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(13)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(14)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
-    \iid(15)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(16)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+    \iid(6)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
+Inc., NCI-Frederick, Frederick, MD 21702, USA \\
+    \iid(7)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
+    \iid(8)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
+    \iid(9)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(11)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
+    \iid(12)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
+    \iid(13)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(14)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(15)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
+    \iid(16)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
+    \iid(17)Division of Molecular Toxicology, Institute of Environmental Medicine, %
         Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
@@
 		in ODOSOS, leading to development of many useful resources freely
-		available to the chemistry community. But how best to engage with the
-		wider chemistry community outside of the Blue Obelisk remains an open
-		question.
+		available to the chemistry community.
 \end{abstract}
@@
 
-However, Blue Obelisk software and data is used outside
+However, Blue Obelisk software and data is also used outside
 cheminformatics and certainly in the five main areas that, for
@@
 Recently, a new version of the HTML
-specification, HTML5, defines a well-developed framework
-for creating native web applications in JavaScript and opens up
+specification, HTML5, defined a well-developed framework
+for creating native web applications in JavaScript and this opens up
 new possibilities for visualising chemical data.
@@
 the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values,
-units, and versioned descriptor implementations to allow for
-integrating several descriptor software in the same calculation.
+units, and versioned descriptor implementations to allow
+descriptors from different software to be integrated into the same calculation.
 Hence, a dataset described in QSAR-ML is completely reproducible. To
@@
 in ODOSOS, leading to development of many useful resources freely
-available to the chemistry community. But how best to engage with the
+available to the chemistry community. However, how best to engage with the
 wider chemistry community outside of the Blue Obelisk remains an open
@@
 Fellowship (PD/2009/13).
+This project has been funded in whole or in part with federal funds from
+the National Cancer Institute, National Institutes of Health, under
+contract HHSN261200800001E. The content of this publication does not
+necessarily reflect the views of the policies of the Department of
+Health and Human Services, nor does mention of trade names, commercial
+products, or organisations imply endorsement by the US Government.
 

### 2011-06-17 paper.tex
Minor correction
Change acknowledgement so that it's clear that only OSRA was dunded by NIH

@@
 Fellowship (PD/2009/13).
-This project has been funded in whole or in part with federal funds from
+The OSRA project has been funded in whole or in part with federal funds from
 the National Cancer Institute, National Institutes of Health, under
@@
   \hline \textbf{Name} & \textbf{License/Waiver} & \textbf{Description} \\ \hline
-  \textbf{ChemPedia} & CC0 & Crowd-sourced chemical names (project discontinued) \\ \hline
+  \textbf{Chempedia} & CC0 & Crowd-sourced chemical names (project discontinued) \\ \hline
   \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature \\ \hline

### 2011-06-20 paper.tex
Correct spelling of Mychem

@@
 open source chemistry database cartridges (PgChem~\cite{WebPgChem},
-MyChem~\cite{WebMyChem}, OrChem~\cite{RijnbeekS10} and
+Mychem~\cite{WebMychem}, OrChem~\cite{RijnbeekS10} and
 more recently Bingo~\cite{WebBingo}) with a view to making their database APIs more
 similar and collaborating on benchmark datasets for assessing
-performance. For two of these projects, PgChem and MyChem, which are both based on
+performance. For two of these projects, PgChem and Mychem, which are both based on
 Open Babel, there is the additional possibility of working together on a shared
@@
         \textbf{Chempound (Chem\#)} & https://bitbucket.org/chempound \\ \hline
-        \textbf{MyChem}  & http://mychem.sf.net \\ \hline
+        \textbf{Mychem}  & http://mychem.sf.net \\ \hline
         \textbf{OrChem} & http://orchem.sf.net \\ \hline

### 2011-06-22 paper.tex
Halfway through final edits

@@
   \and
-  Samuel E Adams$^2$%
+  Rajarshi Guha$^2$%
+  \email{Rajarshi Guha - guhar@mail.nih.gov}
+  \and
+  Egon L Willighagen$^{3}$%
+  \email{Egon L Willighagen - egon.willighagen@ki.se}
+  \and
+  Samuel E Adams$^4$%
   \email{Samuel E Adams - sea36@cam.ac.uk}
   \and
-  Jonathan Alvarsson$^3$%
+  Jonathan Alvarsson$^5$%
   \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
   \and
-  Richard L Apodaca$^{4}$%
+  Richard L Apodaca$^{6}$%
   \email{Richard L Apodaca - rapodaca@metamolecular.com}
   \and
-  Jean-Claude Bradley$^5$%
+  Jean-Claude Bradley$^7$%
   \email{Jean-Claude Bradley - bradlejc@drexel.edu}
   \and
-  Igor V Filippov$^6$%
+  Igor V Filippov$^8$%
   \email{Igor V Filippov - igorf@helix.nih.gov}
   \and
-  Rajarshi Guha$^7$%
-  \email{Rajarshi Guha - guhar@mail.nih.gov}
-  \and
-  Robert M Hanson$^8$%
+  Robert M Hanson$^9$%
   \email{Robert M Hanson - hansonr@stolaf.edu}
   \and
-  Marcus D Hanwell$^{9}$%
+  Marcus D Hanwell$^{10}$%
   \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
   \and
-  Geoffrey R Hutchison$^{10}$%
+  Geoffrey R Hutchison$^{11}$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Andrew SID Lang$^{11}$%
+  Craig A James$^{12}$%
+  \email{Craig A James - cjames@emolecules.com}
+  \and
+  Nina Jeliazkova$^{13}$%
+  \email{Nina Jeliazkova - jeliazkova.nina@gmail.com}
+  \and
+  Andrew SID Lang$^{14}$%
   \email{Andrew SID Lang - alang@oru.edu}
   \and
-  Karol M Langner$^{12}$%
+  Karol M Langner$^{15}$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
   \and
-  Daniel M Lowe$^2$%
+  David C Lonie$^{16}$%
+  \email{David C Lonie - loniedavid@gmail.com}
+  \and
+  Daniel M Lowe$^4$%
   \email{Daniel Lowe - dl387@cam.ac.uk}
   \and
-  Dmitry Pavlov$^{13}$%
+  J\'{e}r\^{o}me Pansanel$^{17}$%
+  \email{J\'{e}r\^{o}me Pansanel - j.pansanel@pansanel.net}
+  \and
+  Dmitry Pavlov$^{18}$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
   \and
-  Ola Spjuth$^3$%
+  Ola Spjuth$^{5}$%
   \email{Ola Spjuth - ola.spjuth@farmbio.uu.se}
   \and
-  Christoph Steinbeck$^{14}$%
+  Christoph Steinbeck$^{19}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{15}$%
+  Adam L Tenderholt$^{20}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{16}$%
+  Kevin J Theisen$^{21}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
   \and
-  Egon L Willighagen$^{17}$%
-  \email{Egon L Willighagen - egon.willighagen@ki.se}
-  \and
-  Peter Murray-Rust$^2$%
+  Peter Murray-Rust$^4$%
   \email{Peter Murray-Rust - pm286@cam.ac.uk}
@@
     \iid(1)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland \\
-    \iid(2)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
-    \iid(3)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(4)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
-    \iid(5)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
-    \iid(6)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
+    \iid(2)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
+    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
+        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
+    \iid(4)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
+    \iid(5)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
+    \iid(6)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
+    \iid(7)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
+    \iid(8)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
 Inc., NCI-Frederick, Frederick, MD 21702, USA \\
-    \iid(7)NIH Center for Translational Therapeutics, 9800 Medical Center Drive, Rockville, MD 20878, USA \\
-    \iid(8)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
-    \iid(9)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(11)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
-    \iid(12)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
-    \iid(13)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(14)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(15)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
-    \iid(16)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA \\
-    \iid(17)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden 
+    \iid(9)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
+    \iid(10)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(11)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(12)eMolecules Inc., 380 Stevens Ave., Solana Beach, California 92075, USA \\
+    \iid(13)Ideaconsult Ltd., 4.A.Kanchev str., Sofia 1000, Bulgaria \\
+    \iid(14)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
+    \iid(15)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
+    \iid(16)Department of Chemistry, State University of New York at Buffalo, Buffalo, NY 14260-3000, USA \\
+    \iid(17)Universit\'{e} de Strasbourg, IPHC, CNRS, UMR7178, 23 rue du Loess 67037, Strasbourg, France \\
+    \iid(18)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(19)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(20)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
+    \iid(21)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA 
 }%
@@
 A final example of second-generation Blue Obelisk software is the 
-AMBIT2~\cite{Jeliazkova2011} software, a GUI that facilitates registration
-of chemicals for the REACH EU directive on toxicity, and which is
-based on the CDK.
+AMBIT2~\cite{Jeliazkova2011,InSilicoTox_2010} software, which was developed to facilitate registration of chemicals for the REACH EU directive, and is based on the CDK.
+It was distributed initially as a standalone Java Swing GUI, and more recently as downloadable web application archive, offering a web services interface to a searchable chemical structures database.
+Also integrated are descriptor calculations, as well as the ability to run and build predictive models, including modules of the open source Toxtree~\cite{WebToxTree,InSilicoTox_2010,Toxtree} software for toxicity prediction.
 
@@
 handling of cis/trans bond symbols at ring closures. In 2007, Craig
-James (eMolecules) initiated work on the OpenSMILES specification, a
+James (eMolecules) initiated work on the OpenSMILES specification~\cite{WebOpenSMILES}, a
 complete specification of the SMILES language as an Open Standard
@@
 to an enhancement in the quality of community reference data.
+The Blue Obelisk provides also a complementary project, the Chemical
+Structure Repository~\cite{BODR}. It aims to provide 3D coordinates, InChi code and several
+physico-chemical descriptors for a set of 570 organic compounds.
 
@@
         \textbf{Brunn}  & http://brunn.sf.net \\ \hline
+        \textbf{Toxtree}  & http://toxtree.sf.net \\ \hline
         \textbf{XtalOpt}  & http://xtalopt.openmolecules.net \\ \hline

### 2011-06-22 paper.tex
Final edits

@@
 %Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
 
 %Publication style settings
-\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
+%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}
 
@@
 interfaces for chemical content. The project is already being adopted by others,
-including iBabel~\cite{iBabel}, ChemSpotlight~\cite{chemspotlight} and the RSC ChemSpider~\cite{chemspider_chemdoodle}.
+including iBabel~\cite{iBabel}, ChemSpotlight~\cite{chemspotlight} and the RSC ChemSpider~\cite{chemspider,chemspider_chemdoodle}.
 
@@
 which is available under the Creative Commons Share-Alike
-Attribution license~\cite{Overington2009}.
+Attribution license~\cite{Overington2009}. The RSC ChemSpider database~\cite{chemspider}, although
+not fully Open, also hosts Open Data; for example, spectral data when deposited
+can be marked as Open.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Authors contributions}
-   The inital layout of the manuscript grew from discussions between
-NMOB, RG and ELW. All of the authors contributed sections to this
-article, and approved the final manuscript.
+\section*{Authors' contributions}
+The overall layout of the manuscript grew from discussions between NMOB, RG and ELW.
+The authorship of the paper is drawn from those people connected with fully Open Data/Standards/Source (OSI-compliant or OKF-compliant) projects associated with the Blue Obelisk.
+There are a large number of people contributing to these projects and because those projects are published in their own right it is not appropriate to include all their developers by default.
+We invited a number of `project gurus' who have been active in promoting the Blue Obelisk, to be authors on this paper and most have accepted and contributed.
 

### 2011-06-23 paper.tex
Remove gChem

@@
 
-Web services offer a potentially powerful application of Open Source software,
-particularly when combined with Open Data. For example, Metamolecular~\cite{Metamolecular}have released
-gChem~\cite{gChem}, an Open Source library for Google Spreadsheets that enables
-non-programmers to build and maintain chemically-aware spreadsheet documents
-that can be easily shared and published. gChem makes use of the National Cancer Institute's
-Chemical Identifier Resolver~\cite{ChemicalIdentifierResolver} to interconvert
-substance names, database identifiers, SMILES and InChI codes, and to generate
-embedded chemical structure images.
-
 \subsection*{The business end}
@@
         \textbf{ChemDoodle Web Components} & http://web.chemdoodle.com \\ \hline
-        \textbf{gChem} & http://metamolecular.com/gchem/ \\ \hline
         \textbf{Jmol} & http://jmol.org \\ \hline

### 2011-06-27 paper.tex
Remove Rich's name.

@@
   \and
-  Richard L Apodaca$^{6}$%
-  \email{Richard L Apodaca - rapodaca@metamolecular.com}
-  \and
-  Jean-Claude Bradley$^7$%
+  Jean-Claude Bradley$^6$%
   \email{Jean-Claude Bradley - bradlejc@drexel.edu}
   \and
-  Igor V Filippov$^8$%
+  Igor V Filippov$^7$%
   \email{Igor V Filippov - igorf@helix.nih.gov}
   \and
-  Robert M Hanson$^9$%
+  Robert M Hanson$^8$%
   \email{Robert M Hanson - hansonr@stolaf.edu}
   \and
-  Marcus D Hanwell$^{10}$%
+  Marcus D Hanwell$^{9}$%
   \email{Marcus D Hanwell - marcus.hanwell@kitware.com}
   \and
-  Geoffrey R Hutchison$^{11}$%
+  Geoffrey R Hutchison$^{10}$%
   \email{Geoffrey R Hutchison - geoffh@pitt.edu}
   \and
-  Craig A James$^{12}$%
+  Craig A James$^{11}$%
   \email{Craig A James - cjames@emolecules.com}
   \and
-  Nina Jeliazkova$^{13}$%
+  Nina Jeliazkova$^{12}$%
   \email{Nina Jeliazkova - jeliazkova.nina@gmail.com}
   \and
-  Andrew SID Lang$^{14}$%
+  Andrew SID Lang$^{13}$%
   \email{Andrew SID Lang - alang@oru.edu}
   \and
-  Karol M Langner$^{15}$%
+  Karol M Langner$^{14}$%
   \email{Karol M Langner - langnerkm@chem.leidenuniv.nl}
   \and
-  David C Lonie$^{16}$%
+  David C Lonie$^{15}$%
   \email{David C Lonie - loniedavid@gmail.com}
@@
   \and
-  J\'{e}r\^{o}me Pansanel$^{17}$%
+  J\'{e}r\^{o}me Pansanel$^{16}$%
   \email{J\'{e}r\^{o}me Pansanel - j.pansanel@pansanel.net}
   \and
-  Dmitry Pavlov$^{18}$%
+  Dmitry Pavlov$^{17}$%
   \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
@@
   \and
-  Christoph Steinbeck$^{19}$%
+  Christoph Steinbeck$^{18}$%
   \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
   \and
-  Adam L Tenderholt$^{20}$%
+  Adam L Tenderholt$^{19}$%
   \email{Adam L Tenderholt - adamlt82@u.washington.edu}
   \and
-  Kevin J Theisen$^{21}$%
+  Kevin J Theisen$^{20}$%
   \email{Kevin J Theisen - kevin@ichemlabs.com}
@@
     \iid(5)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden \\
-    \iid(6)Metamolecular, LLC, 8070 La Jolla Shores Drive \#464, La Jolla, CA 92037, USA \\
-    \iid(7)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
-    \iid(8)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
+    \iid(6)Department of Chemistry, Drexel University, 32nd and Chestnut streets, Philadelphia, PA 19104, USA \\
+    \iid(7)Chemical Biology Laboratory, Basic Research Program, SAIC-Frederick,
 Inc., NCI-Frederick, Frederick, MD 21702, USA \\
-    \iid(9)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
-    \iid(10)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
-    \iid(11)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
-    \iid(12)eMolecules Inc., 380 Stevens Ave., Solana Beach, California 92075, USA \\
-    \iid(13)Ideaconsult Ltd., 4.A.Kanchev str., Sofia 1000, Bulgaria \\
-    \iid(14)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
-    \iid(15)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
-    \iid(16)Department of Chemistry, State University of New York at Buffalo, Buffalo, NY 14260-3000, USA \\
-    \iid(17)Universit\'{e} de Strasbourg, IPHC, CNRS, UMR7178, 23 rue du Loess 67037, Strasbourg, France \\
-    \iid(18)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
-    \iid(19)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
-    \iid(20)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
-    \iid(21)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA 
+    \iid(8)St. Olaf College, 1520 St. Olaf Ave., Northfield, MN 55057, USA \\
+    \iid(9)Kitware, Inc., 28 Corporate Drive, Clifton Park, NY 12065, USA \\
+    \iid(10)Department of Chemistry, University of Pittsburgh, 219 Parkman Avenue, Pittsburgh, PA 15260, USA \\
+    \iid(11)eMolecules Inc., 380 Stevens Ave., Solana Beach, California 92075, USA \\
+    \iid(12)Ideaconsult Ltd., 4.A.Kanchev str., Sofia 1000, Bulgaria \\
+    \iid(13)Department of Engineering, Computer Science, Physics, and Mathematics, Oral Roberts University, 7777 S. Lewis Ave. Tulsa, OK 74171, USA \\
+    \iid(14)Leiden Institute of Chemistry, Leiden University, Einsteinweg 55, 2333 CC Leiden, The Netherlands \\
+    \iid(15)Department of Chemistry, State University of New York at Buffalo, Buffalo, NY 14260-3000, USA \\
+    \iid(16)Universit\'{e} de Strasbourg, IPHC, CNRS, UMR7178, 23 rue du Loess 67037, Strasbourg, France \\
+    \iid(17)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia \\
+    \iid(18)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK \\
+    \iid(19)Department of Chemistry, University of Washington, Seattle, WA 98195, USA \\
+    \iid(20)iChemLabs, 200 Centennial Ave., Suite 200, Piscataway, NJ 08854, USA 
 }%

### 2011-08-16 paper.tex
Proper spelling of väg.
Response to reviewers. Note: a little more work required for reviewer 2, but I think reviewer 1 is finished

@@
     \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
-        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
+        Nobels v\"ag 13, Karolinska Institutet, 171 77 Stockholm, Sweden \\
     \iid(4)Unilever Centre for Molecular Sciences Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK \\
@@
         \paragraph*{Results:} This
-		contribution looks back on the past 5 years and surveys progress and
+		contribution looks back on the work carried out by the
+Blue Obelisk in the past 5 years and surveys progress and
 		remaining challenges in the areas of Open Data, Open Standards, and
@@
 2006. Its home on the web is at http://blueobelisk.org.
-This contribution looks back on the past 5 years and surveys
-progress and remaining challenges in the areas of Open Data, Open
+This contribution looks back on the work carried out by the Blue
+Obelisk over the past 5 years in the areas of Open Data, Open
 Source, and Open Standards in chemistry.
@@
 interest in tools and approaches based on the web philosophy and
-practice. Several groups are experimenting with 
+practice. Several groups~\cite{willighagen_resource_2011,chen_chem2bio2rdf_2010} are experimenting with 
 the Resource Description Framework (RDF) language

### 2011-08-17 paper.tex
Final edits responding to reviewer's comments

@@
 
-Despite this positive news, little chemical data has become
-available from the traditional chemical fields of organic,
-inorganic, solid state chemistry. Table~2 lists a few notable
-exceptions: Metamolecular's Chempedia~\cite{Chempedia} (a discontinued substance registration
-and naming service whose raw data can still be accessed),
-CrystalEye~\cite{WebCrystalEye},
-and the Open Notebook Science Solubility
-data~\cite{ONS2010}. There is also data available using licenses
+Despite this positive news, little chemical data compatible
+with these principles has become
+available from the traditional chemistry fields of organic,
+inorganic, and solid state chemistry. Table~2 lists a few notable
+exceptions, some of which are discussed further below.
+There is also data available using licenses
 not compatible with the Panton Principles, but where the user
 is allowed to modify and redistribute the data. A new data
-set in this category is the data from the ChEMBL database,
+set in this category is the data from the ChEMBL database~\cite{Overington2009},
 which is available under the Creative Commons Share-Alike
-Attribution license~\cite{Overington2009}. The RSC ChemSpider database~\cite{chemspider}, although
+Attribution license. The RSC ChemSpider database~\cite{chemspider}, although
 not fully Open, also hosts Open Data; for example, spectral data when deposited
@@
 The Blue Obelisk provides also a complementary project, the Chemical
-Structure Repository~\cite{BODR}. It aims to provide 3D coordinates, InChi code and several
+Structure Repository~\cite{BODR}. It aims to provide 3D coordinates, InChIs and several
 physico-chemical descriptors for a set of 570 organic compounds.
@@
   \hline \textbf{Name} & \textbf{License/Waiver} & \textbf{Description} \\ \hline
-  \textbf{Chempedia} & CC0 & Crowd-sourced chemical names (project discontinued) \\ \hline
+  \textbf{Chempedia}~\cite{Chempedia} & CC0 & Crowd-sourced chemical
+names (project discontinued but data still available) \\ \hline
   \textbf{CrystalEye}     & PPDL & Crystal structures from primary literature \\ \hline

### 2011-09-05 paper.tex
Editorial changes

@@
 %\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+\usepackage{longtable}
 \urlstyle{rm}
@@
 
+\newpage
 \begin{abstract}
@@
 of small molecules was created as a resource about chemical structure and nomenclature
-by biologists.\cite{DeMatos:2009p3839}
+by biologists \cite{DeMatos:2009p3839}.
 
@@
 Open Data Commons Attribution License, and the
-Open Data Commons Open Database License (ODbL).\cite{WebOpenData}
+Open Data Commons Open Database License (ODbL) \cite{WebOpenData}.
 
@@
 
+\section*{Competing interests}
+Please include your completing interests statement here.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%
+\bigskip
 \section*{Tables}
   \subsection*{Table 1 - Blue Obelisk Open Source Software projects discussed in the text}
-    \par \mbox{}
-    \par
-    \mbox{
-     \begin{tabular}{|c|c|c|}
+   % \par \mbox{}
+   % \par
+   % \mbox{
+\begin{longtable}{|c|c|c|}
+   %  \begin{tabular}{|c|c|c|}
        \hline \textbf{Name} & \textbf{Website}  \\ \hline
@@
         \hline
-       \end{tabular}
-      }
-
+   %    \end{tabular}
+     % }
+\end{longtable}
+
+\bigskip
   \subsection*{Table 2 - Open Data in chemistry.}
-    Overview of major open chemical data available under a license or waiver
-    compatible with the Panton Principles.
     \par \mbox{}
@@
     \mbox{
-\begin{tabular}{|l|c|l|}
+\begin{tabular}{|l|c|p{9cm}|}
   %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
@@
       }
+\\
+\bigskip Overview of major open chemical data available under a license or waiver compatible with the Panton Principles.
 
